---
title: "Spring phenology tier2 project"
subtitle: "R code extracting the species-specific spring climate (PEP725 data)"
author: "Constantin Zohner"
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

format:
  html:
    toc: false
    theme: default
    highlight-style: haddock
    df-print: paged
    embed-resources: true

execute:
  #warning: false
  message: false
  echo: true
  output: true

always_allow_html: true
editor: visual
editor_options:
  chunk_output_type: inline
---

<br>

### Driver names

-   Tmean: Mean daily temperature
-   Tmin: Minimum daily temperature
-   Tmax: Maximum daily temperature

<br>

### Species names

-   Deciduous trees: Aesculus hippocastanum, Alnus glutinosa, Betula pendula, Fagus sylvatica, Fraxinus excelsior, Quercus robur
-   Deciduous shrub: Ribes uva-crispa
-   Deciduous needleleaf: Larix decidua
-   Evergreen needleleaf: Picea abies, Pinus sylvestris

<br>

# 1. Setup

<details>

<summary>Packages</summary>

```{r}
require(data.table)
require(dplyr)
require(pbmcapply)
```

</details>

<details>

<summary>Directories</summary>

```{r}
#-------------------------------------------------
# Define the base directory
#-------------------------------------------------

base.dir <- "/Users/goalie/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis"


#-------------------------------------------------
# Build specific directories from the base
#-------------------------------------------------

# Climate
EOBS.dir = file.path(base.dir, "Analysis_input/EOBS_data")

# Phenology
PEP.dir = file.path(base.dir, "PEP_data")

# Output
output.dir = file.path(base.dir, "Analysis_input/Spring_climate")
```

</details>

<details>

<summary>Data</summary>

```{r}
#-------------------------------------------------
# Phenology data
#-------------------------------------------------

PEP.df = fread(paste(PEP.dir, "PEP725_data_clean.csv", sep="/")) %>% 
  arrange(species, s_id, year) %>% 
  group_by(species) %>% 
  mutate(
    # Calculate species-level mean leaf-put dates
    leaf_out_species_mean = round(mean(leaf_out)), 
    # Timeseries x year identifier
    ts_year = paste0(timeseries, "_", year),
    # Site x year identifier
    site_year = paste0(s_id, "_", year)) %>% 
  ungroup()


#-------------------------------------------------
## Import daily climatic datasets from GLDAS
#-------------------------------------------------

#define climate variables
list.files(EOBS.dir)
vn <- c("tg_ens_mean_0.1deg_reg_v30.0e.csv", # daily mean temperature 
        "tn_ens_mean_0.1deg_reg_v30.0e.csv", # daily minimum temperature
        "tx_ens_mean_0.1deg_reg_v30.0e.csv") # daily maximum temperature

#create empty list
DataList <- replicate(length(vn),data.frame())
#loop through climate variables
for(i in 1:length(vn)) {
  #read data
  data = fread(paste0(EOBS.dir, "/", vn[i]))
  #add table to list
  DataList[[i]] = data }
#add names to list
names(DataList)=vn


#-------------------------------------------------
# Merge data and add names to list
#-------------------------------------------------

# Add additional datasets
DataList[[length(DataList) + 1]] <- PEP.df

# Remove data
rm(PEP.df)

# Set all names
names(DataList) = c(vn, "PEP725")
names(DataList)

```

</details>

<br>

# 2. Parallel computing

<details>

<summary>Create loop</summary>

```{r}
##########################################################
# Loop through each observation using parallel computing #
##########################################################



# Identifier for loop (all timeseries x year combinations)
timeseries_year  = unique(DataList[["PEP725"]]$ts_year)

# Parallel processing function
parallelCalc <- function(timeseries_years, DataList) {
 
  #-------------------------------------------------
  # extract PEP data per site x year combination
  #-------------------------------------------------
  
  pheno.sub <- DataList[["PEP725"]][DataList[["PEP725"]]$ts_year == timeseries_years, ]
  # Skip if doesn't exist
  if (nrow(pheno.sub) == 0) return(NULL)

  
  #--------------------------------------------------------------
  # Define the length of the period of interest in calendar units
  #--------------------------------------------------------------
  
  year1 <- as.character(pheno.sub$year - 1)
  year2 <- as.character(pheno.sub$year)
  start_doy <- paste0(year1, "-01-01")
  end_doy <- paste0(year2, "-06-30")
  days <- seq(as.Date(start_doy), as.Date(end_doy), by = "days")
  
  #get number of days in year 1 and year 2
  days_year1 <- length(seq(as.Date(start_doy), as.Date(paste0(year1, "-12-31")), by = "days"))
  days_year2 <- length(seq(as.Date(paste0(year2, "-01-01")), as.Date(end_doy), by = "days"))
  
  
  #-------------------------------------------------
  # Load and merge all climate variables
  #-------------------------------------------------
  
  # Function to load climate data
  get_subdata <- function(var_index) {
    # Year 1
    sub1 <- DataList[[var_index]][DataList[[var_index]]$s_id == pheno.sub$s_id &
                                    DataList[[var_index]]$year == year1, as.character(1:days_year1), with = FALSE]
    # Year 2
    sub2 <- DataList[[var_index]][DataList[[var_index]]$site_year == pheno.sub$site_year, as.character(1:days_year2), with = FALSE]
    if (nrow(sub1) == 0 || nrow(sub2) == 0) return(NULL)
    cbind(sub1, sub2)
    }
    
    # ------------------------------------------------------------------------
    # Short Name        ; Long Name                               ; Unit [d-1]
    # ------------------------------------------------------------------------
    # TMIN              ; Minimum Air Temperature                 ; degC
    # TMEAN             ; Mean Air Temperature                    ; degC
    # TMAX              ; Maximum Air Temperature                 ; degC
    # ------------------------------------------------------------------------
    
    Tmean <- get_subdata(1)
    Tmin  <- get_subdata(2)
    Tmax  <- get_subdata(3)
    
    if (any(sapply(list(Tmean, Tmin, Tmax), is.null))) {return(NULL)}
    
    
    #-------------------------------------------------
    # Table of climate variables
    #-------------------------------------------------
    
    daily_vals <- data.frame(
      Year  = lubridate::year(days),
      Month = lubridate::month(days),
      Day   = lubridate::day(days),
      DOY   = 1:length(days),
      Tmin  = as.numeric(Tmin),
      Tmean = as.numeric(Tmean),
      Tmax  = as.numeric(Tmax) )
     
    
    #-------------------------------------------------
    # Generate sub-dataframe to store results
    #-------------------------------------------------
    
    factors.sub <- pheno.sub %>% 
      dplyr::select(s_id, species, timeseries, year, lat, lon)

    
    #-------------------------------------------------
    # Get key dates
    #-------------------------------------------------
    
    #row of mean leaf-out date
    DOY_out_mean = days_year1 + pheno.sub$leaf_out_species_mean

    
    #-------------------------------------------------
    # Preseason means for each climate driver
    #-------------------------------------------------
    
    #*********************************************************************************
    # Calculates preseason temperature: 
    # - averages
    # - intra-annual variation 
    # - slopes
    #*********************************************************************************
    
    # get preseason length vector (10 to 120 days with 30-day steps)
    preseason.lengths = seq(30, 120, 30)
    
    #loop through preseasons
    for(len in preseason.lengths) {
      # define preseason period
      idx = (DOY_out_mean - len) : DOY_out_mean
      # Means
      factors.sub[[paste0("Tmean.spring", len)]] <- mean(daily_vals$Tmean[idx], na.rm = TRUE)
      factors.sub[[paste0("Tmax.spring", len)]]  <- mean(daily_vals$Tmax[idx],  na.rm = TRUE)
      factors.sub[[paste0("Tmin.spring", len)]]  <- mean(daily_vals$Tmin[idx],  na.rm = TRUE)
      # SDs
      factors.sub[[paste0("Tmean.IntraSD.spring", len)]] <- sd(daily_vals$Tmean[idx], na.rm = TRUE)
      factors.sub[[paste0("Tmax.IntraSD.spring", len)]]  <- sd(daily_vals$Tmax[idx],  na.rm = TRUE)
      factors.sub[[paste0("Tmin.IntraSD.spring", len)]]  <- sd(daily_vals$Tmin[idx],  na.rm = TRUE)
      # Slopes
      factors.sub[[paste0("Tmean.slope.spring", len)]] <- coef(lm(daily_vals$Tmean[idx] ~ daily_vals$DOY[idx]))[2]
      factors.sub[[paste0("Tmax.slope.spring", len)]] <- coef(lm(daily_vals$Tmax[idx] ~ daily_vals$DOY[idx]))[2]
      factors.sub[[paste0("Tmin.slope.spring", len)]] <- coef(lm(daily_vals$Tmin[idx] ~ daily_vals$DOY[idx]))[2]
      }

    
    #-------------------------------------------------
    # Safe the table  
    #-------------------------------------------------
    
    return(factors.sub) 
    
 }
```

</details>

<details>

<summary>Run loop and safe data</summary>

```{r}
#-------------------------------------------------
# Run in parallel
#-------------------------------------------------

outputlist <- pbmclapply(timeseries_year, function(y) parallelCalc(y, DataList), mc.cores = 12, mc.preschedule = T)
climate.factors.df <- rbindlist(outputlist, fill = TRUE)


#-------------------------------------------------
# Save results
#-------------------------------------------------

write.csv(climate.factors.df, file.path(output.dir, "pep_drivers_spring_climate.csv"))
```

</details>

<br>

# 3. Reproducibility

<details>

<summary>Show</summary>

```{r}
# date time
Sys.time()

# Session info
sessionInfo()
```

</details>
