---
title: "Spring phenology tier2 project"
subtitle: "Alternate Bayesian model figures (PEP725 data)"
author: "Constantin Zohner"
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

format:
  html:
    toc: false
    theme: default
    highlight-style: haddock
    df-print: paged
    embed-resources: true

execute:
  #warning: false
  message: false
  echo: true
  output: true

always_allow_html: true
editor: visual
editor_options:
  chunk_output_type: inline
---

<br>

### Models

-   Full_noSpring: no specific preseason temperature, just mean Jan to March temperatures
-   Full_noScale: no site-level centering (and scaling) of variables, just global scaling
-   Full_noYear: no year term

### Variable names

#### Preseason

##### (start and end dates vary across time series; start date = fitted; end date = time series-level mean leaf-out)

-   Tmean/max: Preseason mean/maximum daily air temperature
-   SWrad: Preseason short-wave radiation
-   RH: Preseason relative humidity

#### Fixed periods

-   Tmean.EQ1.SO1: Mean daily air temperature (March equinox to summer solstice of preceding year)
-   Tmean.SO1.EQ2: Mean daily air temperature (Summer solstice to September equinox of preceding year)
-   Tmean.EQ2.SO2: Mean daily air temperature (September equinox to winter solstice of preceding year)

#### Phenology

-   leaf_out_prev: Leaf-out date of preceding year
-   leaf_off_prev: Leaf senescence date of preceding year

#### Late frosts in preceding year

-   LSF2: Binary indicators of late spring frost events (\<-2°C) occurring after leaf-out in the previous year

<br>

# 1. Setup

<details>

<summary>Packages</summary>

```{r, message=F}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggpattern)
library(brms)
library(posterior)
library(rstan)
library(pracma)
```

</details>

<details>

<summary>Functions</summary>

```{r, message=F}
#-----------------------------------------------------------
# Extract key convergence diagnostics from a brmsfit model
#-----------------------------------------------------------

extract_diagnostics <- function(model, period, species) {
  draws <- as_draws_df(model)
  diag_summary <- summarise_draws(draws)
  data.frame(
    period = period,
    species = species,
    max_rhat = max(diag_summary$rhat, na.rm = TRUE),               # Rhat convergence statistic
    min_ess_bulk = min(diag_summary$ess_bulk, na.rm = TRUE),       # Minimum effective sample size (bulk)
    min_ess_tail = min(diag_summary$ess_tail, na.rm = TRUE),       # Minimum effective sample size (tail)
    n_divergent = sum(model$diagnostic_summary$divergent__),       # Total number of divergent transitions
    treedepth_exceeded = sum(model$diagnostic_summary$treedepth__) # Total number of transitions exceeding tree depth
  )
}


#-----------------------------------------------------------
# Interpolation function
#-----------------------------------------------------------

lin_interp = function(x, y, length.out=100) {
  approx(x, y, xout=seq(min(x), max(x), length.out=length.out))$y
}
```

</details>

<details>

<summary>Plot theme</summary>

```{r, message=F}
plotTheme1 = theme(legend.position   = "none",
                   legend.background = element_rect(fill=NA, linewidth=0.5, linetype="solid"),
                   legend.text       = element_text(color="black"),
                   panel.grid.major  = element_line(color = "grey90"),
                   panel.background  = element_blank(),
                   panel.border      = element_rect(colour = "black", fill=NA),
                   axis.line         = element_line(color = "black"),
                   axis.text         = element_text(colour = "black"),
                   strip.background  = element_rect(fill=NA),
                   strip.text        = element_text(colour = 'black', face="italic"),
                   plot.title        = element_text(hjust = 0.5))
```

</details>

<details>

<summary>Directories</summary>

```{r}
#-------------------------------------------------
# Define the base directory
#-------------------------------------------------

base.dir = "/Users/goalie/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_output"


#-------------------------------------------------
# Build specific directories from the base
#-------------------------------------------------

# Model results
model.dir = file.path(base.dir, "ModelResults/Bayesian_models")

# Figure output
output.dir = file.path(base.dir, "ModelFigures/Bayesian_models")
```

</details>

<details>

<summary>Data</summary>

```{r}
#----------------------------------------------------
# no spring preseason model
#----------------------------------------------------

model.noSpring    = readRDS(file.path(model.dir, "brms_full_models_noSpring.rds"))


#----------------------------------------------------
# no site-level scaling model
#----------------------------------------------------

model.noScale    = readRDS(file.path(model.dir, "brms_full_models_noScale.rds"))


#----------------------------------------------------
# no Year model
#----------------------------------------------------

model.noYear    = readRDS(file.path(model.dir, "brms_full_models_noYear.rds"))
```

</details>

<br>

# 2. Model diagnostics

<details>

<summary>No Spring model</summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.noSpring, function(period_list, period) {
  imap_dfr(period_list, function(species_list, species) {
    extract_diagnostics(species_list, period = period, species = species)
  })
}) %>%
  # Tidy and reorder variables
  mutate(species = factor(species, levels = rev(sort(unique(species)))))  # alphabetical order
       
  
#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models by model type and period
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Group data by phenological period (early/late)
  group_by(period) %>%
  # Calculate summary statistics within each group
  summarise(
    prop_rhat_ok = mean(max_rhat < 1.01),        # Proportion of models with Rhat < 1.02 (indicates good convergence)
    prop_ess_bulk_ok = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 200 (good sampling in the posterior bulk)
    prop_ess_tail_ok = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 200 (good sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),        # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded), # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = species, y = max_rhat, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  labs(
    x = "Species",
    y = "Maximum Rhat (Model convergence)",
    color = "Model",
    shape = "Period"
  ) +
  coord_flip() +
  plotTheme1+
  theme(legend.position = "top",
        axis.text.y = element_text(face = "italic")) 

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = species, y = min_ess_bulk, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Bulk ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = species, y = min_ess_tail, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Tail ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_noSpring.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_noSpring.plot, file="diagnostics_full_model_noSpring.pdf", path=output.dir,
       width=8, height=8)

diagnostics_noSpring.plot 
```

</details>

<details>

<summary>No Scale model</summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.noScale, function(period_list, period) {
  imap_dfr(period_list, function(species_list, species) {
    extract_diagnostics(species_list, period = period, species = species)
  })
}) %>%
  # Tidy and reorder variables
  mutate(species = factor(species, levels = rev(sort(unique(species)))))  # alphabetical order
         

#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models by model type and period
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Group data by phenological period (early/late)
  group_by(period) %>%
  # Calculate summary statistics within each group
  summarise(
    prop_rhat_ok = mean(max_rhat < 1.01),        # Proportion of models with Rhat < 1.02 (indicates good convergence)
    prop_ess_bulk_ok = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 200 (good sampling in the posterior bulk)
    prop_ess_tail_ok = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 200 (good sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),        # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded), # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = species, y = max_rhat, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  labs(
    x = "Species",
    y = "Maximum Rhat (Model convergence)",
    color = "Model",
    shape = "Period"
  ) +
  coord_flip() +
  plotTheme1+
  theme(legend.position = "top",
        axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = species, y = min_ess_bulk, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Bulk ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = species, y = min_ess_tail, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Tail ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_noScale.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_noScale.plot , file="diagnostics_full_model_noScale.pdf", path=output.dir,
       width=8, height=8)

diagnostics_noScale.plot 
```

</details>

<details>

<summary>No Year model</summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.noYear, function(period_list, period) {
  imap_dfr(period_list, function(species_list, species) {
    extract_diagnostics(species_list, period = period, species = species)
  })
}) %>%
  # Tidy and reorder variables
  mutate(species = factor(species, levels = rev(sort(unique(species)))))  # alphabetical order
       
  
#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models by model type and period
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Group data by phenological period (early/late)
  group_by(period) %>%
  # Calculate summary statistics within each group
  summarise(
    prop_rhat_ok = mean(max_rhat < 1.01),        # Proportion of models with Rhat < 1.02 (indicates good convergence)
    prop_ess_bulk_ok = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 200 (good sampling in the posterior bulk)
    prop_ess_tail_ok = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 200 (good sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),        # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded), # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = species, y = max_rhat, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  labs(
    x = "Species",
    y = "Maximum Rhat (Model convergence)",
    color = "Model",
    shape = "Period"
  ) +
  coord_flip() +
  plotTheme1+
  theme(legend.position = "top",
        axis.text.y = element_text(face = "italic")) 

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = species, y = min_ess_bulk, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Bulk ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = species, y = min_ess_tail, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Tail ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_noYear.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_noYear.plot, file="diagnostics_full_model_noYear.pdf", path=output.dir,
       width=8, height=8)

diagnostics_noYear.plot 
```

</details>

<br>

# 3. Model results - Extract posterior stats

<details>

<summary>No Spring model</summary>

```{r, warning=FALSE}
#-----------------------------------------------------------
# 1. Loop through 2-level nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.noSpring, function(period_list, period_name) {
    purrr::imap_dfr(period_list, function(model_obj, species_name) {
      if (inherits(model_obj, "brmsfit")) {
        broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
          mutate(
            period = period_name,
            species = species_name,
            .before = term
          )
      } else {
        NULL  # skip if not a brms model
      }
    })
}) %>% 
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
      species = factor(species, levels = rev(sort(unique(species)))), # alphabetical order
      term = plyr::revalue(term, warn_missing=F,
                           c("leaf_out_prev_z"  = "Leaf-out",
                             "LSF2"             = "Late frost",
                             "Tmean.EQ1.SO1_z"  = "TQ2",
                             "Tmean.SO1.EQ2_z"  = "TQ3",
                             "leaf_off_z"       = "Leaf-off",
                             "Tmean.EQ2.SO2_z"  = "TQ4",
                             "RH_z"             = "Humidity",
                             "SWrad_z"          = "Radiation",
                             "Tmax_spring_z"    = "Tmax")))


#-----------------------------------------------------------
# 2. Summarize data and create plotting helpers
#-----------------------------------------------------------

# Define consistent term order
term.order <- c("Leaf-out","Late frost",
                "TQ2","TQ3","Leaf-off","TQ4",
                "Humidity","Radiation","Tmax")

# Summarize data
summary_model.df <- model.df %>%
    group_by(term, period) %>%
    summarise(
      mean_estimate = mean(estimate, na.rm = TRUE),
      # 100% of the individual species' 95% CIs fall within this range
      lower = min(conf.low, na.rm = TRUE),
      upper = max(conf.high,na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(term = factor(term, levels = term.order))
  
# get the number of terms
terms <- unique(model.df$term)
n_terms <- length(terms)
  
# Create numeric positions halfway between rows (used for spacing lines between terms)
grid_lines <- seq(1.5, n_terms - 0.5, by = 1)
  
# Define dodge & jitter
dodge <- position_dodge(width = 0.6)
point_dodge_jitter <- position_jitterdodge(
  dodge.width = 0.6,
  jitter.width = 0.35,
  jitter.height = 0
)


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_noSpring.plot = ggplot(model.df, aes(x = term, group = period)) +
  
  # Horizontal reference line
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
    
  # Summary error bars with dashed for early, solid for late
  geom_errorbar(data = summary_model.df,
                aes(x = term, y = mean_estimate, ymin = lower, ymax = upper, linetype = period),
                position = dodge, width = 0.3, linewidth = 0.6,
                inherit.aes = FALSE, color = "black") +
    
  # Species-level error bars
  geom_errorbar(aes(y = estimate, ymin = conf.low, ymax = conf.high, color = species),
                position = dodge, width = 0.2, linewidth = 0.5, alpha = 0.5) +
    
  # Species-level points
  geom_point(aes(y = estimate, color = species, shape = period),
             position = point_dodge_jitter, size = 1.8, alpha = 0.7) +
    
  # Shape, color, linetype scales
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  scale_linetype_manual(values = c("early" = "dashed", "late" = "solid")) +
  scale_color_manual(values = c(
    "#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99",
    "#e31a1c", "#fdbf6f", "#ff7f00", "#cab2d6", "#6a3d9a"
  )) +
    
  # Flip axes and label
  coord_flip(ylim = c(-0.75, 0.75)) +
  xlab("") + ylab("") +
  labs(title = "Fixed preseason model") +
  
  # Theme
  plotTheme1 +
  theme(
    axis.text.x = element_text(angle = 0),
    axis.text.y = element_text(face = "plain"),
    panel.grid.major = element_blank(),
  ) +
    
  # Tidy legend
  guides(color = guide_legend(override.aes = list(alpha = 1)))+
  # add horizontal lines at those positions
  geom_vline(xintercept = grid_lines, linetype = "solid", color = "grey80")
  
results_noSpring.plot
```

</details>

<details>

<summary>No Scale model</summary>

```{r, warning=FALSE}
#-----------------------------------------------------------
# 1. Loop through 2-level nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.noScale, function(period_list, period_name) {
    purrr::imap_dfr(period_list, function(model_obj, species_name) {
      if (inherits(model_obj, "brmsfit")) {
        broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
          mutate(
            period = period_name,
            species = species_name,
            .before = term
          )
      } else {
        NULL  # skip if not a brms model
      }
    })
}) %>% 
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
      species = factor(species, levels = rev(sort(unique(species)))), # alphabetical order
      term = plyr::revalue(term, warn_missing=F,
                           c("leaf_out_prev_z_global"  = "Leaf-out",
                             "LSF2"                    = "Late frost",
                             "Tmean.EQ1.SO1_z_global"  = "TQ2",
                             "Tmean.SO1.EQ2_z_global"  = "TQ3",
                             "leaf_off_z_global"       = "Leaf-off",
                             "Tmean.EQ2.SO2_z_global"  = "TQ4",
                             "RH_z_global"             = "Humidity",
                             "SWrad_z_global"          = "Radiation",
                             "Tmax_z_global"           = "Tmax")))


#-----------------------------------------------------------
# 2. Summarize data and create plotting helpers
#-----------------------------------------------------------

# Define consistent term order
term.order <- c("Leaf-out","Late frost",
                "TQ2","TQ3","Leaf-off","TQ4",
                "Humidity","Radiation","Tmax")
  
summary_model.df <- model.df %>%
    group_by(term, period) %>%
    summarise(
      mean_estimate = mean(estimate, na.rm = TRUE),
      # 100% of the individual species' 95% CIs fall within this range
      lower = min(conf.low, na.rm = TRUE),
      upper = max(conf.high,na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(term = factor(term, levels = term.order))
  
# First, get the number of terms
terms <- unique(model.df$term)
n_terms <- length(terms)
  
# Create numeric positions halfway between rows (used for spacing lines between terms)
grid_lines <- seq(1.5, n_terms - 0.5, by = 1)
  
# Define dodge & jitter
dodge <- position_dodge(width = 0.6)
point_dodge_jitter <- position_jitterdodge(
  dodge.width = 0.6,
  jitter.width = 0.35,
  jitter.height = 0
)


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_noScale.plot = ggplot(model.df, aes(x = term, group = period)) +
  # Horizontal reference line
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
    
  # Summary error bars with dashed for early, solid for late
  geom_errorbar(data = summary_model.df,
                aes(x = term, y = mean_estimate, ymin = lower, ymax = upper, linetype = period),
                position = dodge, width = 0.3, linewidth = 0.6,
                inherit.aes = FALSE, color = "black") +
    
  # Species-level error bars
  geom_errorbar(aes(y = estimate, ymin = conf.low, ymax = conf.high, color = species),
                position = dodge, width = 0.2, linewidth = 0.5, alpha = 0.5) +
    
  # Species-level points
  geom_point(aes(y = estimate, color = species, shape = period),
             position = point_dodge_jitter, size = 1.8, alpha = 0.7) +
    
  # Shape, color, linetype scales
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  scale_linetype_manual(values = c("early" = "dashed", "late" = "solid")) +
  scale_color_manual(values = c(
    "#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99",
    "#e31a1c", "#fdbf6f", "#ff7f00", "#cab2d6", "#6a3d9a"
  )) +
    
  # Flip axes and label
  coord_flip(ylim = c(-0.75, 0.75)) +
  xlab("") + ylab("") +
  labs(title = "Uncentered predictor model") +

  # Theme
  plotTheme1 +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 0),
    axis.text.y = element_text(face = "plain"),
    panel.grid.major = element_blank()
  ) +
    
  # Tidy legend
  guides(color = guide_legend(override.aes = list(alpha = 1)))+
  # add horizontal lines at those positions
  geom_vline(xintercept = grid_lines, linetype = "solid", color = "grey80")
  
results_noScale.plot
```

</details>

<details>

<summary>No Year model</summary>

```{r, warning=FALSE}
#-----------------------------------------------------------
# 1. Loop through 2-level nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.noYear, function(period_list, period_name) {
    purrr::imap_dfr(period_list, function(model_obj, species_name) {
      if (inherits(model_obj, "brmsfit")) {
        broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
          mutate(
            period = period_name,
            species = species_name,
            .before = term
          )
      } else {
        NULL  # skip if not a brms model
      }
    })
}) %>% 
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
      species = factor(species, levels = rev(sort(unique(species)))), # alphabetical order
      term = plyr::revalue(term, warn_missing=F,
                           c("leaf_out_prev_z"  = "Leaf-out",
                             "LSF2"             = "Late frost",
                             "Tmean.EQ1.SO1_z"  = "TQ2",
                             "Tmean.SO1.EQ2_z"  = "TQ3",
                             "leaf_off_z"       = "Leaf-off",
                             "Tmean.EQ2.SO2_z"  = "TQ4",
                             "RH_z"             = "Humidity",
                             "SWrad_z"          = "Radiation",
                             "Tmax_z"           = "Tmax")))


#-----------------------------------------------------------
# 2. Summarize data and create plotting helpers
#-----------------------------------------------------------

# Define consistent term order
term.order <- c("Leaf-out","Late frost",
                "TQ2","TQ3","Leaf-off","TQ4",
                "Humidity","Radiation","Tmax")

# Summarize data
summary_model.df <- model.df %>%
    group_by(term, period) %>%
    summarise(
      mean_estimate = mean(estimate, na.rm = TRUE),
      # 100% of the individual species' 95% CIs fall within this range
      lower = min(conf.low, na.rm = TRUE),
      upper = max(conf.high,na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(term = factor(term, levels = term.order))
  
# get the number of terms
terms <- unique(model.df$term)
n_terms <- length(terms)
  
# Create numeric positions halfway between rows (used for spacing lines between terms)
grid_lines <- seq(1.5, n_terms - 0.5, by = 1)
  
# Define dodge & jitter
dodge <- position_dodge(width = 0.6)
point_dodge_jitter <- position_jitterdodge(
  dodge.width = 0.6,
  jitter.width = 0.35,
  jitter.height = 0
)


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_noYear.plot = ggplot(model.df, aes(x = term, group = period)) +
  
  # Horizontal reference line
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
    
  # Summary error bars with dashed for early, solid for late
  geom_errorbar(data = summary_model.df,
                aes(x = term, y = mean_estimate, ymin = lower, ymax = upper, linetype = period),
                position = dodge, width = 0.3, linewidth = 0.6,
                inherit.aes = FALSE, color = "black") +
    
  # Species-level error bars
  geom_errorbar(aes(y = estimate, ymin = conf.low, ymax = conf.high, color = species),
                position = dodge, width = 0.2, linewidth = 0.5, alpha = 0.5) +
    
  # Species-level points
  geom_point(aes(y = estimate, color = species, shape = period),
             position = point_dodge_jitter, size = 1.8, alpha = 0.7) +
    
  # Shape, color, linetype scales
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  scale_linetype_manual(values = c("early" = "dashed", "late" = "solid")) +
  scale_color_manual(values = c(
    "#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99",
    "#e31a1c", "#fdbf6f", "#ff7f00", "#cab2d6", "#6a3d9a"
  )) +
    
  # Flip axes and label
  coord_flip(ylim = c(-.75, .75)) +
  xlab("") + ylab("Standardized effect (posterior mean ± 95% CI)") +
  labs(title = "No year term model") +
  
  # Theme
  plotTheme1 +
  theme(
    axis.text.x = element_text(angle = 0),
    axis.text.y = element_text(face = "plain"),
    panel.grid.major = element_blank(),
  ) +
    
  # Tidy legend
  guides(color = guide_legend(override.aes = list(alpha = 1)))+
  # add horizontal lines at those positions
  geom_vline(xintercept = grid_lines, linetype = "solid", color = "grey80")
  
results_noYear.plot
```

</details>

<details>

<summary>Save plots</summary>

```{r, warning=FALSE}
#-------------------------------------------------------
# Define plot layouts
#-------------------------------------------------------

layout <- "
A
B
C"


#-------------------------------------------------------
# Merge plots
#-------------------------------------------------------

results.plot = results_noSpring.plot + results_noScale.plot + results_noYear.plot +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))


#-------------------------------------------------------
# Save figure as .pdf
#-------------------------------------------------------

ggsave(results.plot, file="results_full_model_alternate_formulations.pdf", path=output.dir,
       width=7, height=10)
```

</details>

<br>

# 4. Reproducibility

<details>

<summary>show</summary>

```{r}
# Datetime
Sys.time()

# Session info
sessionInfo()
```

</details>
