---
title: "Spring phenology tier2 project"
subtitle: "Bayesian model figures (PEP725 data)"
author: "Constantin Zohner"
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

format:
  html:
    toc: false
    theme: default
    highlight-style: haddock
    df-print: paged
    embed-resources: true

execute:
  #warning: false
  message: false
  echo: true
  output: true

always_allow_html: true
editor: visual
editor_options:
  chunk_output_type: inline
---

<br>

### Models

-   Full: Multivariate model
-   Full alternative: Replacing LSF2 with LSF4, Tmax with Tmean, and leaf_out_prev with leaf_out_prev2
-   Monthly: Models including monthly mean temperatures

### Variable names

#### Preseason

##### (start and end dates vary across time series; start date = fitted; end date = time series-level mean leaf-out)

-   Tmean/max: Preseason mean/maximum daily air temperature
-   SWrad: Preseason short-wave radiation
-   RH: Preseason relative humidity

#### Fixed periods

-   Tmean.GS: Growing-season air temperature (May 1 to September 30 of preceding year)
-   Tmean.EQ1.SO1: Mean daily air temperature (March equinox to summer solstice of preceding year)
-   Tmean.SO1.EQ2: Mean daily air temperature (Summer solstice to September equinox of preceding year)
-   Tmean.EQ2.SO2: Mean daily air temperature (September equinox to winter solstice of preceding year)
-   Tmean.8-17: Mean monthly air temperature from August of previous year to May of current year

#### Phenology

-   leaf_out_prev: Leaf-out date of preceding year
-   leaf_out_prev2: Leaf-out date two years ago
-   leaf_off_prev: Leaf senescence date of preceding year

#### Late frosts in preceding year

-   LSF2 / LSF4: Binary indicators of late spring frost events occurring after leaf-out in the previous year:
    -   LSF2: Set to 1 if temperature dropped below –2°C
    -   LSF4: Set to 1 if temperature dropped below –4°C

<br>

# 1. Setup

<details>

<summary>Packages</summary>

```{r, message=F}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggpattern)
library(brms)
library(posterior)
library(rstan)
library(pracma)
```

</details>

<details>

<summary>Functions</summary>

```{r, message=F}
#-----------------------------------------------------------
# Extract key convergence diagnostics from a brmsfit model
#-----------------------------------------------------------

extract_diagnostics <- function(model, period, species) {
  draws <- as_draws_df(model)
  diag_summary <- summarise_draws(draws)
  data.frame(
    period = period,
    species = species,
    max_rhat = max(diag_summary$rhat, na.rm = TRUE),               # Rhat convergence statistic
    min_ess_bulk = min(diag_summary$ess_bulk, na.rm = TRUE),       # Minimum effective sample size (bulk)
    min_ess_tail = min(diag_summary$ess_tail, na.rm = TRUE),       # Minimum effective sample size (tail)
    n_divergent = sum(model$diagnostic_summary$divergent__),       # Total number of divergent transitions
    treedepth_exceeded = sum(model$diagnostic_summary$treedepth__) # Total number of transitions exceeding tree depth
  )
}


#-----------------------------------------------------------
# Interpolation function
#-----------------------------------------------------------

lin_interp = function(x, y, length.out=100) {
  approx(x, y, xout=seq(min(x), max(x), length.out=length.out))$y
}
```

</details>

<details>

<summary>Plot theme</summary>

```{r, message=F}
plotTheme1 = theme(legend.position   = "none",
                   legend.background = element_rect(fill=NA, linewidth=0.5, linetype="solid"),
                   legend.text       = element_text(color="black"),
                   panel.grid.major  = element_line(color = "grey90"),
                   panel.background  = element_blank(),
                   panel.border      = element_rect(colour = "black", fill=NA),
                   axis.line         = element_line(color = "black"),
                   axis.text         = element_text(colour = "black"),
                   strip.background  = element_rect(fill=NA),
                   strip.text        = element_text(colour = 'black', face="italic"),
                   plot.title        = element_text(face="bold"))
```

</details>

<details>

<summary>Directories</summary>

```{r}
#-------------------------------------------------
# Define the base directory
#-------------------------------------------------

base.dir = "/Users/goalie/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_output"


#-------------------------------------------------
# Build specific directories from the base
#-------------------------------------------------

# Model results
model.dir = file.path(base.dir, "ModelResults/Bayesian_models")

# Figure output
output.dir = file.path(base.dir, "ModelFigures/Bayesian_models")
```

</details>

<details>

<summary>Data</summary>

```{r}
#----------------------------------------------------
# Full: Multivariate model
#----------------------------------------------------

model.full     = readRDS(file.path(model.dir, "brms_full_models.rds"))


#----------------------------------------------------
# Full alternative: Replacing LSF2 with LSF4, Tmax with Tmean, and leaf_out_prev with leaf_out_prev2
#----------------------------------------------------

model.full.alt = readRDS(file.path(model.dir, "brms_full_models_alternative.rds"))


#----------------------------------------------------
# Full: Multivariate model alternative (student distribution + response centered but not scaled)
#----------------------------------------------------

model.full.SC    = readRDS(file.path(model.dir, "brms_full_models_studentCenter.rds")) 


#----------------------------------------------------
# Monthly: Models including monthly mean temperatures
#----------------------------------------------------

model.monthly  = readRDS(file.path(model.dir, "brms_monthly_models.rds"))
```

</details>

<br>

# 2. Model diagnostics

<details>

<summary>Full model</summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.full, function(period_list, period) {
  imap_dfr(period_list, function(species_list, species) {
    extract_diagnostics(species_list, period = period, species = species)
  })
}) %>%
  # Tidy and reorder variables
  mutate(species = factor(species, levels = rev(sort(unique(species)))))  # alphabetical order
       
  
#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models by model type and period
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Group data by phenological period (early/late)
  group_by(period) %>%
  # Calculate summary statistics within each group
  summarise(
    prop_rhat_ok = mean(max_rhat < 1.01),        # Proportion of models with Rhat < 1.02 (indicates good convergence)
    prop_ess_bulk_ok = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 200 (good sampling in the posterior bulk)
    prop_ess_tail_ok = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 200 (good sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),        # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded), # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = species, y = max_rhat, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  labs(
    x = "Species",
    y = "Maximum Rhat (Model convergence)",
    color = "Model",
    shape = "Period"
  ) +
  coord_flip() +
  plotTheme1+
  theme(legend.position = "top",
        axis.text.y = element_text(face = "italic")) 

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = species, y = min_ess_bulk, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Bulk ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = species, y = min_ess_tail, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Tail ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_full_model.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_full_model.plot, file="diagnostics_full_model.pdf", path=output.dir,
       width=8, height=8)

diagnostics_full_model.plot 
```

</details>

<details>

<summary>Full model alternative</summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.full.alt, function(period_list, period) {
  imap_dfr(period_list, function(species_list, species) {
    extract_diagnostics(species_list, period = period, species = species)
  })
}) %>%
  # Tidy and reorder variables
  mutate(species = factor(species, levels = rev(sort(unique(species)))))  # alphabetical order
         

#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models by model type and period
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Group data by phenological period (early/late)
  group_by(period) %>%
  # Calculate summary statistics within each group
  summarise(
    prop_rhat_ok = mean(max_rhat < 1.01),        # Proportion of models with Rhat < 1.02 (indicates good convergence)
    prop_ess_bulk_ok = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 200 (good sampling in the posterior bulk)
    prop_ess_tail_ok = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 200 (good sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),        # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded), # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = species, y = max_rhat, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  labs(
    x = "Species",
    y = "Maximum Rhat (Model convergence)",
    color = "Model",
    shape = "Period"
  ) +
  coord_flip() +
  plotTheme1+
  theme(legend.position = "top",
        axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = species, y = min_ess_bulk, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Bulk ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = species, y = min_ess_tail, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Tail ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_full_model_alt.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_full_model_alt.plot , file="diagnostics_full_model_alternative.pdf", path=output.dir,
       width=8, height=8)

diagnostics_full_model_alt.plot 
```

</details>

<details>

<summary>Full studentCenter model </summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.full.SC, function(period_list, period) {
  imap_dfr(period_list, function(species_list, species) {
    extract_diagnostics(species_list, period = period, species = species)
  })
}) %>%
  # Tidy and reorder variables
  mutate(species = factor(species, levels = rev(sort(unique(species)))))  # alphabetical order
       
  
#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models by model type and period
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Group data by phenological period (early/late)
  group_by(period) %>%
  # Calculate summary statistics within each group
  summarise(
    prop_rhat_ok = mean(max_rhat < 1.01),        # Proportion of models with Rhat < 1.02 (indicates good convergence)
    prop_ess_bulk_ok = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 200 (good sampling in the posterior bulk)
    prop_ess_tail_ok = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 200 (good sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),        # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded), # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = species, y = max_rhat, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  labs(
    x = "Species",
    y = "Maximum Rhat (Model convergence)",
    color = "Model",
    shape = "Period"
  ) +
  coord_flip() +
  plotTheme1+
  theme(legend.position = "top",
        axis.text.y = element_text(face = "italic")) 

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = species, y = min_ess_bulk, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Bulk ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = species, y = min_ess_tail, color = period, shape = period)) +
  geom_jitter(width = 0.2, height = 0, size = 2.5, alpha = 0.9) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Tail ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_full_SC_model.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_full_SC_model.plot, file="diagnostics_full_studentCenter_model.pdf", path=output.dir,
       width=8, height=8)

diagnostics_full_SC_model.plot 
```

</details>

<details>

<summary>Monthly model</summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.monthly, function(period_list, period) {
  imap_dfr(period_list, function(species_list, species) {
    extract_diagnostics(species_list, period = period, species = species)
  })
}) %>%
  # Tidy and reorder variables
  mutate(species = factor(species, levels = rev(sort(unique(species)))))  # alphabetical order
         

#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models by model type and period
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Calculate summary statistics within each group
  summarise(
    prop_rhat_ok      = mean(max_rhat < 1.01),    # Proportion of models with Rhat < 1.02 (indicates good convergence)
    prop_ess_bulk_ok  = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 200 (good sampling in the posterior bulk)
    prop_ess_tail_ok  = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 200 (good sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),         # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded),  # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = species, y = max_rhat)) +
  geom_point() +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  labs(
    x = "Species",
    y = "Maximum Rhat (Model convergence)",
    color = "Model"
  ) +
  coord_flip() +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = species, y = min_ess_bulk)) +
  geom_point() +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Bulk ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = species, y = min_ess_tail)) +
  geom_point() +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "Species", y = "Minimum Tail ESS") +
  coord_flip() +
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_monthly_model.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_monthly_model.plot, file="diagnostics_monthly_model.pdf", path=output.dir,
       width=8, height=8)

diagnostics_monthly_model.plot
```

</details>

<br>

# 3. Model results - Extract posterior stats

<details>

<summary>Full model</summary>

```{r, warning=FALSE}
#-----------------------------------------------------------
# 1. Loop through 2-level nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.full, function(period_list, period_name) {
    purrr::imap_dfr(period_list, function(model_obj, species_name) {
      if (inherits(model_obj, "brmsfit")) {
        broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
          mutate(
            period = period_name,
            species = species_name,
            .before = term
          )
      } else {
        NULL  # skip if not a brms model
      }
    })
}) %>% 
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
      species = factor(species, levels = rev(sort(unique(species)))), # alphabetical order
      term = plyr::revalue(term, warn_missing=F,
                           c("leaf_out_prev" = "Leaf-out",
                             "LSF2"           = "Late frost",
                             "Tmean.EQ1.SO1"  = "Tmean.Q2",
                             "Tmean.SO1.EQ2"  = "Tmean.Q3",
                             "leaf_off"       = "Leaf-off",
                             "Tmean.EQ2.SO2"  = "Tmean.Q4",
                             "SWrad"          = "SWR",
                             "Tmax"           = "Tmax")))


#-----------------------------------------------------------
# 2. Summarize data and create plotting helpers
#-----------------------------------------------------------

# Define consistent term order
term.order <- c("Leaf-out","Late frost",
                "Tmean.Q2","Tmean.Q3","Leaf-off","Tmean.Q4",
                "RH","SWR","Tmax")

# Summarize data
summary_model.df <- model.df %>%
    group_by(term, period) %>%
    summarise(
      mean_estimate = mean(estimate, na.rm = TRUE),
      # 100% of the individual species' 95% CIs fall within this range
      lower = min(conf.low, na.rm = TRUE),
      upper = max(conf.high,na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(term = factor(term, levels = term.order))
  
# get the number of terms
terms <- unique(model.df$term)
n_terms <- length(terms)
  
# Create numeric positions halfway between rows (used for spacing lines between terms)
grid_lines <- seq(1.5, n_terms - 0.5, by = 1)
  
# Define dodge & jitter
dodge <- position_dodge(width = 0.6)
point_dodge_jitter <- position_jitterdodge(
  dodge.width = 0.6,
  jitter.width = 0.35,
  jitter.height = 0
)


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_full_model.plot = ggplot(model.df, aes(x = term, group = period)) +
  
  # Horizontal reference line
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
    
  # Summary error bars with dashed for early, solid for late
  geom_errorbar(data = summary_model.df,
                aes(x = term, y = mean_estimate, ymin = lower, ymax = upper, linetype = period),
                position = dodge, width = 0.3, linewidth = 0.6,
                inherit.aes = FALSE, color = "black") +
    
  # Species-level error bars
  geom_errorbar(aes(y = estimate, ymin = conf.low, ymax = conf.high, color = species),
                position = dodge, width = 0.2, linewidth = 0.5, alpha = 0.5) +
    
  # Species-level points
  geom_point(aes(y = estimate, color = species, shape = period),
             position = point_dodge_jitter, size = 1.8, alpha = 0.7) +
    
  # Shape, color, linetype scales
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  scale_linetype_manual(values = c("early" = "dashed", "late" = "solid")) +
  scale_color_manual(values = c(
    "#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99",
    "#e31a1c", "#fdbf6f", "#ff7f00", "#cab2d6", "#6a3d9a"
  )) +
    
  # Flip axes and label
  coord_flip(ylim = c(-0.75, 0.75)) +
  xlab("") + ylab("Standardized effect") +
    
  # Theme
  plotTheme1 +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 0),
    axis.text.y = element_text(face = "plain"),
    panel.grid.major = element_blank(),
  ) +
    
  # Tidy legend
  guides(color = guide_legend(override.aes = list(alpha = 1)))+
  # add horizontal lines at those positions
  geom_vline(xintercept = grid_lines, linetype = "solid", color = "grey80")
  
#save plot as .pdf
ggsave(results_full_model.plot, file="results_full_model.pdf", path=output.dir,
       width=7, height=3.5)
  
results_full_model.plot
```

</details>

<details>

<summary>Full model alternative</summary>

```{r, warning=FALSE}
#-----------------------------------------------------------
# 1. Loop through 2-level nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.full.alt, function(period_list, period_name) {
    purrr::imap_dfr(period_list, function(model_obj, species_name) {
      if (inherits(model_obj, "brmsfit")) {
        broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
          mutate(
            period = period_name,
            species = species_name,
            .before = term
          )
      } else {
        NULL  # skip if not a brms model
      }
    })
}) %>% 
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
      species = factor(species, levels = rev(sort(unique(species)))), # alphabetical order
      term = plyr::revalue(term, warn_missing=F,
                           c("leaf_out_prev2" = "Leaf-out 2",
                             "LSF4"           = "Late frost",
                             "Tmean.GS"       = "TGS",
                             "leaf_off"       = "Leaf-off",
                             "Tmean.EQ2.SO2"  = "Tmean.Q4",
                             "SWrad"          = "SWR",
                             "Tmean"          = "Tmean")))


#-----------------------------------------------------------
# 2. Summarize data and create plotting helpers
#-----------------------------------------------------------

# Define consistent term order
term.order <- c("Leaf-out 2","Late frost",
                "TGS","Leaf-off","Tmean.Q4",
                "RH","SWR","Tmean")
  
summary_model.df <- model.df %>%
    group_by(term, period) %>%
    summarise(
      mean_estimate = mean(estimate, na.rm = TRUE),
      # 100% of the individual species' 95% CIs fall within this range
      lower = min(conf.low, na.rm = TRUE),
      upper = max(conf.high,na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(term = factor(term, levels = term.order))
  
# First, get the number of terms
terms <- unique(model.df$term)
n_terms <- length(terms)
  
# Create numeric positions halfway between rows (used for spacing lines between terms)
grid_lines <- seq(1.5, n_terms - 0.5, by = 1)
  
# Define dodge & jitter
dodge <- position_dodge(width = 0.6)
point_dodge_jitter <- position_jitterdodge(
  dodge.width = 0.6,
  jitter.width = 0.35,
  jitter.height = 0
)


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_full_model_alternative.plot = ggplot(model.df, aes(x = term, group = period)) +
  # Horizontal reference line
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
    
  # Summary error bars with dashed for early, solid for late
  geom_errorbar(data = summary_model.df,
                aes(x = term, y = mean_estimate, ymin = lower, ymax = upper, linetype = period),
                position = dodge, width = 0.3, linewidth = 0.6,
                inherit.aes = FALSE, color = "black") +
    
  # Species-level error bars
  geom_errorbar(aes(y = estimate, ymin = conf.low, ymax = conf.high, color = species),
                position = dodge, width = 0.2, linewidth = 0.5, alpha = 0.5) +
    
  # Species-level points
  geom_point(aes(y = estimate, color = species, shape = period),
             position = point_dodge_jitter, size = 1.8, alpha = 0.7) +
    
  # Shape, color, linetype scales
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  scale_linetype_manual(values = c("early" = "dashed", "late" = "solid")) +
  scale_color_manual(values = c(
    "#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99",
    "#e31a1c", "#fdbf6f", "#ff7f00", "#cab2d6", "#6a3d9a"
  )) +
    
  # Flip axes and label
  coord_flip(ylim = c(-0.75, 0.75)) +
  xlab("") + ylab("Standardized effect") +
    
  # Theme
  plotTheme1 +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 0),
    axis.text.y = element_text(face = "plain"),
    panel.grid.major = element_blank()
  ) +
    
  # Tidy legend
  guides(color = guide_legend(override.aes = list(alpha = 1)))+
  # add horizontal lines at those positions
  geom_vline(xintercept = grid_lines, linetype = "solid", color = "grey80")
  
#save plot as .pdf
ggsave(results_full_model_alternative.plot, file="results_full_model_alternative.pdf", path=output.dir,
       width=7, height=3.5)
  
results_full_model_alternative.plot
```

</details>

<details>

<summary>Full studentCenter model</summary>

```{r, warning=FALSE}
#-----------------------------------------------------------
# 1. Loop through 2-level nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.full.SC, function(period_list, period_name) {
    purrr::imap_dfr(period_list, function(model_obj, species_name) {
      if (inherits(model_obj, "brmsfit")) {
        broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
          mutate(
            period = period_name,
            species = species_name,
            .before = term
          )
      } else {
        NULL  # skip if not a brms model
      }
    })
}) %>% 
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
      species = factor(species, levels = rev(sort(unique(species)))), # alphabetical order
      term = plyr::revalue(term, warn_missing=F,
                           c("leaf_out_prev_z"  = "Leaf-out",
                             "LSF2"             = "Late frost",
                             "Tmean.EQ1.SO1_z"  = "Tmean.Q2",
                             "Tmean.SO1.EQ2_z"  = "Tmean.Q3",
                             "leaf_off_z"       = "Leaf-off",
                             "Tmean.EQ2.SO2_z"  = "Tmean.Q4",
                             "RH_z"             = "Humidity",
                             "SWrad_z"          = "SWR",
                             "Tmax_z"           = "Tmax")))


#-----------------------------------------------------------
# 2. Summarize data and create plotting helpers
#-----------------------------------------------------------

# Define consistent term order
term.order <- c("Leaf-out","Late frost",
                "Tmean.Q2","Tmean.Q3","Leaf-off","Tmean.Q4",
                "Humidity","SWR","Tmax")

# Summarize data
summary_model.df <- model.df %>%
    group_by(term, period) %>%
    summarise(
      mean_estimate = mean(estimate, na.rm = TRUE),
      # 100% of the individual species' 95% CIs fall within this range
      lower = min(conf.low, na.rm = TRUE),
      upper = max(conf.high,na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(term = factor(term, levels = term.order))
  
# get the number of terms
terms <- unique(model.df$term)
n_terms <- length(terms)
  
# Create numeric positions halfway between rows (used for spacing lines between terms)
grid_lines <- seq(1.5, n_terms - 0.5, by = 1)
  
# Define dodge & jitter
dodge <- position_dodge(width = 0.6)
point_dodge_jitter <- position_jitterdodge(
  dodge.width = 0.6,
  jitter.width = 0.35,
  jitter.height = 0
)


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_full_SC_model.plot = ggplot(model.df, aes(x = term, group = period)) +
  
  # Horizontal reference line
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
    
  # Summary error bars with dashed for early, solid for late
  geom_errorbar(data = summary_model.df,
                aes(x = term, y = mean_estimate, ymin = lower, ymax = upper, linetype = period),
                position = dodge, width = 0.3, linewidth = 0.6,
                inherit.aes = FALSE, color = "black") +
    
  # Species-level error bars
  geom_errorbar(aes(y = estimate, ymin = conf.low, ymax = conf.high, color = species),
                position = dodge, width = 0.2, linewidth = 0.5, alpha = 0.5) +
    
  # Species-level points
  geom_point(aes(y = estimate, color = species, shape = period),
             position = point_dodge_jitter, size = 1.8, alpha = 0.7) +
    
  # Shape, color, linetype scales
  scale_shape_manual(values = c("early" = 16, "late" = 17)) +
  scale_linetype_manual(values = c("early" = "dashed", "late" = "solid")) +
  scale_color_manual(values = c(
    "#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99",
    "#e31a1c", "#fdbf6f", "#ff7f00", "#cab2d6", "#6a3d9a"
  )) +
    
  # Flip axes and label
  coord_flip(ylim = c(-8, 8)) +
  xlab("") + ylab("Standardized effect") +
    
  # Theme
  plotTheme1 +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 0),
    axis.text.y = element_text(face = "plain"),
    panel.grid.major = element_blank(),
  ) +
    
  # Tidy legend
  guides(color = guide_legend(override.aes = list(alpha = 1)))+
  # add horizontal lines at those positions
  geom_vline(xintercept = grid_lines, linetype = "solid", color = "grey80")
  
#save plot as .pdf
ggsave(results_full_SC_model.plot, file="results_full_studentCenter_model.pdf", path=output.dir,
       width=7, height=3.5)
  
results_full_SC_model.plot
```

</details>

<details>

<summary>Monthly model</summary>

```{r}
#-----------------------------------------------------------
# 1. Loop through 2-level nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.monthly, function(period_list, period_name) {
    purrr::imap_dfr(period_list, function(model_obj, species_name) {
      if (inherits(model_obj, "brmsfit")) {
        broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
          mutate(
            period = period_name,
            species = species_name,
            .before = term
          )
      } else {
        NULL  # skip if not a brms model
      }
    })
}) %>% 
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
      species = factor(species, levels = rev(sort(unique(species)))), # alphabetical order
      term = str_remove(term, "[.]"),
      term = readr::parse_number(term))


#-----------------------------------------------------------
# 2. Interpolation of monthly estimates 
#-----------------------------------------------------------

#create identifier
equation.species = unique(model.df$species)

#create interpolation dataframe
df.interp = data.frame()
df.AUC = data.frame()

#loop over variable x equation x species vector
for (species.name in equation.species){

  #subset table
  df.sub = model.df  %>% 
    filter(species == species.name) 
  
  # Interpolate data
  created.interp = lin_interp(df.sub$term, df.sub$term)
  score.interp   = lin_interp(df.sub$term, df.sub$estimate)
  df.interp.sub  = data.frame(created=created.interp, score=score.interp)
  
  # Make a grouping variable for each pos/neg segment
  cat.rle        = rle(df.interp.sub$score < 0)
  df.interp.sub  = df.interp.sub %>%
    mutate(group = rep.int(1:length(cat.rle$lengths),  times=cat.rle$lengths),
           species = unique(df.sub$species))
  #rbind sub dataframes 
  df.interp = rbind(df.interp, df.interp.sub)
  
  #get Area under curve (%)
  df.AUC.sub = df.interp.sub %>%
    mutate(positive = ifelse(score<0, 0, score),
           negative = ifelse(score>0, 0, score))%>%
    summarise(sum.pos = trapz(created, positive), 
              sum.neg = abs(trapz(created, negative)))%>%
    mutate(percent.neg = round(sum.neg/(sum.pos+sum.neg)*100),
           percent.pos = round(sum.pos/(sum.pos+sum.neg)*100),
           species = unique(df.sub$species))
  
  #rbind sub dataframes 
  df.AUC = rbind(df.AUC, df.AUC.sub)
}


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_monthly_model.plot = ggplot() + 
  
  geom_area(data = df.interp, aes(x = created, y = score, fill=score>0, group=group)) + 
  
  scale_fill_manual(values = c('#F21A00', '#3B9AB2'))+
  
  geom_point(data=model.df, aes(x=term, y=estimate), size=0.2)+
  
  geom_errorbar(data=model.df, aes(x=term, ymin=conf.low, ymax=conf.high), width=.2,
                position=position_dodge(.9)) +
  
  geom_hline(yintercept=0)+
  
  geom_vline(xintercept=12.3, linewidth=2, alpha=0.4)+
  
  geom_text(data = df.AUC, mapping = aes(x = -Inf, y = -Inf, hjust = -0.2, vjust = -1.5,
                                         label = paste0(percent.pos,'% / ',percent.neg, '%')))+
  
  coord_cartesian(xlim=c(8.34,15.66))+
  
  xlab("") + ylab("St (days °C-1)")+
  
  scale_x_continuous(breaks = seq(8,16,by=1),
                     labels = c('Aug','Sep','Oct','Nov','Dec','Jan','Feb','Mar','Apr'))+
  
  plotTheme1 + 
  
  facet_wrap(~species, ncol=2)

#save plot as .pdf
ggsave(results_monthly_model.plot, file="results_monthly_model.pdf", path=output.dir,
       width=5, height=10)
  
results_monthly_model.plot
```

</details>

<br>

# 4. Reproducibility

<details>

<summary>show</summary>

```{r}
# Datetime
Sys.time()

# Session info
sessionInfo()
```

</details>
