---
title: "Spring phenology tier2 project"
subtitle: "Moving window figures for fixed preseasons (PEP725 data)"
author: "Constantin Zohner"
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

format:
  html:
    toc: false
    theme: default
    highlight-style: haddock
    df-print: paged
    embed-resources: true

execute:
  #warning: false
  message: false
  echo: true
  output: true

always_allow_html: true
editor: visual
editor_options:
  chunk_output_type: inline
---

<br>

### Model names

-   St_abs: Absolute model (days per °C)
-   St_log: log-log transformation
-   St_z: log-log + z-scaled predictor

<br>

### Preseason definitions

-   PS60: Mean daily maximum temperature 60 days prior to series-level mean leaf-out
-   PSfix: Mean daily maximum temperature 60 days prior to species-level mean leaf-out

<br>

# 1. Setup

<details>

<summary>Packages</summary>

```{r, message = F}
require(tidyverse)
require(data.table)
require(ggplot2)
require(broom)
```

</details>

<details>

<summary>Directories</summary>

```{r}
#-------------------------------------------------
# Define the base directory
#-------------------------------------------------

base.dir <- "/Users/goalie/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis"


#-------------------------------------------------
# Build specific directories from the base
#-------------------------------------------------

# Moving window model results
MW_model.dir    = file.path(base.dir, "Analysis_output/ModelResults/Moving_window_analysis/Individual_files_fixed_preseason_models")

# Output
output.dir = file.path(base.dir, "Analysis_output/ModelFigures/Moving_windows/St_fixed_preseason")
```

</details>

<details>

<summary>Plot theme</summary>

```{r}
plotTheme1 = theme(legend.position   = "right",
                   legend.background = element_rect(fill=NA, linewidth=0.5, linetype="solid"),
                   legend.text       = element_text(color="black",face = "italic"),
                   panel.grid.major  = element_blank(),
                   panel.grid.minor  = element_blank(),
                   panel.background  = element_blank(),
                   panel.border      = element_rect(colour = "black", fill=NA),
                   axis.line         = element_line(color = "black"),
                   axis.text         = element_text(colour = "black"),
                   axis.text.x       = element_text(angle = 45, hjust=1),
                   strip.background  = element_rect(fill=NA),
                   plot.title        = element_text(face="bold"))
```

</details>

<details>

<summary>Functions</summary>

```{r}
#-------------------------------------------------------
# lower (lo) and upper (hi) confidence intervals of mean
#-------------------------------------------------------

lo <- function(x){
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  t.test(x, conf.level = 0.95)$conf.int[1]
}

hi <- function(x){
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  t.test(x, conf.level = 0.95)$conf.int[2]
}


#-------------------------------------------------------
# Standard error of the mean
#-------------------------------------------------------

stderror <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  sd(x) / sqrt(length(x))
}
```

</details>

<details>

<summary>Data</summary>

```{r}
#-------------------------------------------------------
# Moving window model data
#-------------------------------------------------------

model.df <- rbindlist(
  lapply(
    list.files(path = MW_model.dir, pattern = "\\.csv$", full.names = TRUE),
    fread),
  fill = TRUE) %>% 
  
  # transform sign of coefficients
  mutate(estimate = -1 * estimate)
```

</details>

<br>

# 2. Summarize models

<details>

<summary>Show</summary>

```{r, message=F}
#-------------------------------------------------------
# Create tables for absolute and relative metrics
#-------------------------------------------------------

# Absolute metrics (coefficients & correlations)
#-----------------------------------------------
metrics_abs_long <- model.df %>%
  # keep the identifying columns you need
  dplyr::select(
    phenology, equation, species, timeseries, MWyear,
    estimate, cor_xy
  ) %>%
  # pivot the metrics you want to baseline-normalize
  tidyr::pivot_longer(
    cols = c(estimate, cor_xy),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  dplyr::mutate(scale = "absolute")

# Relative metrics (coefficients, SDs, and variance ratios)
#----------------------------------------------------------
metrics_rel_long <- model.df %>%
  # keep the identifying columns you need
  dplyr::select(phenology, equation, species, timeseries, MWyear, 
                estimate, sd_x, sd_y, var_ratio) %>%
  # pivot the metrics you want to baseline-normalize
  tidyr::pivot_longer(
    cols = c(estimate, sd_x, sd_y, var_ratio),
    names_to  = "metric",
    values_to = "value_abs"
  ) %>%
  # create baseline and calculate relative metric
  group_by(phenology, equation, species, metric) %>%
  mutate(
    base  = mean(value_abs, na.rm = TRUE),
    value = 100 * (value_abs / base - 1)
  ) %>% 
  ungroup() %>% 
  dplyr::select(-value_abs, -base)%>%
  dplyr::mutate(
    metric = paste0(metric, "_rel"),
    scale = "relative")

# Anomaly metrics (leaf-out and temperature means)
#----------------------------------------------------------
metrics_anom_long <- model.df %>%
  # keep the identifying columns you need
  dplyr::select(phenology, equation, species, timeseries, MWyear, 
                mean_x, mean_y) %>%
  
  # pivot the metrics you want to baseline-normalize
  tidyr::pivot_longer(
    cols = c(mean_x, mean_y),
    names_to  = "metric",
    values_to = "value_abs"
  ) %>%
  # create baseline and calculate anomaly
  group_by(phenology, equation, species, metric) %>%
  mutate(
    base  = mean(value_abs, na.rm = TRUE),
    value = value_abs - base
  ) %>%
  ungroup() %>% 
  dplyr::select(-value_abs, -base)%>%
  dplyr::mutate(
    metric = paste0(metric, "_anom"),
    scale = "anomaly")

# Bind everything into one long table
#------------------------------------
metrics_long <- dplyr::bind_rows(
  metrics_abs_long,
  metrics_rel_long,
  metrics_anom_long
)


#-------------------------------------------------------
# 3. Summarize at species × MWyear × metric
#-------------------------------------------------------

model.summary.df <- metrics_long %>%
  group_by(phenology, equation, species, MWyear, metric, scale) %>%
  summarize(
    mean = mean(value, na.rm = TRUE),
    lo   = lo(value),
    hi   = hi(value),
    .groups = "drop"
  )


#-------------------------------------------------------
# 4. Summarize at MWyear × metric
#-------------------------------------------------------

model.summary.all.df <- model.summary.df %>%
  group_by(phenology, equation, MWyear, metric, scale) %>%
  summarize(
    mean_all = mean(mean, na.rm = TRUE),
    lo       = lo(mean),
    hi       = hi(mean),
    .groups  = "drop"
  )
```

</details>

<br>

# 3. Leaf-out model plots

<details>

<summary>Model coefficients</summary>

```{r, warning=F, message=F}
#-------------------------------------------------------
# Linear model coefficients of metric change over time
#-------------------------------------------------------

linear.trends.df <- model.summary.all.df %>%
  
  # 1) Define the grouping for which you want *one* trend each
  group_by(phenology, equation, metric) %>%
  
  # 2) Fit one linear model per group and store it as a list-column.
  summarise(
    model = list(lm(mean_all ~ MWyear, data = pick(MWyear, mean_all))),
    .groups = "drop"   # drop grouping after summarise
  ) %>%
  
  # 3) Extract model results into two standard broom tables:
  mutate(tidied  = purrr::map(model, broom::tidy)) %>%
  dplyr::select(-model) %>%   # We no longer need the raw model objects
  
  # 4) Expand the "tidied" list-column so you get one row per model term
  tidyr::unnest(tidied) %>%
  
  # 5) Keep only the slope term (drop the intercept row)
  filter(term == "MWyear") %>%
  
  # 6) Rename to your preferred names
  rename(slope = estimate, se = std.error, pval = p.value) %>%
  
  # 7) Build an annotation label (and provide x/y placement defaults for geom_text)
  mutate(
    
    # significance labeling
    p_lab = case_when(
      pval < 0.001 ~ "***",
      pval < 0.01  ~ "**",
      pval < 0.05  ~ "*",
      TRUE         ~ sprintf("p = %.2f", pval)
    ),
    
    # ---- Units logic ----
    unit = dplyr::case_when(
      metric == "estimate" ~ "St per year",
      metric == "cor_xy"   ~ "per year",
      
      # Any metric name containing "rel" -> percent per year
      stringr::str_detect(metric, "rel") ~ "% per year",
      
      # mean_x_anom special-cases
      metric == "mean_x_anom" & equation == "St_abs" ~ "\u00B0C per year",
      metric == "mean_x_anom" & equation != "St_abs" ~ "per year",
      
      # mean_y_anom special-cases
      metric == "mean_y_anom" & equation == "St_abs" ~ "days per year",
      metric == "mean_y_anom" & equation != "St_abs" ~ "per year",
      
      # Fallback (in case new metrics appear)
      TRUE ~ "per year"
    ),
    
    # label text shown on plots
    label = sprintf("trend: %.3f \u00B1 %.3f %s %s", slope, 2 * se, unit, p_lab)
  ) %>%
  
  # 8) Keep only what you need for joining/filtering + plotting annotation
  dplyr::select(phenology, equation, metric, slope, se, pval, label)

linear.trends.df
```

</details>

<details>

<summary>Relative Ts change (Fig. Sx)</summary>

```{r, warning=F, message=F}
#===============================================================================
# Relative decline in temperature sensitivity (St) from early → late moving windows
#   - Computes mean St per species × equation × phenology for:
#       (i) earliest 15 moving windows  = "early"
#      (ii) latest   15 moving windows  = "late"
#   - Returns relative decline (%) = 100 * (St_early - St_late) / St_early
#   - Plots distributions across species as violin + boxplot + jittered points
#===============================================================================


#------------------------------------------------------
# 1) Define time windows
#------------------------------------------------------

mw_min <- min(model.summary.df$MWyear, na.rm = TRUE)
mw_max <- max(model.summary.df$MWyear, na.rm = TRUE)

early <- mw_min:(mw_min + 14)   # earliest 15 moving windows
late  <- (mw_max - 14):mw_max   # latest 15 moving windows


#------------------------------------------------------
# 2) Phenology model ordering + plot labels + colors
#------------------------------------------------------

phen_order <- c(
  "leaf_out",
  "leaf_out_GDD",
  "leaf_out_GDD.chill.lo",
  "leaf_out_GDD.chill.hi",
  "leaf_out_GDD.chill.DL"
)

phen_labels <- c(
  leaf_out              = "Observations",
  leaf_out_GDD          = "Forcing-only",
  leaf_out_GDD.chill.lo = "Low chilling",
  leaf_out_GDD.chill.hi = "High chilling",
  leaf_out_GDD.chill.DL = "High chilling + Daylength"
)


#------------------------------------------------------
# 3) Compute relative decline in St (species-level)
#------------------------------------------------------

relative.Ts.df <- model.summary.df %>%
  # keep only early/late windows, target equations, and target model variants
  filter(
    MWyear %in% c(early, late),
    metric == "estimate",
    phenology %in% phen_order
  ) %>%
  # classify MWyear into early vs late
  mutate(period = ifelse(MWyear %in% early, "early", "late")) %>%
  # average across the 15 moving windows within each period
  group_by(species, equation, phenology, period) %>%
  summarise(St = mean(mean, na.rm = TRUE), .groups = "drop") %>%
  # reshape to have columns early and late
  pivot_wider(names_from = period, values_from = St) %>%
  # drop incomplete pairs (and avoid division by zero)
  filter(!is.na(early), !is.na(late), early != 0) %>%
  mutate(
  change_pct = 100 * (late / early -1),  # 0 = no change; + = increase; - = decrease
  # enforce consistent ordering in plot
  phenology = factor(phenology, levels = phen_order),
  equation  = factor(equation, levels = c("St_PS60_abs", "St_PS60_log", "St_PS60_z",
                                          "St_PSfix_abs", "St_PSfix_log", "St_PSfix_z"))
  ) %>%
  arrange(species, equation, phenology)

# panel letters (placed manually in the upper-left of each facet)
panel_labs <- tibble(
  equation = factor(c("St_PS60_abs", "St_PS60_log", "St_PS60_z",
                      "St_PSfix_abs", "St_PSfix_log", "St_PSfix_z"), 
                    levels = levels(relative.Ts.df$equation)),
  lab      = c("a", "b", "c", "d", "e", "f")
)


#------------------------------------------------------
# 4) Plot: box + species points, faceted by equation
#------------------------------------------------------

plot.relative.St <- ggplot(relative.Ts.df, aes(x = phenology, y = change_pct)) +

  # Reference: 100% = no change
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +

  # Make all distributions grey (no phenology fill)
  #geom_violin(trim = TRUE, alpha = 0.85, fill = "grey85", color = NA) +
  geom_boxplot(width = 0.4, outlier.shape = NA, 
               fill = "grey90", color = "grey30") +

  # Color points by species (manual palette)
  geom_point(
  aes(color = species),
  position = position_jitter(width = 0.18, height = 0),
  size = 1.5, alpha = 0.85) +

  facet_wrap(~ equation, nrow = 2, scales = "fixed") +

  # Panel letters a / b (upper-left)
  geom_text(
    data = panel_labs,
    aes(x = -Inf, y = Inf, label = lab),
    inherit.aes = FALSE,
    hjust = -0.2, vjust = 1.2, fontface = "bold", size = 5
  ) +

  # Rename phenology classes on x-axis
  scale_x_discrete(labels = phen_labels) +

  # Manual species color palette (10 species)
  scale_color_manual(
    "Species",
    values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
               '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a')
  ) +

  labs(x = NULL, y = "Change in St relative to early period (%)") +
  coord_cartesian(ylim = c(-50, 40)) +
  plotTheme1 +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 25, hjust = 1),
    panel.spacing.x = unit(0.8, "lines"),
    legend.position = "right"
  )
plot.relative.St

# Save plot
ggsave(plot.relative.St, file="FigSx_Relative_St_decline_fixed_preseason.pdf", path=output.dir,
       width=9, height=6)
```

</details>

<details>

<summary>Relative St and variance ratio plots (Fig. Sx)</summary>

```{r, warning=F, message=F}
#------------------------------------------------------------
# Relative St and variance ratio plots
#------------------------------------------------------------


# equations
equations = c("St_PS60_abs",
              "St_PS60_log",
              "St_PS60_z",
              
              "St_PSfix_abs",
              "St_PSfix_log",
              "St_PSfix_z")

# Define desired phenology order
phenology_levels <- c("leaf_out",
                      "leaf_out_GDD", 
                      "leaf_out_GDD.chill.lo", 
                      "leaf_out_GDD.chill.hi", 
                      "leaf_out_GDD.chill.DL")


#------------------------------------------------------------
# Loop over equations
#------------------------------------------------------------

for (equation.name in equations) {
  
  # Create data
  #------------
  # Model summay data
  St.summary.all.df = model.summary.all.df %>% 
    filter(
      equation == equation.name,
      metric %in% c("estimate_rel","cor_xy",
                    "sd_x_rel","var_ratio_rel")) %>% 
    mutate(phenology = factor(phenology, levels = phenology_levels))
  
  # Linear trend (slope ± 2*SE)
  ann_slope = linear.trends.df %>% 
    filter(
      equation == equation.name,
      metric %in% c("estimate_rel","cor_xy",
                    "sd_x_rel","var_ratio_rel")) %>% 
    mutate(x = -Inf, y = -Inf,
           phenology = factor(phenology, levels = phenology_levels))
  
  
  # Create plot
  #------------
  plot = model.summary.df %>% 
    filter(
      equation == equation.name,
      metric %in% c("estimate_rel","cor_xy",
                    "sd_x_rel","var_ratio_rel")) %>% 
  
    dplyr::mutate(phenology = factor(phenology, levels = phenology_levels)) %>% # sort factors
  
    ggplot() + 
    
    geom_hline(yintercept = 0) +
  
    # Confidence ribbon for overall trend
    geom_ribbon(data=St.summary.all.df,
                aes(x = MWyear, y = mean_all, ymin = lo, ymax = hi),
                fill = "darkgrey", color=NA, alpha = 0.9) + 
  
    # Individual species lines (filtered from pipe input)
    geom_line(aes(x = MWyear, y = mean, group=species, color=species), linewidth = 0.4) +
  
    # Linear trend over the aggregated series (dashed)
    geom_smooth(data = St.summary.all.df,
                aes(x = MWyear, y = mean_all),
                method='lm', formula = y~x, se = FALSE, 
                linewidth=1, linetype="dashed", color="black") +
  
    # Smoothed line: overall trend (black line)
    geom_line(data=St.summary.all.df, aes(x = MWyear, y = mean_all),
              color="black", size=1) +
  
    # Manual species color palette
    scale_color_manual("Species", 
                       values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                                  '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a'))+
  
    # Axis limits: restrict x-range to window years; y-range customized
    coord_cartesian(xlim=c(1967.9,2007.1))+
  
    xlab("") + ylab(expression("Relative change"))+
  
    # X-axis breaks for moving-window labels (left blank intentionally)
    scale_x_continuous(breaks = seq(1966,2008,by=5),
                       labels = c('1966-1980','1971-1985','1976-1990',
                                  '1981-1995','1986-2000','1991-2005',
                                  '1996-2010','2001-2015','2006-2020')) +
  
    # Slope annotation
    geom_text(
    data = ann_slope,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    hjust = -0.05, vjust = -1, size = 3.5) +
  
    facet_grid(metric~phenology, scales = "free_y") +
  
    # Plot theme and legend
    plotTheme1

  print(plot)
  
  # Fig. Sx-y: Relative St + Variance ratio plots
  ggsave(plot, 
       file=paste0("FigSX_VarianceRatios_", equation.name, ".pdf"), 
       path=output.dir,
       width=14, height=8)
  
}
```

</details>

<br>

# 4. Reproducibility

<details>

<summary>Show</summary>

```{r}
# Datetime
Sys.time()

# Session info
sessionInfo()
```

</details>
