---
title: "Spring phenology tier2 project"
subtitle: "Moving window figures (PEP725 data)"
author: "Constantin Zohner"
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

format:
  html:
    toc: false
    theme: default
    highlight-style: haddock
    df-print: paged
    embed-resources: true

execute:
  #warning: false
  message: false
  echo: true
  output: true

always_allow_html: true
editor: visual
editor_options:
  chunk_output_type: inline
---

<br>

### Models

-   St: Univariate models using Tmax (daily maximum temperature) as predictor
    -   abs: absolute sensitivity (days per °C)
    -   log: log-log transformed
    -   z: log-log and z-scaled predictor \[log(days) per SD log(temp)\]
-   Multivariate model with scaled predictor and response variables

<br>

# 1. Setup

<details>

<summary>Packages</summary>

```{r, message = F}
require(tidyverse)
require(data.table)
require(ggplot2)
require(patchwork)
require(wesanderson)
require(pracma)
require(ggpubr)
require(broom)
require(segmented)
require(lme4)
```

</details>

<details>

<summary>Directories</summary>

```{r}
#-------------------------------------------------
# Define the base directory
#-------------------------------------------------

base.dir <- "/Users/goalie/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis"


#-------------------------------------------------
# Build specific directories from the base
#-------------------------------------------------

# PEP moving window data
PEP.dir         = file.path(base.dir, "Analysis_input/PEP_drivers_final_spring_MW/MW15/Merged_file")

# Spring climate
Spring_clim.dir = file.path(base.dir, "Analysis_input/Spring_climate")

# Moving window model results
MW_model.dir    = file.path(base.dir, "Analysis_output/ModelResults/Moving_window_analysis/Individual_files_models")

# Moving window mixed model results
MW_mixed_model.dir = file.path(base.dir, "Analysis_output/ModelResults/Moving_window_analysis/Individual_files_mixed_models")

# Output
output.dir = file.path(base.dir, "Analysis_output/ModelFigures/Moving_windows")
```

</details>

<details>

<summary>Plot theme</summary>

```{r}
plotTheme1 = theme(legend.position   = "right",
                   legend.background = element_rect(fill=NA, linewidth=0.5, linetype="solid"),
                   legend.text       = element_text(color="black",face = "italic"),
                   panel.grid.major  = element_blank(),
                   panel.grid.minor  = element_blank(),
                   panel.background  = element_blank(),
                   panel.border      = element_rect(colour = "black", fill=NA),
                   axis.line         = element_line(color = "black"),
                   axis.text         = element_text(colour = "black"),
                   axis.text.x       = element_text(angle = 45, hjust=1),
                   strip.background  = element_rect(fill=NA),
                   plot.title        = element_text(face="bold"))
```

</details>

<details>

<summary>Functions</summary>

```{r}
#-------------------------------------------------------
# lower (lo) and upper (hi) confidence intervals of mean
#-------------------------------------------------------

lo <- function(x){
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  t.test(x, conf.level = 0.95)$conf.int[1]
}

hi <- function(x){
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  t.test(x, conf.level = 0.95)$conf.int[2]
}


#-------------------------------------------------------
# Standard error of the mean
#-------------------------------------------------------

stderror <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  sd(x) / sqrt(length(x))
}
```

</details>

<br>

# 2. Load and summarize data

<details>

<summary>Leaf-out observations and simulations and spring data</summary>

```{r, message=F}
#-------------------------------------------------------
# 1. PEP725 moving window data
#-------------------------------------------------------

PEP.df <- fread(paste(PEP.dir, "pep_MW15_merged_final.csv", sep="/")) %>% 
  #species-level leaf-out and temperature anomalies
  group_by(phenology, species) %>% 
  mutate(leafout.anomaly = leaf_out - mean(leaf_out)) %>% 
  ungroup() %>% 
  #select columns
  dplyr::select(c(phenology, species, s_id, timeseries, year, MWyear,
                  leafout.anomaly))


#-------------------------------------------------------
# 2. Spring climate data
#-------------------------------------------------------

Spring_clim.df <- fread(paste(Spring_clim.dir, "pep_drivers_spring_climate.csv", sep="/")) %>% 
  #delete NAs
  filter(!is.na(Tmax.spring60)) 

# Moving-window transformation

# Year vector
year.vector  = c(min(Spring_clim.df$year):(max(Spring_clim.df$year)-14))
# Counter
i=1
# Create List object
List.year = replicate(length(year.vector), data.frame())
names(List.year) = year.vector

# Loop to bind individual moving-window data frames
for(MW.year in year.vector) {
  
  Spring_clim.df.sub = Spring_clim.df %>% 
    # filter years within 15-year moving window
    filter(year >= MW.year,
           year < MW.year + 15) %>% 
    # Group by timeseries
    group_by(timeseries)%>%
    # Delete timeseries <12 years
    filter(n() >= 12) %>%
    ungroup() %>% 
    mutate(MWyear = MW.year,
           ts_yr  = paste(timeseries,year,MW.year, sep="_")) 
  
  # store dataframe in year list
  List.year[[i]] = Spring_clim.df.sub
  
  # Increase counter and print
  i=i+1
  #print(MW.year)
}  

# Merge list to data frame
Spring_clim.MW.df = do.call(rbind.data.frame, List.year)

# Add data
Spring_clim.MW.df = Spring_clim.MW.df %>% 
  #species-level temperature anomalies
  group_by(species) %>% 
  mutate(Tmax.spring60 = Tmax.spring60 - mean(Tmax.spring60)) %>% 
  ungroup() %>% 
  #select columns
  dplyr::select(c(species, s_id, timeseries, year, MWyear,
                  Tmax.spring60))


#-------------------------------------------------------
# 3. Merge PEP and spring climate data
#-------------------------------------------------------

PEP.df = PEP.df %>% 
  left_join(., Spring_clim.MW.df, by = join_by(species, s_id, timeseries, year, MWyear))

rm(Spring_clim.MW.df, Spring_clim.df, Spring_clim.df.sub)


#-------------------------------------------------------
# 4. Summarise data
#-------------------------------------------------------

VariablesVector = c("leafout.anomaly", "Tmax.spring60")
summary.df = PEP.df %>% 
  # get mean and sd of leaf-out and Tmax for each timeseries x MW combination
  group_by(phenology, species, timeseries, MWyear) %>% 
  summarize_at(VariablesVector, tibble::lst(mean)) %>% 
  # long format
  pivot_longer(-c(phenology, species, timeseries, MWyear), names_to = "variable", values_to = "value") %>% 
  # get means per species X MW combination
  group_by(phenology, species, MWyear, variable) %>% 
  summarize(value = mean(value, na.rm = TRUE))


#-------------------------------------------------------
# 5. Summarise species-level results
#-------------------------------------------------------

summary.all.df = summary.df %>% 
  group_by(phenology, MWyear, variable) %>% 
  summarize(mean = mean(value),
            lo   = lo(value),
            hi   = hi(value), 
            .groups="drop")
```

</details>

<details>

<summary>Univariate models</summary>

```{r, message=F}
#-------------------------------------------------------
# 1. Load moving window model data
#-------------------------------------------------------

model.df <- rbindlist(
  lapply(
    list.files(path = MW_model.dir, pattern = "\\.csv$", full.names = TRUE),
    fread),
  fill = TRUE) %>% 
  
  # transform sign of coefficients
  mutate(estimate = -1 * estimate)


#-------------------------------------------------------
# 2. Create tables for absolute and relative metrics
#-------------------------------------------------------

# Absolute metrics (coefficients & correlations)
#-----------------------------------------------
metrics_abs_long <- model.df %>%
  # keep the identifying columns you need
  dplyr::select(
    phenology, equation, species, timeseries, MWyear,
    estimate, cor_xy
  ) %>%
  # pivot the metrics you want to baseline-normalize
  tidyr::pivot_longer(
    cols = c(estimate, cor_xy),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  dplyr::mutate(scale = "absolute")

# Relative metrics (coefficients, SDs, and variance ratios)
#----------------------------------------------------------
metrics_rel_long <- model.df %>%
  # keep the identifying columns you need
  dplyr::select(phenology, equation, species, timeseries, MWyear, 
                estimate, sd_x, sd_y, var_ratio) %>%
  # pivot the metrics you want to baseline-normalize
  tidyr::pivot_longer(
    cols = c(estimate, sd_x, sd_y, var_ratio),
    names_to  = "metric",
    values_to = "value_abs"
  ) %>%
  # create baseline and calculate relative metric
  group_by(phenology, equation, species, metric) %>%
  mutate(
    base  = mean(value_abs, na.rm = TRUE),
    value = 100 * (value_abs / base - 1)
  ) %>% 
  ungroup() %>% 
  dplyr::select(-value_abs, -base)%>%
  dplyr::mutate(
    metric = paste0(metric, "_rel"),
    scale = "relative")

# Anomaly metrics (leaf-out and temperature means)
#----------------------------------------------------------
metrics_anom_long <- model.df %>%
  # keep the identifying columns you need
  dplyr::select(phenology, equation, species, timeseries, MWyear, 
                mean_x, mean_y) %>%
  
  # pivot the metrics you want to baseline-normalize
  tidyr::pivot_longer(
    cols = c(mean_x, mean_y),
    names_to  = "metric",
    values_to = "value_abs"
  ) %>%
  # create baseline and calculate anomaly
  group_by(phenology, equation, species, metric) %>%
  mutate(
    base  = mean(value_abs, na.rm = TRUE),
    value = value_abs - base
  ) %>%
  ungroup() %>% 
  dplyr::select(-value_abs, -base)%>%
  dplyr::mutate(
    metric = paste0(metric, "_anom"),
    scale = "anomaly")

# Bind everything into one long table
#------------------------------------
metrics_long <- dplyr::bind_rows(
  metrics_abs_long,
  metrics_rel_long,
  metrics_anom_long
)


#-------------------------------------------------------
# 3. Summarize at species × MWyear × metric
#-------------------------------------------------------

model.summary.df <- metrics_long %>%
  group_by(phenology, equation, species, MWyear, metric, scale) %>%
  summarize(
    mean = mean(value, na.rm = TRUE),
    lo   = lo(value),
    hi   = hi(value),
    .groups = "drop"
  )


#-------------------------------------------------------
# 4. Summarize at MWyear × metric
#-------------------------------------------------------

model.summary.all.df <- model.summary.df %>%
  group_by(phenology, equation, MWyear, metric, scale) %>%
  summarize(
    mean_all = mean(mean, na.rm = TRUE),
    lo       = lo(mean),
    hi       = hi(mean),
    .groups  = "drop"
  )
```

</details>

<details>

<summary>Multivariate mixed model</summary>

```{r, message=F}
#-------------------------------------------------------
# 1. Load data
#-------------------------------------------------------

mixed.model.df = rbindlist(lapply(list.files(path = MW_mixed_model.dir), 
                            function(n) fread(file.path(MW_mixed_model.dir, n))),fill=TRUE) %>% 
  # Rename variables
  mutate(term = plyr::revalue(term, warn_missing=F,
                              c("leaf_out_prev"  = "Leaf-out",
                                "Tmean.EQ2.SO2"  = "T[Q4]",
                                "SWrad"          = "Radiation",
                                "Tmax"           = "T[max]")),
         
         term = factor(term, levels = c("T[max]","Radiation","T[Q4]","Leaf-out")))


#-------------------------------------------------------
# 2. Summarise species-level mixed model results
#-------------------------------------------------------

mixed.model.summary.df = mixed.model.df %>% 
  group_by(phenology, term, MWyear) %>% 
  summarize(
    # summarize model coefficients
    coeff.mean = mean(estimate),
    coeff.lo   = lo(estimate),
    coeff.hi   = hi(estimate) )
```

</details>

<br>

# 3. Leaf-out date and climate trend plots

<details>

<summary>Model coefficients and breakpoints</summary>

```{r}
#-------------------------------------------------------
# Linear model coefficients of variable change over time
#-------------------------------------------------------

linear.trends.df <- summary.all.df %>%
  
  # 1) Define the grouping for which you want *one* trend each
  group_by(phenology, variable) %>%
  
  # 2) Fit one linear model per group and store it as a list-column.
  summarise(
    model = list(lm(mean ~ MWyear, data = pick(MWyear, mean))),
    .groups = "drop"   # drop grouping after summarise
  ) %>%
  
  # 3) Extract model results into standard broom table:
  mutate(tidied  = purrr::map(model, broom::tidy)) %>%
  dplyr::select(-model) %>%   # We no longer need the raw model objects
  
  # 4) Expand the "tidied" list-column so you get one row per model term
  tidyr::unnest(tidied) %>%
  
  # 5) Keep only the slope term (drop the intercept row)
  filter(term == "MWyear") %>%
  
  # 6) Rename to your preferred names
  rename(slope = estimate, se = std.error, pval = p.value) %>%
  
  # 7) Build an annotation label (and provide x/y placement defaults for geom_text)
  mutate(
    
    # significance labeling
    p_lab = case_when(
      pval < 0.001 ~ "***",
      pval < 0.01  ~ "**",
      pval < 0.05  ~ "*",
      TRUE         ~ sprintf("p = %.2f", pval)
    ),
    
    # ---- Units logic ----
    unit = dplyr::case_when(
      variable == "leafout.anomaly_mean" ~ "days per year",
      variable == "Tmax.spring60_mean"   ~ "\u00B0C per year",
      
      # Fallback (in case new metrics appear)
      TRUE ~ "per year"
    ),
    
    # label text shown on plots
    label = sprintf("trend: %.3f \u00B1 %.3f %s %s", slope, 2 * se, unit, p_lab)
  ) %>%
  
  # 8) Keep only what you need for joining/filtering + plotting annotation
  dplyr::select(phenology, variable, slope, se, pval, label)

linear.trends.df


#-------------------------------------------------------
# Breakpoints
#-------------------------------------------------------

BP.results.df = summary.df %>% 
  
  # group data
  group_by(phenology, variable) %>%  
  
  do({
    
    #run linear model
    linear.mod    = lm(value ~ MWyear, data=.)
    
    #run breakpoint model
    segmented.mod = segmented(linear.mod, seg.Z=~MWyear, psi=list(MWyear = c(1970, 1988)), 
                              seg.control(n.boot=100, fix.npsi=F))
    
    # get the fitted data
    my.fitted = fitted(segmented.mod)
    
    # extract breakpoints in case model predicts them
    if(length(segmented.mod$coefficients)>2){
      
      # get the breakpoints
      breakpoint = data.frame(segmented.mod$psi) %>% 
        mutate(Est. = round(Est.)) 
      
      # get vector of breakpint numbers
      bp.vec = c(rep(1, length(min(.$MWyear):breakpoint[1,]$Est.)),
                 rep(2, length((breakpoint[1,]$Est.+1):breakpoint[2,]$Est.)),
                 rep(3, length((breakpoint[2,]$Est.+1):max(.$MWyear))))
      
      #create empty data if no breakpoints were estimated
    } else {
      breakpoint = data.frame(Initital=c(NA,NA), Est.=c(NA,NA) )
      bp.vec     = c(rep(1, length(min(.$MWyear):1980)),
                     rep(2, length(1981:1990)),
                     rep(3, length(1991:max(.$MWyear))))
    }
    
    # combine data
    data.frame(MWyear      = unique(.$MWyear), 
               breakpoint  = bp.vec,
               breakpoint1 = breakpoint[1,]$Est.,
               breakpoint2 = breakpoint[2,]$Est.,
               breakpoint1.std.err = breakpoint[1,]$St.Err,
               breakpoint2.std.err = breakpoint[2,]$St.Err)
  })

# Get mean breakpoint years and confidence range across species
BreakYears.df = BP.results.df %>% 
  distinct(phenology, variable, .keep_all = T) %>% 
  dplyr::select(phenology, variable, breakpoint1, breakpoint2, breakpoint1.std.err, breakpoint2.std.err)

# Summarize coefficients and slopes for each breakpoint
BP.coeff.all.df = summary.df %>% left_join(., 
                                           BP.results.df %>% dplyr::select(phenology, MWyear, variable, breakpoint), 
                                           by=c("phenology", "MWyear", "variable")) %>% 
  #summarize for each species
  group_by(phenology, variable, breakpoint, species) %>% 
  summarize_at("value", tibble::lst(coeff = ~ mean(., na.rm=T)) ) %>% 
  #summarize across species
  group_by(phenology, variable, breakpoint) %>% 
  summarize_at("coeff", tibble::lst(mean.coeff = ~ mean(., na.rm=T), se.coeff = stderror) )
```

</details>

<details>

<summary>Leaf-out anomaly plots (Figs. 2b and 3a)</summary>

```{r}
#-------------------------------------------------------
# Leaf-out anomaly plot (Fig. 2b)
#-------------------------------------------------------

# Filter datasets once using phenology + variable names
Pheno_BreakYears.df = BreakYears.df %>% 
  filter(phenology == "leaf_out",
         variable == "leafout.anomaly_mean")

Pheno_summary.all.df = summary.all.df %>% 
  filter(phenology == "leaf_out",
         variable == "leafout.anomaly_mean")

Pheno_BP.coeff.all.df = BP.coeff.all.df %>% 
  filter(phenology == "leaf_out",
         variable == "leafout.anomaly_mean")

# --- Linear trend (slope ± 2*SE)
ann_slope <- linear.trends.df %>% 
  filter(phenology == "leaf_out",
         variable == "leafout.anomaly_mean") %>% 
  mutate(x = Inf, y = Inf)

# Plotting: filter main data and construct ggplot
#------------------------------------------------
plot.leafout = summary.df %>% 
  
  # Keep only relevant phenology & temperature variables
  filter(phenology == "leaf_out",
         variable == "leafout.anomaly_mean") %>% 
  
  ggplot() + 
  
   # Horizontal reference line at y=0
  geom_hline(yintercept=0) +
  
  # Add shaded rectangles for breakpoint confidence intervals (breakpoint 1 & 2)
  geom_rect(data=Pheno_BreakYears.df, 
            mapping=aes(xmin=breakpoint1 - 2*breakpoint1.std.err, 
                        xmax=breakpoint1 + 2*breakpoint1.std.err, ymin=-Inf, ymax=Inf), 
            color=NA, alpha=0.2, fill="black") +
  geom_rect(data=Pheno_BreakYears.df, 
            mapping=aes(xmin=breakpoint2-2*breakpoint2.std.err, 
                        xmax=breakpoint2+2*breakpoint2.std.err, ymin=-Inf, ymax=Inf), 
            color=NA, alpha=0.2, fill="black") +
  
  # Add dashed vertical lines at breakpoints
  geom_vline(data=Pheno_BreakYears.df, aes(xintercept=breakpoint1), color="black", lty="dashed") +
  geom_vline(data=Pheno_BreakYears.df, aes(xintercept=breakpoint2), color="black", lty="dashed") +
  
  # Add ribbon for CI of smoothed mean trend
  geom_ribbon(data=Pheno_summary.all.df,
              aes(x = MWyear, y = mean, ymin = lo, ymax = hi),
              fill = "darkgrey", color=NA, alpha = 0.9) + 
  
  # Add individual species time series
  geom_line(aes(x = MWyear, y = value, group=species, color=species), linewidth = 0.4) +
  
  # Add overall mean trend line
  geom_line(data=Pheno_summary.all.df,
            aes(x = MWyear, y = mean), color="black", linewidth=1) +
  
  # Manual color palette for species
  scale_color_manual("Species", 
                     values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                                '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a'))+
  
  # Zoom in on period of interest
  coord_cartesian(xlim=c(1967.9,2007.1))+
  
  # Axis labels
  xlab("") + ylab('Leaf-out anomaly (days)')+
  
  # Custom x-axis labels showing moving windows
  scale_x_continuous(breaks = seq(1966,2008,by=5),
                     labels = c('','','','','','','','','')) +
  
  # Add breakpoint coefficient labels (mean ± 2*SE) for each breakpoint
  geom_text(data = Pheno_BP.coeff.all.df %>% filter(breakpoint == 1),
            aes(x = 1967, y = -Inf, 
                label = paste0(round(mean.coeff, 1), " ± ", round(2*se.coeff, 1))),
            hjust = 0, vjust = -.5, size=3.5) +
  geom_text(data = Pheno_BP.coeff.all.df %>% filter(breakpoint == 2),
            aes(x = 1981, y = -Inf, 
                label = paste0(round(mean.coeff, 1), " ± ", round(2*se.coeff, 1))),
            hjust = 0, vjust = -.5, size=3.5) +
  geom_text(data = Pheno_BP.coeff.all.df %>% filter(breakpoint == 3),
            aes(x = 1995, y = -Inf, 
                label = paste0(round(mean.coeff, 1), " ± ", round(2*se.coeff, 1))),
            hjust = 0, vjust = -.5, size=3.5) +
  
  # Slope annotation
  geom_text(
  data = ann_slope,
  aes(x = x, y = y, label = label),
  inherit.aes = FALSE,
  hjust = 1.1, vjust = 2, size = 3.5) +
  
  # Apply predefined theme
  plotTheme1

plot.leafout


#-------------------------------------------------------
# Leaf-out simulation anomaly plots (Fig. 3a)
#-------------------------------------------------------

# Define desired phenology order
phenology_levels <- c("leaf_out_GDD", 
                      "leaf_out_GDD.chill.lo", 
                      "leaf_out_GDD.chill.hi", 
                      "leaf_out_GDD.chill.DL")

# Filter datasets once using phenology + variable names
Pheno2_BreakYears.df = BreakYears.df %>% 
  filter(phenology != "leaf_out",
         variable == "leafout.anomaly_mean") %>%
  mutate(phenology = factor(phenology, levels = phenology_levels))

Pheno2_summary.all.df = summary.all.df %>% 
  filter(phenology != "leaf_out",
         variable == "leafout.anomaly_mean") %>%
  mutate(phenology = factor(phenology, levels = phenology_levels))

Pheno2_BP.coeff.all.df = BP.coeff.all.df %>% 
  filter(phenology != "leaf_out",
         variable == "leafout.anomaly_mean") %>%
  mutate(phenology = factor(phenology, levels = phenology_levels))

# --- Linear trend (slope ± 2*SE)
ann_slope <- linear.trends.df %>% 
  filter(phenology != "leaf_out",
         variable == "leafout.anomaly_mean") %>% 
  mutate(x = Inf, y = Inf,
         phenology = factor(phenology, levels = phenology_levels))

# Plotting: filter main data and construct ggplot
#------------------------------------------------
plot.leafout.simulation = summary.df %>% 
  
  # Keep only relevant phenology & temperature variables
  filter(phenology != "leaf_out",
         variable == "leafout.anomaly_mean") %>%
  mutate(phenology = factor(phenology, levels = phenology_levels)) %>% 
  
  ggplot() + 
  
   # Horizontal reference line at y=0
  geom_hline(yintercept=0) +
  
  # Add shaded rectangles for breakpoint confidence intervals (breakpoint 1 & 2)
  geom_rect(data=Pheno2_BreakYears.df, 
            mapping=aes(xmin=breakpoint1 - 2*breakpoint1.std.err, 
                        xmax=breakpoint1 + 2*breakpoint1.std.err, ymin=-Inf, ymax=Inf), 
            color=NA, alpha=0.2, fill="black") +
  geom_rect(data=Pheno2_BreakYears.df, 
            mapping=aes(xmin=breakpoint2-2*breakpoint2.std.err, 
                        xmax=breakpoint2+2*breakpoint2.std.err, ymin=-Inf, ymax=Inf), 
            color=NA, alpha=0.2, fill="black") +
  
  # Add dashed vertical lines at breakpoints
  geom_vline(data=Pheno2_BreakYears.df, aes(xintercept=breakpoint1), color="black", lty="dashed") +
  geom_vline(data=Pheno2_BreakYears.df, aes(xintercept=breakpoint2), color="black", lty="dashed") +
  
  # Add ribbon for CI of smoothed mean trend
  geom_ribbon(data=Pheno2_summary.all.df,
              aes(x = MWyear, y = mean, ymin = lo, ymax = hi),
              fill = "darkgrey", color=NA, alpha = 0.9) + 
  
  # Add individual species time series
  geom_line(aes(x = MWyear, y = value, group=species, color=species), linewidth = 0.4) +
  
  # Linear trend over the aggregated series (dashed)
  geom_smooth(data=Pheno2_summary.all.df,
              aes(x = MWyear, y = mean),
              method='lm', formula = y~x, se = FALSE, 
              linewidth=1, linetype="dashed", color="black") +
  
  # Add overall mean trend line
  geom_line(data=Pheno2_summary.all.df,
            aes(x = MWyear, y = mean), color="black", linewidth=1) +
  
  # Manual color palette for species
  scale_color_manual("Species", 
                     values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                                '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a'))+
  
  # Zoom in on period of interest
  coord_cartesian(xlim=c(1967.9,2007.1))+
  
  # Axis labels
  xlab("") + ylab('Leaf-out anomaly (days)')+
  
  # Custom x-axis labels showing moving windows
  scale_x_continuous(breaks = seq(1966,2008,by=5),
                     labels = c('','','','','','','','','')) +
  
  # Add breakpoint coefficient labels (mean ± 2*SE) for each breakpoint
  geom_text(data = Pheno2_BP.coeff.all.df %>% filter(breakpoint == 1),
            aes(x = 1967, y = -Inf, 
                label = paste0(round(mean.coeff, 1), " ± ", round(2*se.coeff, 1))),
            hjust = 0, vjust = -.5, size=3.5) +
  geom_text(data = Pheno2_BP.coeff.all.df %>% filter(breakpoint == 2),
            aes(x = 1981, y = -Inf, 
                label = paste0(round(mean.coeff, 1), " ± ", round(2*se.coeff, 1))),
            hjust = 0, vjust = -.5, size=3.5) +
  geom_text(data = Pheno2_BP.coeff.all.df %>% filter(breakpoint == 3),
            aes(x = 1995, y = -Inf, 
                label = paste0(round(mean.coeff, 1), " ± ", round(2*se.coeff, 1))),
            hjust = 0, vjust = -.5, size=3.5) +
  
  # Slope annotation
  geom_text(
  data = ann_slope,
  aes(x = x, y = y, label = label),
  inherit.aes = FALSE,
  hjust = 1.05, vjust = 2, size = 3.5) +
  
  facet_grid(~phenology) +
  
  # Apply predefined theme
  plotTheme1 +
  theme(legend.position = "none")

plot.leafout.simulation
```

</details>

<details>

<summary>Temperature anomaly plot (Fig. 2a)</summary>

```{r}
#-------------------------------------------------------
# Spring temperature anomaly plot (Fig. 2a)
#-------------------------------------------------------

# Filter datasets once using phenology + temp_vars
Temp_BreakYears.df = BreakYears.df %>% 
  filter(phenology == "leaf_out",
         variable == "Tmax.spring60_mean")

Temp_summary.all.df = summary.all.df %>% 
  filter(phenology == "leaf_out",
         variable == "Tmax.spring60_mean")

Temp_BP.coeff.all.df = BP.coeff.all.df %>% 
  filter(phenology == "leaf_out",
         variable == "Tmax.spring60_mean")

# --- Linear trend (slope ± 2*SE)
ann_slope <- linear.trends.df %>% 
  filter(phenology == "leaf_out",
         variable == "Tmax.spring60_mean") %>% 
  mutate(x = -Inf, y = Inf)

# Plotting: filter main data and construct ggplot
#------------------------------------------------
plot.spring.temperature = summary.df %>% 
  
  filter(phenology %in% c("leaf_out"),
         variable %in% c("Tmax.spring60_mean")) %>% 
  
  ggplot() + 
  
  # Horizontal reference line at y=0
  geom_hline(yintercept = 0) +
  
  # Add shaded rectangles for breakpoint confidence intervals (breakpoint 1 & 2)
  geom_rect(data=Temp_BreakYears.df, 
            aes(xmin=breakpoint1-2*breakpoint1.std.err, 
                xmax=breakpoint1+2*breakpoint1.std.err, ymin=-Inf, ymax=Inf), 
            color=NA, alpha=0.2, fill="black")+
  geom_rect(data=Temp_BreakYears.df, 
            aes(xmin=breakpoint2-2*breakpoint2.std.err, 
                xmax=breakpoint2+2*breakpoint2.std.err, ymin=-Inf, ymax=Inf), 
            color=NA, alpha=0.2, fill="black")+
  
  # Add dashed vertical lines at breakpoints
  geom_vline(data=Temp_BreakYears.df, aes(xintercept=breakpoint1), color="black", lty="dashed")+
  geom_vline(data=Temp_BreakYears.df, aes(xintercept=breakpoint2), color="black", lty="dashed")+
  
  # Add ribbon for CI of smoothed mean trend
  geom_ribbon(data=Temp_summary.all.df,
              aes(x = MWyear, y = mean, ymin = lo, ymax = hi),
              fill = "darkgrey", color=NA, alpha = 0.9) + 
  
  # Add individual species time series
  geom_line(aes(x = MWyear, y = value, group=species, color=species), linewidth = 0.4) +
  
  # Add overall mean trend line
  geom_line(data=Temp_summary.all.df,
            aes(x = MWyear, y = mean),
            color="black", linewidth=1) +
  
  # Manual color palette for species
  scale_color_manual("Species", 
                     values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                                '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a'))+
  
  # Zoom in on period of interest
  coord_cartesian(xlim=c(1967.9,2007.1)) +
  
  # Axis labels
  xlab("") + ylab('Spring temperature anomaly (°C)') +
  
  # Custom x-axis labels showing moving windows
  scale_x_continuous(breaks = seq(1966,2008,by=5),
                     labels = c('','','','','','','','','')) +
  
  # Add breakpoint coefficient labels (mean ± 2*SE) for each breakpoint
  geom_text(data = Temp_BP.coeff.all.df %>% filter(breakpoint == 1),
            aes(x = 1967, y = -Inf, 
                label = paste0(round(mean.coeff, 2), " ± ", round(2*se.coeff, 2))),
            hjust = 0, vjust = -.5, size=3.5) +
  geom_text(data = Temp_BP.coeff.all.df %>% filter(breakpoint == 2),
            aes(x = 1981, y = -Inf, 
                label = paste0(round(mean.coeff, 2), " ± ", round(2*se.coeff, 2))),
            hjust = 0, vjust = -.5, size=3.5) +
  geom_text(data = Temp_BP.coeff.all.df %>% filter(breakpoint == 3),
            aes(x = 1993, y = -Inf, 
                label = paste0(round(mean.coeff, 2), " ± ", round(2*se.coeff, 2))),
            hjust = 0, vjust = -.5, size=3.5) +
  
   # Slope annotation
  geom_text(
  data = ann_slope,
  aes(x = x, y = y, label = label),
  inherit.aes = FALSE,
  hjust = -.1, vjust = 2, size = 3.5) +
  
  # Apply predefined theme
  plotTheme1 +
  theme(legend.position = "none") 

plot.spring.temperature
```

</details>

<br>

# 4. Leaf-out model plots

<details>

<summary>Model coefficients and breakpoints</summary>

```{r, warning=F, message=F}
#-------------------------------------------------------
# 1. Linear model coefficients of metric change over time
#-------------------------------------------------------

linear.trends.df <- model.summary.all.df %>%
  
  # 1) Define the grouping for which you want *one* trend each
  group_by(phenology, equation, metric) %>%
  
  # 2) Fit one linear model per group and store it as a list-column.
  summarise(
    model = list(lm(mean_all ~ MWyear, data = pick(MWyear, mean_all))),
    .groups = "drop"   # drop grouping after summarise
  ) %>%
  
  # 3) Extract model results into two standard broom tables:
  mutate(tidied  = purrr::map(model, broom::tidy)) %>%
  dplyr::select(-model) %>%   # We no longer need the raw model objects
  
  # 4) Expand the "tidied" list-column so you get one row per model term
  tidyr::unnest(tidied) %>%
  
  # 5) Keep only the slope term (drop the intercept row)
  filter(term == "MWyear") %>%
  
  # 6) Rename to your preferred names
  rename(slope = estimate, se = std.error, pval = p.value) %>%
  
  # 7) Build an annotation label (and provide x/y placement defaults for geom_text)
  mutate(
    
    # significance labeling
    p_lab = case_when(
      pval < 0.001 ~ "***",
      pval < 0.01  ~ "**",
      pval < 0.05  ~ "*",
      TRUE         ~ sprintf("; p = %.2f", pval)),
    
    # ---- Units logic ----
    unit = dplyr::case_when(
      metric == "estimate" ~ "St per year",
      metric == "cor_xy"   ~ "per year",
      
      # Any metric name containing "rel" -> percent per year
      stringr::str_detect(metric, "rel") ~ "% per year",
      
      # mean_x_anom special-cases
      metric == "mean_x_anom" & equation == "St_abs" ~ "\u00B0C per year",
      metric == "mean_x_anom" & equation != "St_abs" ~ "per year",
      
      # mean_y_anom special-cases
      metric == "mean_y_anom" & equation == "St_abs" ~ "days per year",
      metric == "mean_y_anom" & equation != "St_abs" ~ "per year",
      
      # Fallback (in case new metrics appear)
      TRUE ~ "per year"),
    
    # choose formatting
    fmt = case_when(
      metric == "estimate" ~ "%.4f \u00B1 %.4f %s %s",
      equation == "St_log" & metric == "mean_x_anom" ~ "%.3f \u00B1 %.3f %s %s",
      TRUE ~ "%.2f \u00B1 %.2f %s %s"),
    
    # label text shown on plots
    label = sprintf(fmt, slope, 2 * se, unit, p_lab)) %>%
  
  # 8) Keep only what you need for joining/filtering + plotting annotation
  dplyr::select(phenology, equation, metric, slope, se, pval, label)

linear.trends.df


#-------------------------------------------------------
# 2. Look at species-level trends
#-------------------------------------------------------

species.model = model.df %>% 
  filter(equation %in% c("St_abs", "St_log")) %>% 
  dplyr::group_by(phenology, equation, species) %>% 
  dplyr::group_modify(~{
     # run model
    model <- 
      lmer(estimate ~ MWyear + (1|timeseries), 
           data = .x, REML = FALSE)
    # fixed effects table
    tdy <- broom.mixed::tidy(model, effects = "fixed")  # term, estimate, std.error, statistic
  }) %>% 
  dplyr::filter(term == "MWyear") %>% 
  dplyr::select(phenology, equation, species, estimate, std.error)
species.model

# Ribes change in St over time
Ribes = species.model[species.model$species=="Ribes uva-crispa" &
                      species.model$equation=="St_abs" &
                      species.model$phenology %in% c("leaf_out", "leaf_out_GDD"), ]$estimate
# Other species change in St over time
mean.species = mean(species.model[species.model$species!="Ribes uva-crispa" &
                                  species.model$equation=="St_abs" &
                                  species.model$phenology=="leaf_out",]$estimate)
# Other species change in St simulation over time
mean.species.sim = mean(species.model[species.model$species!="Ribes uva-crispa" &
                                      species.model$equation=="St_abs" &
                                      species.model$phenology=="leaf_out_GDD",]$estimate)
# Biggest change in St over time
max.species = min(species.model[species.model$equation=="St_abs" &
                                species.model$phenology=="leaf_out",]$estimate)
# Biggest change in St over time
max.species.sim = min(species.model[species.model$equation=="St_abs" &
                                    species.model$phenology=="leaf_out_GDD",]$estimate)

paste0("Other species have a ", round(mean.species/Ribes[1], 1), " times greater change in St over time than Ribes")
paste0("Other species have a ", round(mean.species.sim/Ribes[2], 1), " times greater simulated change in St over time than Ribes")
paste0("Fastest changing species (Betula) has a ", round(max.species/Ribes[1], 1), " times greater change in St over time than Ribes")
paste0("Fastest changing species (Fagus) has a ", round(max.species.sim/Ribes[2], 1), " times greater simulated change in St over time than Ribes")


#-------------------------------------------------------
# 3. Breakpoints of regression coefficients
#-------------------------------------------------------

BP.model.results.df = model.summary.df %>% 
  
  # filter regression coefficients
  filter(metric == "estimate") %>% 
  
  # group data
  group_by(phenology, equation, metric) %>%  
  
  do({
    
    #run linear model
    linear.mod    = lm(mean ~ MWyear, data=.)
    
    #run breakpoint model
    segmented.mod = segmented(linear.mod, seg.Z=~MWyear, psi=list(MWyear = c(1970, 1988)),
                              seg.control(n.boot=100, fix.npsi=F))
    
    # get the fitted data
    my.fitted     = fitted(segmented.mod)
    
    # extract breakpoints in case model predicts them
    if(length(segmented.mod$coefficients)>2){
      
      # get the breakpoints
      breakpoint = data.frame(segmented.mod$psi) %>% 
        mutate(Est. = round(Est.)) 
      
      # get vector of breakpoint numbers
      bp.vec = c(rep(1, length(min(.$MWyear):breakpoint[1,]$Est.)),
                 rep(2, length((breakpoint[1,]$Est.+1):breakpoint[2,]$Est.)),
                 rep(3, length((breakpoint[2,]$Est.+1):max(.$MWyear))))
      
      # create empty data if no breakpoints were estimated
    } else {
      breakpoint = data.frame(Initital=c(NA,NA), Est.=c(NA,NA) )
      bp.vec = rep(0, length(min(.$MWyear):max(.$MWyear)))
    }
    
    # combine data
    data.frame(MWyear      = unique(.$MWyear), 
               breakpoint  = bp.vec,
               breakpoint1 = breakpoint[1,]$Est.,
               breakpoint2 = breakpoint[2,]$Est.,
               breakpoint1.std.err = breakpoint[1,]$St.Err,
               breakpoint2.std.err = breakpoint[2,]$St.Err)
  })

# Get mean breakpoint years and confidence range across species
BreakYears.df = BP.model.results.df %>% 
  distinct(phenology, equation, .keep_all = T) %>% 
  dplyr::select(phenology, equation, breakpoint1, breakpoint2, breakpoint1.std.err, breakpoint2.std.err)

# Summarize coefficients and slopes for each breakpoint
BP.coeff.all.df = model.summary.df %>% 
  filter(metric == "estimate") %>% 
  left_join(., 
            BP.model.results.df %>% 
              dplyr::select(phenology, MWyear, equation, breakpoint), 
            by=c("phenology", "MWyear", "equation")) %>% 
  #summarize for each species
  group_by(phenology, equation, breakpoint, species) %>% 
  summarize_at("mean", tibble::lst(coeff = ~ mean(., na.rm=T)) ) %>% 
  #summarize across species
  group_by(phenology, equation, breakpoint) %>% 
  summarize_at("coeff", tibble::lst(mean.coeff = ~ mean(., na.rm=T), se.coeff = stderror) )
```

</details>

<details>

<summary>Temperature sensitivity plots (Fig. 2c, d)</summary>

```{r, warning=F, message=F}
#-------------------------------------------------------
# Temperature sensitivity plot (St = days per °C; Fig. 2c)
#-------------------------------------------------------

# --- Filter breakpoint data (change-point analysis) ---
St_BreakYears.df = BreakYears.df %>% 
  filter(phenology == "leaf_out", equation == "St_abs", metric == "estimate")

# --- Filter and transform model summary data ---
St_model.summary.all.df = model.summary.all.df %>% 
  filter(phenology == "leaf_out", equation == "St_abs", metric == "estimate")

# --- Filter breakpoint coefficient data for annotations ---
St_BP.coeff.all.df = BP.coeff.all.df %>% 
  filter(phenology == "leaf_out", equation == "St_abs")

# --- Linear trend (slope ± 2*SE)
ann_slope <- linear.trends.df %>% 
  filter(phenology == "leaf_out", equation == "St_abs", metric == "estimate") %>% 
  mutate(x = Inf, y = Inf)

# Plotting: filter main data and construct ggplot
#------------------------------------------------
plot.St = model.summary.df %>% 
  filter(phenology == "leaf_out", equation == "St_abs", metric == "estimate") %>% 
  
  ggplot() + 
  
  # (1) Shaded rectangles: 95% CI around breakpoints 1 and 2
  geom_rect(data=St_BreakYears.df, 
            mapping=aes(xmin=breakpoint1-2*breakpoint1.std.err, 
                        xmax=breakpoint1+2*breakpoint1.std.err, 
                        ymin=-Inf, ymax=Inf), 
            color=NA, alpha=0.2, fill="black")+
  geom_rect(data=St_BreakYears.df, 
            mapping=aes(xmin=breakpoint2-2*breakpoint2.std.err, 
                        xmax=breakpoint2+2*breakpoint2.std.err, 
                        ymin=-Inf, ymax=Inf), 
            color=NA, alpha=0.2, fill="black")+
  
  # (2) Dashed vertical lines at breakpoints
  geom_vline(data=St_BreakYears.df, mapping=aes(xintercept=breakpoint1), color="black", lty="dashed")+
  geom_vline(data=St_BreakYears.df, mapping=aes(xintercept=breakpoint2), color="black", lty="dashed")+
  
  # (3) CI ribbon for the aggregated series
  geom_ribbon(data=St_model.summary.all.df,
              aes(x = MWyear, y = mean_all, ymin = lo, ymax = hi),
              fill = "darkgrey", color=NA, alpha = 0.9) + 
  
  # (4) Species lines (inherits data from the main pipe)
  geom_line(aes(x = MWyear, y = mean, group=species, color=species), linewidth = 0.4) +
  
  # (5) Linear trend over the aggregated series (dashed)
  geom_smooth(data=St_model.summary.all.df,
              aes(x = MWyear, y = mean_all),
              method='lm', formula = y~x, se = FALSE, 
              linewidth=1, linetype="dashed", color="black") +
  
  # (6) Aggregated series line
  geom_line(data=St_model.summary.all.df, aes(x = MWyear, y = mean_all),
            color="black", size=1) +
  
  # Manual species color palette
  scale_color_manual("Species", 
                     values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                                '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a'))+
  
  # Axis limits: restrict x-range to window years; y-range customized
  coord_cartesian(xlim=c(1967.9,2007.1), ylim=c(0.22, 4.8))+
  
  xlab("") + ylab('St (days per °C)')+
  
  # X-axis breaks for moving-window labels (left blank intentionally)
  scale_x_continuous(breaks = seq(1966,2008,by=5),
                     labels = c('','','','','','','','','')) +
  
  # Breakpoint coefficient annotations (rounded + sign flipped)
  geom_text(data = St_BP.coeff.all.df %>% filter(breakpoint == 1),
            mapping = aes(x = 1967, y = -Inf, 
                          label = paste0(round(mean.coeff, 1), " ± ", round(2*se.coeff, 1))),
            hjust   = 0, vjust   = -.5, size=3.5) +
  geom_text(data = St_BP.coeff.all.df %>% filter(breakpoint == 2),
            mapping = aes(x = 1981, y = -Inf, 
                          label = paste0(round(mean.coeff, 1), " ± ", round(2*se.coeff, 1))),
            hjust = 0, vjust = -.5, size=3.5) +
  geom_text(data = St_BP.coeff.all.df %>% filter(breakpoint == 3),
            mapping = aes(x = 1995, y = -Inf, 
                          label = paste0(round(mean.coeff, 1), " ± ", round(2*se.coeff, 1))),
            hjust = 0, vjust = -.5, size=3.5) +
  
  # Slope annotation
  geom_text(
  data = ann_slope,
  aes(x = x, y = y, label = label),
  inherit.aes = FALSE,
  hjust = 1.1, vjust = 2, size = 3.5) +
  
   # Plot theme and legend
  plotTheme1 +
  theme(legend.position = "none") 

plot.St


#-------------------------------------------------------
# Pearson correlation coefficient (r) plot (Fig. 2d)
#-------------------------------------------------------

# --- Filter and transform model summary data ---
Cor_model.summary.all.df = model.summary.all.df %>% 
  filter(phenology == "leaf_out", equation == "St_abs", metric == "cor_xy")

# --- Linear trend (slope ± 2*SE)
ann_slope <- linear.trends.df %>% 
  filter(phenology == "leaf_out", equation == "St_abs", metric == "cor_xy") %>% 
  mutate(x = 1967, y = Inf)

# Plotting: filter main data and construct ggplot
#------------------------------------------------
plot.Cor = model.summary.df %>% 
  filter(phenology == "leaf_out", equation == "St_abs", metric == "cor_xy") %>% 
  
  ggplot() + 
  
  # (1) CI ribbon for the aggregated series
  geom_ribbon(data = Cor_model.summary.all.df,
              aes(x = MWyear, y = mean_all, ymin = lo, ymax = hi),
              fill = "darkgrey", color=NA, alpha = 0.9) + 
  
  # (2) Species lines (inherits data from the main pipe)
  geom_line(aes(x = MWyear, y = mean, group=species, color=species), linewidth = 0.4) +
  
  # (3) Linear trend over the aggregated series (dashed)
  geom_smooth(data = Cor_model.summary.all.df,
              aes(x = MWyear, y = mean_all),
              method='lm', formula = y~x, se = FALSE, 
              linewidth=1, linetype="dashed", color="black") +
  
  # (4) Aggregated series line
  geom_line(data = Cor_model.summary.all.df, 
            aes(x = MWyear, y = mean_all),
            color="black", linewidth=1) +
  
  # Manual species color palette
  scale_color_manual("Species", 
                     values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                                '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a'))+
  
  # Axis limits: restrict x-range to window years; y-range customized
  coord_cartesian(xlim=c(1967.9,2007.1), ylim=c(-.9,-.04))+
  
  xlab("") + ylab('Pearson`s r')+
  
  # X-axis breaks for moving-window labels (left blank intentionally)
  scale_x_continuous(breaks = seq(1966,2008,by=5),
                     labels = c('1966-1980','1971-1985','1976-1990',
                                '1981-1995','1986-2000','1991-2005',
                                '1996-2010','2001-2015','2006-2020')) +
  
  # Slope annotation
  geom_text(
  data = ann_slope,
  aes(x = x, y = y, label = label),
  inherit.aes = FALSE,
  hjust = -.1, vjust = 2.5, size = 3.5) +
  
  # Plot theme and legend
  plotTheme1 +
  theme(legend.position = "none") 

plot.Cor
```

</details>

<details>

<summary>Temperature sensitivity simulation plots (Figs. 3b and 3c)</summary>

```{r, warning=F, message=F}
#------------------------------------------------------------
# Temperature sensitivity simulation plots (St = days per °C; Fig. 3b)
#------------------------------------------------------------

# Define desired phenology order
phenology_levels <- c("leaf_out_GDD", 
                      "leaf_out_GDD.chill.lo", 
                      "leaf_out_GDD.chill.hi", 
                      "leaf_out_GDD.chill.DL")

# --- Filter and transform model summary data ---
St_simulation_model.summary.all.df = model.summary.all.df %>% 
  filter(phenology != "leaf_out", equation == "St_abs", metric == "estimate") %>% 
  mutate(phenology = factor(phenology, levels = phenology_levels))
  
# --- Linear trend (slope ± 2*SE)
ann_slope <- linear.trends.df %>% 
  filter(phenology != "leaf_out", equation == "St_abs", metric == "estimate") %>% 
  mutate(x = -Inf, y = -Inf,
         phenology = factor(phenology, levels = phenology_levels))

# Plotting: filter main data and construct ggplot
#------------------------------------------------
plot.St.simulation = model.summary.df %>% 
  filter(phenology != "leaf_out", equation == "St_abs", metric == "estimate") %>% 
  dplyr::mutate(phenology = factor(phenology, levels = phenology_levels)) %>% # sort factors
  
  ggplot() + 
  
  # Confidence ribbon for overall trend
  geom_ribbon(data=St_simulation_model.summary.all.df,
              aes(x = MWyear, y = mean_all, ymin = lo, ymax = hi),
              fill = "darkgrey", color=NA, alpha = 0.9) + 
  
  # Individual species lines (filtered from pipe input)
  geom_line(aes(x = MWyear, y = mean, group=species, color=species), linewidth = 0.4) +
  
  # Linear trend over the aggregated series (dashed)
  geom_smooth(data=St_simulation_model.summary.all.df,
              aes(x = MWyear, y = mean_all),
              method='lm', formula = y~x, se = FALSE, 
              linewidth=1, linetype="dashed", color="black") +
  
  # Smoothed line: overall trend (black line)
  geom_line(data=St_simulation_model.summary.all.df, aes(x = MWyear, y = mean_all),
            color="black", size=1) +
  
  # Manual species color palette
  scale_color_manual("Species", 
                     values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                                '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a'))+
  
  # Axis limits: restrict x-range to window years; y-range customized
  coord_cartesian(xlim=c(1967.9,2007.1), ylim=c(0.38, 8))+
  
  xlab("") + ylab('St (days per °C)')+
  
  # X-axis breaks for moving-window labels (left blank intentionally)
  scale_x_continuous(breaks = seq(1966,2008,by=5),
                     labels = c('','','','','','','','','')) +
  
  # Slope annotation
  geom_text(
  data = ann_slope,
  aes(x = x, y = y, label = label),
  inherit.aes = FALSE,
  hjust = -0.05, vjust = -1, size = 3.5) +
  
  facet_grid(~phenology) +
  
  # Plot theme and legend
  plotTheme1

plot.St.simulation


#------------------------------------------------------------
# Pearson correlation coefficient (r) plots (Fig. 3c)
#------------------------------------------------------------

# --- Filter and transform model summary data ---
Cor_simulation_model.summary.all.df = model.summary.all.df %>% 
  filter(phenology != "leaf_out", equation == "St_abs", metric == "cor_xy") %>% 
  mutate(phenology = factor(phenology, levels = phenology_levels))

# --- Linear trend (slope ± 2*SE)
ann_slope <- linear.trends.df %>% 
  filter(phenology != "leaf_out", equation == "St_abs", metric == "cor_xy") %>% 
  mutate(x = -Inf, y = Inf,
         phenology = factor(phenology, levels = phenology_levels))

# Plotting: filter main data and construct ggplot
#------------------------------------------------
plot.cor.simulation = model.summary.df %>% 
  filter(phenology != "leaf_out", equation == "St_abs", metric == "cor_xy") %>% 
  dplyr::mutate(phenology = factor(phenology, levels = phenology_levels)) %>% # sort factors
  
  ggplot() + 
  
  # Confidence ribbon for overall trend
  geom_ribbon(data=Cor_simulation_model.summary.all.df,
              aes(x = MWyear, y = mean_all, ymin = lo, ymax = hi),
              fill = "darkgrey", color=NA, alpha = 0.9) + 
  
  # Individual species lines (filtered from pipe input)
  geom_line(aes(x = MWyear, y = mean, group=species, color=species), linewidth = 0.4) +
  
  # Linear trend over the aggregated series (dashed)
  geom_smooth(data=Cor_simulation_model.summary.all.df,
              aes(x = MWyear, y = mean_all),
              method='lm', formula = y~x, se = FALSE, 
              linewidth=1, linetype="dashed", color="black") +
  
  # Smoothed line: overall trend (black line)
  geom_line(data=Cor_simulation_model.summary.all.df, aes(x = MWyear, y = mean_all),
            color="black", size=1) +
  
  # Manual species color palette
  scale_color_manual("Species", 
                     values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                                '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a'))+
  
  # Axis limits: restrict x-range to window years; y-range customized
  coord_cartesian(xlim=c(1967.9,2007.1), ylim=c(-1,-.05))+
  
  xlab("") + ylab('Pearsons`s r')+
  
  # X-axis breaks for moving-window labels (left blank intentionally)
  scale_x_continuous(breaks = seq(1966,2008,by=5),
                     labels = c('1966-1980','1971-1985','1976-1990',
                                '1981-1995','1986-2000','1991-2005',
                                '1996-2010','2001-2015','2006-2020')) +
  
  # Slope annotation
  geom_text(
  data = ann_slope,
  aes(x = x, y = y, label = label),
  inherit.aes = FALSE,
  hjust = -0.1, vjust = 2.5, size = 3.5) +
  
  facet_grid(~phenology) +
  
  # Plot theme and legend
  plotTheme1 +
  theme(legend.position = "none") 

plot.cor.simulation
```

</details>

<details>

<summary>Relative Ts change (Fig. 4)</summary>

```{r, warning=F, message=F}
#===============================================================================
# Relative decline in temperature sensitivity (St) from early → late moving windows
#   - Computes mean St per species × equation × phenology for:
#       (i) earliest 15 moving windows  = "early"
#      (ii) latest   15 moving windows  = "late"
#   - Returns relative decline (%) = 100 * (St_early - St_late) / St_early
#   - Plots distributions across species as violin + boxplot + jittered points
#===============================================================================


#------------------------------------------------------
# 1) Define time windows
#------------------------------------------------------

mw_min <- min(model.summary.df$MWyear, na.rm = TRUE)
mw_max <- max(model.summary.df$MWyear, na.rm = TRUE)

early <- mw_min:(mw_min + 14)   # earliest 15 moving windows
late  <- (mw_max - 14):mw_max   # latest 15 moving windows


#------------------------------------------------------
# 2) Phenology model ordering + plot labels + colors
#------------------------------------------------------

phen_order <- c(
  "leaf_out",
  "leaf_out_GDD",
  "leaf_out_GDD.chill.lo",
  "leaf_out_GDD.chill.hi",
  "leaf_out_GDD.chill.DL"
)

phen_labels <- c(
  leaf_out              = "Observations",
  leaf_out_GDD          = "Forcing-only",
  leaf_out_GDD.chill.lo = "Low chilling",
  leaf_out_GDD.chill.hi = "High chilling",
  leaf_out_GDD.chill.DL = "High chilling + Daylength"
)


#------------------------------------------------------
# 3) Compute relative decline in St (species-level)
#------------------------------------------------------

relative.Ts.df <- model.summary.df %>%
  # keep only early/late windows, target equations, and target model variants
  filter(
    MWyear %in% c(early, late),
    metric == "estimate",
    phenology %in% phen_order
  ) %>%
  # classify MWyear into early vs late
  mutate(period = ifelse(MWyear %in% early, "early", "late")) %>%
  # average across the 15 moving windows within each period
  group_by(species, equation, phenology, period) %>%
  summarise(St = mean(mean, na.rm = TRUE), .groups = "drop") %>%
  # reshape to have columns early and late
  pivot_wider(names_from = period, values_from = St) %>%
  # drop incomplete pairs (and avoid division by zero)
  filter(!is.na(early), !is.na(late), early != 0) %>%
  mutate(
  change_pct = 100 * (late / early -1),  # 0 = no change; + = increase; - = decrease
  # enforce consistent ordering in plot
  phenology = factor(phenology, levels = phen_order),
  equation  = factor(equation, levels = c("St_abs", "St_log", "St_z"))
  ) %>%
  arrange(species, equation, phenology)

# panel letters (placed manually in the upper-left of each facet)
panel_labs <- tibble(
  equation = factor(c("St_abs", "St_log", "St_z"), 
                    levels = levels(relative.Ts.df$equation)),
  lab      = c("a", "b", "c")
)


#------------------------------------------------------
# 4) Plot: box + species points, faceted by equation
#------------------------------------------------------

plot.relative.St <- ggplot(relative.Ts.df, aes(x = phenology, y = change_pct)) +

  # Reference: 100% = no change
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +

  # Make all distributions grey (no phenology fill)
  #geom_violin(trim = TRUE, alpha = 0.85, fill = "grey85", color = NA) +
  geom_boxplot(width = 0.4, outlier.shape = NA, 
               fill = "grey90", color = "grey30") +

  # Color points by species (manual palette)
  geom_point(
  aes(color = species),
  position = position_jitter(width = 0.18, height = 0),
  size = 1.5, alpha = 0.85) +

  facet_wrap(~ equation, nrow = 1, scales = "fixed") +

  # Panel letters a / b (upper-left)
  geom_text(
    data = panel_labs,
    aes(x = -Inf, y = Inf, label = lab),
    inherit.aes = FALSE,
    hjust = -0.2, vjust = 1.2, fontface = "bold", size = 5
  ) +

  # Rename phenology classes on x-axis
  scale_x_discrete(labels = phen_labels) +

  # Manual species color palette (10 species)
  scale_color_manual(
    "Species",
    values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
               '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a')
  ) +

  labs(x = NULL, y = "Change in St relative to early period (%)") +
  coord_cartesian(ylim = c(-50, 30)) +
  plotTheme1 +
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 25, hjust = 1),
    panel.spacing.x = unit(0.8, "lines"),
    legend.position = "right"
  )
plot.relative.St
```

</details>

<details>

<summary>Relative St and variance ratio plots (Figs. Sx, Sy, and Sx)</summary>

```{r, warning=F, message=F}
#------------------------------------------------------------
# Relative St and variance ratio plots
#------------------------------------------------------------

# equations
equations = c("St_abs",
              "St_log",
              "St_z")

# Define order (raw values in your data)
phenology_levels <- c(
  "leaf_out",
  "leaf_out_GDD",
  "leaf_out_GDD.chill.lo",
  "leaf_out_GDD.chill.hi",
  "leaf_out_GDD.chill.DL"
)

# Define display labels (same order as above)
phenology_labels <- c(
  "Observations",
  "Forcing-only model",
  "Low chilling model",
  "High chilling model",
  "High chilling + daylength model"
)

# Helper: enforce order + labels everywhere
label_phenology <- function(df) {
  df %>%
    mutate(
      phenology = factor(
        phenology,
        levels = phenology_levels,
        labels = phenology_labels
      )
    )
}


#------------------------------------------------------------
# Loop over equations
#------------------------------------------------------------

for (equation.name in equations) {
  
  # Create data
  #------------
  # choose metrics
  if (equation.name == "St_z") {
    metrics = c("estimate","sd_x_rel","sd_y_rel","var_ratio_rel","mean_x_anom") 
  } else {
    metrics = c("estimate_rel","sd_x_rel","sd_y_rel","var_ratio_rel","mean_x_anom") 
    }

  # Model summay data
  St.summary.all.df = model.summary.all.df %>% 
    filter(
      equation == equation.name,
      metric %in% metrics) %>% 
    label_phenology()
  
  # Linear trend (slope ± 2*SE)
  ann_slope = linear.trends.df %>% 
    filter(
      equation == equation.name,
      metric %in% metrics) %>% 
    mutate(x = -Inf, y = -Inf) %>% 
    label_phenology()
 
  # Plot data
  plot_df <- model.summary.df %>%
    filter(
      equation == equation.name,
      metric %in% metrics) %>%
    label_phenology()
  
  
  # Create plot
  #------------
  plot = ggplot() + 
  
  geom_hline(yintercept = 0) +
  
  # Confidence ribbon for overall trend
  geom_ribbon(data=St.summary.all.df,
              aes(x = MWyear, y = mean_all, ymin = lo, ymax = hi),
              fill = "darkgrey", color=NA, alpha = 0.9) + 
  
  # Individual species lines (filtered from pipe input)
  geom_line(data = plot_df,
            aes(x = MWyear, y = mean, group=species, color=species), 
            linewidth = 0.4) +
  
  # Linear trend over the aggregated series (dashed)
  geom_smooth(data = St.summary.all.df,
              aes(x = MWyear, y = mean_all),
              method='lm', formula = y~x, se = FALSE, 
              linewidth=1, linetype="dashed", color="black") +
  
  # Smoothed line: overall trend (black line)
  geom_line(data=St.summary.all.df, 
            aes(x = MWyear, y = mean_all),
            color="black", size=1) +
  
  # Manual species color palette
  scale_color_manual("Species", 
                     values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                                '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a'))+
  
  # Axis limits: restrict x-range to window years; y-range customized
  coord_cartesian(xlim=c(1967.9,2007.1))+
  
  xlab("") + ylab(expression("Relative change"))+
  
  # X-axis breaks for moving-window labels (left blank intentionally)
  scale_x_continuous(breaks = seq(1966,2008,by=5),
                     labels = c('1966-1980','1971-1985','1976-1990',
                                '1981-1995','1986-2000','1991-2005',
                                '1996-2010','2001-2015','2006-2020')) +
  
  # Slope annotation
  geom_text(
  data = ann_slope,
  aes(x = x, y = y, label = label),
  inherit.aes = FALSE,
  hjust = -0.05, vjust = -1, size = 3.5) +
  
  facet_grid(metric~phenology, scales = "free_y") +
  
  # Plot theme and legend
  plotTheme1

  print(plot)
  
  # Fig. Sx-y: Relative St + Variance ratio plots
  ggsave(plot, 
       file=paste0("FigSX_VarianceRatios_", equation.name, ".pdf"), 
       path=output.dir,
       width=14, height=10)
}
```

</details>

<details>

<summary>Standardized driver plots (Fig. S12)</summary>

```{r, warning=F, message=F}
#-------------------------------------------------------
# Standardized driver plots (Fig. S12)
#-------------------------------------------------------

# Filter datasets once using phenology + equation + term
Full_model.summary.df = mixed.model.summary.df %>% 
  filter(phenology == "leaf_out")

# Plotting: filter main data and construct ggplot
#------------------------------------------------
plot.standardized.supp = mixed.model.df %>% 
  
   # Keep only relevant variables              
  filter(phenology == "leaf_out") %>% 
  
  # sort term factors
  mutate(term = factor(term, levels = c("T[max]", "Radiation", "T[Q4]", "Leaf-out"))) %>% 
  
  ggplot() + 
  
  # (1) CI ribbon for the aggregated series
  geom_ribbon(data=Full_model.summary.df,
              aes(x = MWyear, y = coeff.mean, ymin = coeff.lo, ymax = coeff.hi),
              fill = "darkgrey", color=NA, alpha = 0.9) + 
  
   # (2) Horizontal reference line
  geom_hline(yintercept=0)+
  
  # (3) Species lines (inherits data from the main pipe)
  geom_line(aes(x = MWyear, y = estimate, group=species, color=species), linewidth = 0.4) +
  
  # (4) Linear trend over the aggregated series (dashed)
  geom_smooth(data=Full_model.summary.df,
              aes(x = MWyear, y = coeff.mean),
              method='lm', formula = y~x, se = FALSE, 
              linewidth=1, linetype="dashed", color="black") +
  
  # (5) Aggregated series line
  geom_line(data=Full_model.summary.df,
            aes(x = MWyear, y = coeff.mean), 
            color="black", linewidth=1) +
  
  # Manual palette
  scale_color_manual("Species", 
                     values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                                '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a'))+
  
  # X and Y window for moving-window years
  coord_cartesian(xlim=c(1967.9,2006.1), ylim=c(-.87,.2))+
  xlab("") + ylab('Standardized effect (z-score)')+
  
  # Axis tick labels for windows
  scale_x_continuous(breaks = seq(1966,2008,by=5),
                     labels = c('1966-1980','1971-1985','1976-1990',
                                '1981-1995','1986-2000','1991-2005',
                                '1996-2010','2001-2015','2006-2020')) +
  
  
# facet by term
  facet_wrap(~term, ncol=4, dir="h", labeller=label_parsed)+
  
   # Correlation annotation on the per-species data
  ggpubr::stat_cor(aes(x = MWyear, 
                       y = estimate), 
                   color = "black", 
                   r.accuracy = 0.01, p.accuracy = 0.001, 
                   show.legend=F,
                   label.x = 1980, label.y = -0.35) +
  
   # Apply predefined theme
  plotTheme1

plot.standardized.supp
```

</details>

<br>

# 5. Save plots

<details>

<summary>Show</summary>

```{r}
#-------------------------------------------------------
# Define plot layouts
#-------------------------------------------------------

layout4 <- "
A
B
C
D"

layout3 <- "
A
B
C"


#-------------------------------------------------------
# Merge plots
#-------------------------------------------------------

MainPlot = plot.spring.temperature + plot.leafout + plot.St + plot.Cor +
  plot_layout(design = layout4) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

MainPlot

SimulationPlot = plot.leafout.simulation + plot.St.simulation + plot.cor.simulation + 
  plot_layout(design = layout3) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

SimulationPlot


#-------------------------------------------------------
# Save plots as .pdf
#-------------------------------------------------------

# Fig. 2: Main plot (spring temperature anomalies + leaf-out anomalies + St + Z-scores)
ggsave(MainPlot, 
       file="Fig2_St_main.pdf", 
       path=output.dir,
       width=6, height=12)

# Fig. 3: Simulations plots (leaf-out anomalies + St)
ggsave(SimulationPlot, 
       file="Fig3_St_simulation.pdf", 
       path=output.dir,
       width=12, height=9)

# Fig. 4: Relative St change
ggsave(plot.relative.St, 
       file="Fig4_Relative_St_decline.pdf", 
       path=output.dir,
       width=9, height=4)

# Fig. Sx: Multivariate models
ggsave(plot.standardized.supp, 
       file="FigS12_multiple_standardized.pdf", 
       path=output.dir,
       width=11, height=3.5)
```

</details>

<br>

# 6. Reproducibility

<details>

<summary>Show</summary>

```{r}
# Datetime
Sys.time()

# Session info
sessionInfo()
```

</details>
