---
title: "Spring phenology tier2 project"
subtitle: "Model performance figures (PEP725 data)"
author: "Constantin Zohner"
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

format:
  html:
    toc: false
    theme: default
    highlight-style: haddock
    df-print: paged
    embed-resources: true

execute:
  #warning: false
  message: false
  echo: true
  output: true

always_allow_html: true
editor: visual
editor_options:
  chunk_output_type: inline
---

<br>

### Models

-   Forcing-only: Constant site-level GDD requirement
-   Low chilling: GDD requirement modified by chilling factor (C\* = 75)
-   High chilling: GDD requirement modified by chilling factor (C\* = 150)
-   Chilling & Photoperiod: GDD requirement modified by chilling (fC) and photoperiod factors (fP)

<br>

# 1. Setup

<details>

<summary>Packages</summary>

```{r, message = F}
require(tidyverse)
require(data.table)
require(ggplot2)
```

</details>

<details>

<summary>Directories</summary>

```{r}
#-------------------------------------------------
# Define the base directory
#-------------------------------------------------

base.dir <- "/Users/goalie/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis"


#-------------------------------------------------
# Build specific directories from the base
#-------------------------------------------------

# PEP LOO-CV data
Model.dir = file.path(base.dir, "Analysis_input/PEP_models")

# Moving window model results
PEP.dir = file.path(base.dir, "Analysis_input/PEP_drivers_final_spring/Merged_file")

# Output
output.dir = file.path(base.dir, "Analysis_output/ModelFigures/Moving_windows/Model_evaluation")
```

</details>

<details>

<summary>Plot theme</summary>

```{r}
plotTheme1 = theme(legend.position   = "right",
                   legend.background = element_rect(fill=NA, linewidth=0.5, linetype="solid"),
                   legend.text       = element_text(color="black",face = "italic"),
                   panel.grid.major  = element_blank(),
                   panel.grid.minor  = element_blank(),
                   panel.background  = element_blank(),
                   panel.border      = element_rect(colour = "black", fill=NA),
                   axis.line         = element_line(color = "black"),
                   axis.text         = element_text(colour = "black"),
                   axis.text.x       = element_text(angle = 45, hjust=1),
                   strip.background  = element_rect(fill=NA),
                   plot.title        = element_text(face="bold"))
```

</details>

<details>

<summary>Functions</summary>

```{r}
#-------------------------------------------------------
# RMSE — Root Mean Squared Error
#-------------------------------------------------------

rmse <- function(obs, pred, w = NULL) {
  # If weights are supplied, compute a weighted RMSE.
  # weighted.mean() divides by sum(w) (after dropping NA pairs),
  # so this yields sqrt( Σ w_i * (obs - pred)^2 / Σ w_i ).
  if (!is.null(w)) return(sqrt(weighted.mean((obs - pred)^2, w, na.rm = TRUE)))
  
  # Unweighted RMSE: sqrt( mean squared error ), dropping NA pairs.
  sqrt(mean((obs - pred)^2, na.rm = TRUE))
}


#-------------------------------------------------------
# cross-validated R2 (a.k.a. Nash–Sutcliffe efficiency)
#-------------------------------------------------------

# CV-friendly, penalizes bias & scale
# R2 = 1 - SSE/SST, where SSE = Σ (obs - pred)^2,
# and SST = Σ (obs - mean(obs))^2
r2_ss <- function(obs, pred) {
  # Require at least 2 non-NA pairs to define a variance
  if (sum(!is.na(obs) & !is.na(pred)) < 2) return(NA_real_)
  
  # 1 - (sum of squared errors) / (total sum of squares)
  1 - sum((obs - pred)^2, na.rm = TRUE) /
        sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE)
}


#-------------------------------------------------------
# lower (lo) and upper (hi) confidence intervals of mean
#-------------------------------------------------------

lo <- function(x, conf.level = 0.95) {
  x <- x[is.finite(x)]
  n <- length(x)
  if (n < 2) return(NA_real_)
  m <- mean(x); se <- stats::sd(x) / sqrt(n)
  tcrit <- stats::qt((1 + conf.level)/2, df = n - 1)
  m - tcrit * se
}


hi <- function(x, conf.level = 0.95) {
  x <- x[is.finite(x)]
  n <- length(x)
  if (n < 2) return(NA_real_)
  m <- mean(x); se <- stats::sd(x) / sqrt(n)
  tcrit <- stats::qt((1 + conf.level)/2, df = n - 1)
  m + tcrit * se
}
```

</details>

<details>

<summary>Load and prepare data</summary>

```{r}
#-------------------------------------------------------
# PEP725 data
#-------------------------------------------------------

PEP.df <- fread(paste(PEP.dir, "pep_drivers_spring_data_clean.csv", sep="/")) %>% 
  dplyr::select(timeseries, year, alt_dem)


#-------------------------------------------------------
# Model predictions (claibration and LOO-CV data)
#-------------------------------------------------------

LOO_CV.df      <- fread(paste(Model.dir, "pep_model_predictions_LOO.csv", sep="/")) 
Calibration.df <- fread(paste(Model.dir, "pep_model_predictions.csv", sep="/")) 


#-------------------------------------------------------
# Merge and filter data
#-------------------------------------------------------

pred_cols <- c("leaf_out_GDD",
               "leaf_out_GDD.chill.lo",
               "leaf_out_GDD.chill.hi",
               "leaf_out_GDD.chill.DL")

LOO_CV.df <- left_join(LOO_CV.df, PEP.df, by = c("timeseries", "year")) %>%
  group_by(timeseries) %>%
  filter(!(alt_dem > 600), year >= 1966) %>%
  filter(n() >= 20) %>%
  ungroup() %>%
  dplyr::select(-c(alt_dem, V1)) %>%
  # keep obs as 'obs', pivot preds to 'pred' and model name to 'model'
  rename(obs = leaf_out) %>%
  pivot_longer(cols = all_of(pred_cols),
               names_to = "model",
               values_to = "pred")

Calibration.df <- left_join(Calibration.df, PEP.df, by = c("timeseries", "year")) %>%
  group_by(timeseries) %>%
  filter(!(alt_dem > 600), year >= 1966) %>%
  filter(n() >= 20) %>%
  ungroup() %>%
  dplyr::select(-c(alt_dem, V1)) %>%
  # keep obs as 'obs', pivot preds to 'pred' and model name to 'model'
  rename(obs = leaf_out) %>%
  pivot_longer(cols = all_of(pred_cols),
               names_to = "model",
               values_to = "pred")


#-------------------------------------------------------
# Moving windows
#-------------------------------------------------------

year.vector <- seq(min(PEP.df$year), max(PEP.df$year) - 14L)

LOOCV.MW.df <- lapply(year.vector, function(MW.year) {
  LOO_CV.df %>%
    filter(year >= MW.year, year < MW.year + 15L) %>%
    group_by(timeseries, model) %>%
    filter(n() >= 12) %>%    # keep reasonably long series within the window
    ungroup() %>%
    mutate(MWyear = MW.year)
}) %>% bind_rows()

Cal.MW.df <- lapply(year.vector, function(MW.year) {
  Calibration.df %>%
    filter(year >= MW.year, year < MW.year + 15L) %>%
    group_by(timeseries, model) %>%
    filter(n() >= 12) %>%    # keep reasonably long series within the window
    ungroup() %>%
    mutate(MWyear = MW.year)
}) %>% bind_rows()
```

</details>

<br>

# 2. Analysis

<details>

<summary>Summarize calibration models</summary>

```{r, message=F}
#=======================================================
# Summarize model results
#=======================================================



#-------------------------------------------------------
# Compute per-timeseries × model × window metrics
#-------------------------------------------------------

Cal_metrics_by_ts <- Cal.MW.df %>%
  group_by(species, timeseries, model, MWyear) %>%
  summarise(
    #n     = sum(!is.na(obs) & !is.na(pred)),
    RMSE  = rmse(obs, pred),
    MAE   = mean(abs(obs - pred), na.rm = TRUE),
    #Bias  = mean(pred - obs, na.rm = TRUE),              # + means over-prediction
    r     = suppressWarnings(cor(obs, pred, use = "complete.obs")),
    #R2    = r2_ss(obs, pred),                            # 1 - SSE/SST
    #slope = if (n >= 2) coef(lm(obs ~ pred))[2] else NA_real_,
    #intercept = if (n >= 2) coef(lm(obs ~ pred))[1] else NA_real_,
    #AIC_lm = if (n >= 3) AIC(lm(obs ~ pred)) else NA_real_
  , .groups = "drop")


#-------------------------------------------------------
# Summarise time series-level results at species level
#-------------------------------------------------------

# Define used metrics
vars <- c("RMSE","MAE","r")

# Define desired model order
phenology_levels <- c("leaf_out_GDD",
                      "leaf_out_GDD.chill.lo",  
                      "leaf_out_GDD.chill.hi",
                      "leaf_out_GDD.chill.DL")

Cal_metrics_by_species <- Cal_metrics_by_ts %>%
  group_by(species, model, MWyear) %>%
  summarise(
    across(all_of(vars),
           list(. = ~mean(.x, na.rm = TRUE)),
           .names = "{.col}")
  , .groups = "drop") %>% 
  pivot_longer(cols = all_of(vars),
               names_to = "metric",
               values_to = "val") %>% 
  mutate(model  = factor(model, levels = phenology_levels),
         metric = factor(metric, levels = vars))


#-------------------------------------------------------
# Summarise species-level results
#-------------------------------------------------------

Cal_metrics_by_model = Cal_metrics_by_species %>% 
  group_by(model, MWyear, metric) %>%
  summarise(
    mean = mean(val, na.rm = TRUE),
    lo   = lo(val),
    hi   = hi(val)) %>% 
  mutate(model = factor(model, levels = phenology_levels),
         metric = factor(metric, levels = vars))
```

</details>

<details>

<summary>Summarize leave-one-out cross-validation (LOO-CV) models</summary>

```{r, message=F}
#=======================================================
# Summarize model results
#=======================================================



#-------------------------------------------------------
# Compute per-timeseries × model × window metrics
#-------------------------------------------------------

LOOCV_metrics_by_ts <- LOOCV.MW.df %>%
  group_by(species, timeseries, model, MWyear) %>%
  summarise(
    #n     = sum(!is.na(obs) & !is.na(pred)),
    RMSE  = rmse(obs, pred),
    MAE   = mean(abs(obs - pred), na.rm = TRUE),
    #Bias  = mean(pred - obs, na.rm = TRUE),              # + means over-prediction
    r     = suppressWarnings(cor(obs, pred, use = "complete.obs")),
    #R2    = r2_ss(obs, pred),                            # 1 - SSE/SST
    #slope = if (n >= 2) coef(lm(obs ~ pred))[2] else NA_real_,
    #intercept = if (n >= 2) coef(lm(obs ~ pred))[1] else NA_real_,
    #AIC_lm = if (n >= 3) AIC(lm(obs ~ pred)) else NA_real_
  , .groups = "drop")


#-------------------------------------------------------
# Summarise time series-level results at species level
#-------------------------------------------------------

LOOCV_metrics_by_species <- LOOCV_metrics_by_ts %>%
  group_by(species, model, MWyear) %>%
  summarise(
    across(all_of(vars),
           list(. = ~mean(.x, na.rm = TRUE)),
           .names = "{.col}")
  , .groups = "drop") %>% 
  pivot_longer(cols = all_of(vars),
               names_to = "metric",
               values_to = "val") %>% 
  mutate(model  = factor(model, levels = phenology_levels),
         metric = factor(metric, levels = vars))


#-------------------------------------------------------
# Summarise species-level results
#-------------------------------------------------------

LOOCV_metrics_by_model = LOOCV_metrics_by_species %>% 
  group_by(model, MWyear, metric) %>%
  summarise(
    mean = mean(val, na.rm = TRUE),
    lo   = lo(val),
    hi   = hi(val)) %>% 
  mutate(model = factor(model, levels = phenology_levels),
         metric = factor(metric, levels = vars))
```

</details>

<br>

# 3. Create plots

<details>

<summary>Calibration model plots</summary>

```{r}
#-------------------------------------------------------
# Create plot using ggplot2
#-------------------------------------------------------

Cal.plot = Cal_metrics_by_species %>% 
  
  ggplot() + 
  
   # (1) Horizontal reference line at y=0
  geom_hline(yintercept=0.5, color="white") +
  
  # (2) Add ribbon for CI of smoothed mean trend
  geom_ribbon(data=Cal_metrics_by_model,
              aes(x = MWyear, y = mean, ymin = lo, ymax = hi),
              fill = "darkgrey", color=NA, alpha = 0.9) + 
  
  # (3) Add individual species time series
  geom_line(aes(x = MWyear, y = val, group=species, color=species), linewidth = 0.4) +
  
  # (4) Linear trend over the species-level data (dashed)
  geom_smooth(aes(x = MWyear, y = val),
              method='lm', formula = y~x, se = FALSE, 
              linewidth=1, linetype="dashed", color="black") +
  
  # (5) Add overall mean trend line
  geom_line(data=Cal_metrics_by_model,
            aes(x = MWyear, y = mean), color="black", linewidth=1) +
  
  # (6) Regression annotation on the per-species data
  ggpubr::stat_regline_equation(
    aes(x = MWyear, y = val, 
        label = paste(..eq.label.., ..rr.label.., sep = "~~~~")),
    formula = y ~ x,
    label.x.npc = 0.1,  # 2% from left
    label.y.npc = 0.05,  # 2% from top
    show.legend = FALSE
  ) +
  
  # (7) Correlation annotation on the per-species data
  ggpubr::stat_cor(aes(x = MWyear, y = val), 
                   color = "black", 
                   r.accuracy =0.01, p.accuracy =0.001, 
                   show.legend=F,
                   label.x.npc = 0.1,
                   label.y.npc = 0.12) +
  
  # Manual color palette for species
  scale_color_manual("Species", 
                     values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                                '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a'))+
  
  # Zoom in on period of interest
  coord_cartesian(xlim=c(1967.9,2007.1))+
  
  # Axis labels
  xlab("") + ylab('Pearson`s R                   MAE                   RMSE ')+
  
  # Custom x-axis labels showing moving windows
  scale_x_continuous(breaks = seq(1966,2008,by=5),
                     labels = c('1966-1980','1971-1985','1976-1990',
                                '1981-1995','1986-2000','1991-2005',
                                '1996-2010','2001-2015','2006-2020')) +
  
  facet_grid(metric~model, scales="free_y")+
  
  # Apply predefined theme
  plotTheme1


#-------------------------------------------------------
# Save plot as .pdf
#-------------------------------------------------------

ggsave(Cal.plot, file="Calibration.pdf", path=output.dir,
       width=12, height=7.5)


Cal.plot
```

</details>

<details>

<summary>LOO-CV model plots</summary>

```{r}
#-------------------------------------------------------
# Create plot using ggplot2
#-------------------------------------------------------

LOOCV.plot = LOOCV_metrics_by_species %>% 
  
  ggplot() + 
  
   # (1) Horizontal reference line at y=0
  geom_hline(yintercept=0.55, color="white") +
  
  # (2) Add ribbon for CI of smoothed mean trend
  geom_ribbon(data=LOOCV_metrics_by_model,
              aes(x = MWyear, y = mean, ymin = lo, ymax = hi),
              fill = "darkgrey", color=NA, alpha = 0.9) + 
  
  # (3) Add individual species time series
  geom_line(aes(x = MWyear, y = val, group=species, color=species), linewidth = 0.4) +
  
  # (4) Linear trend over the species-level data (dashed)
  geom_smooth(aes(x = MWyear, y = val),
              method='lm', formula = y~x, se = FALSE, 
              linewidth=1, linetype="dashed", color="black") +
  
  # (5) Add overall mean trend line
  geom_line(data=LOOCV_metrics_by_model,
            aes(x = MWyear, y = mean), color="black", linewidth=1) +
  
  # (6) Regression annotation on the per-species data
  ggpubr::stat_regline_equation(
    aes(x = MWyear, y = val, 
        label = paste(..eq.label.., ..rr.label.., sep = "~~~~")),
    formula = y ~ x,
    label.x.npc = 0.1,  # 2% from left
    label.y.npc = 0.05,  # 2% from top
    show.legend = FALSE
  ) +
  
  # (7) Correlation annotation on the per-species data
  ggpubr::stat_cor(aes(x = MWyear, y = val), 
                   color = "black", 
                   r.accuracy =0.01, p.accuracy =0.001, 
                   show.legend=F,
                   label.x.npc = 0.1,
                   label.y.npc = 0.12) +
  
  # Manual color palette for species
  scale_color_manual("Species", 
                     values = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                                '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a'))+
  
  # Zoom in on period of interest
  coord_cartesian(xlim=c(1967.9,2007.1))+
  
  # Axis labels
  xlab("") + ylab('Pearson`s R                   MAE                   RMSE ')+
  
  # Custom x-axis labels showing moving windows
  scale_x_continuous(breaks = seq(1966,2008,by=5),
                     labels = c('1966-1980','1971-1985','1976-1990',
                                '1981-1995','1986-2000','1991-2005',
                                '1996-2010','2001-2015','2006-2020')) +
  
  facet_grid(metric~model, scales="free_y")+
  
  # Apply predefined theme
  plotTheme1


#-------------------------------------------------------
# Save plot as .pdf
#-------------------------------------------------------

ggsave(LOOCV.plot, file="LOO_cross_validation.pdf", path=output.dir,
       width=12, height=7.5)


LOOCV.plot
```

</details>

<br>

# 4. Reproducibility

<details>

<summary>show</summary>

```{r}
# Datetime
Sys.time()

# Session info
sessionInfo()
```

</details>
