---
title: "Spring phenology tier2 project"
subtitle: "Run Bayesian hierarchical models using student distribution and centered response (MODIS data)"
author: "Constantin Zohner"
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

format:
  html:
    toc: false
    theme: default
    highlight-style: haddock
    df-print: paged
    embed-resources: true

execute:
  #warning: false
  message: false
  echo: true
  output: true

always_allow_html: true
editor: visual
editor_options:
  chunk_output_type: inline
---

<br>

### Methods

To isolate interannual, within-site drivers of spring phenology while avoiding spurious associations driven by shared temporal trends or between-site differences, all continuous climate predictors were centered and scaled within site, such that coefficients reflect responses to typical site-level anomalies. Green-up dates were centered within site to remove differences in mean phenological timing among sites.

In the main analyses (see 3.3_Bayesian_models_MODIS.html), green-up anomalies were additionally scaled within site, allowing effect sizes to be interpreted as relative sensitivities (in units of within-site phenological variability) and ensuring equal weighting of sites when comparing the relative importance of climate drivers.

As a robustness check, we refitted all models using unscaled (centered-only) green-up anomalies and a Student-t error distribution, which down-weights occasional extreme observations and provides inference in original phenological units (days). The consistency of results across these complementary model formulations indicates that our conclusions are not driven by site-specific differences in phenological variance or by a small number of anomalous observations.

### Driver names

-   Greenup_prev: Leaf-out date in the previous year
-   LSF2: Binary indicator of late spring frost event (\<-2°C) after leaf-out in previous year
-   MidGreendown_prev: Leaf senescence date in the previous year
-   Tmean: Daily mean air temperature (°C):
    -   Previous year:
        -   EQ1 = March equinox
        -   SO1 = Summer solstice
        -   EQ2 = September equinox
        -   SO2 = December solstice
    -   From 1-17: 1 = January of previous year, 13 = January of current year
-   RH: Spring relative air humidity (%)
-   SWrad: Spring (preseason) incoming short-wave solar radiation (W/m²)
-   Tmax: Spring (preseason) Daily maximum air temperature (°C)

<br>

# 1. Setup

<details>

<summary>Packages</summary>

```{r, message=F}
library(tidyverse)
library(data.table)
library(car)
library(brms)
#install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
#cmdstanr::install_cmdstan()
library(cmdstanr)
options(brms.backend = "cmdstanr")
```

</details>

<details>

<summary>Directories</summary>

```{r}
#-------------------------------------------------
# Define the base directory
#-------------------------------------------------

base.dir = "/Users/goalie/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/MODIS_Analysis"


#-------------------------------------------------
# Build specific directories from the base
#-------------------------------------------------

# Model results
MODIS.dir = file.path(base.dir, "Analysis_input/MODIS_drivers")

# Figure output
output.dir = file.path(base.dir, "Analysis_output/ModelResults")
```

</details>

<details>

<summary>Data</summary>

```{r}
#-----------------------------------------------------------
# define model variables
#-----------------------------------------------------------

# List of responses
response_vars = c("Greenup")

# List of full variables
pred_vars = c(
  "year",
  "Greenup_prev", "MidGreendown_prev",                     # Previous-year phenology data      
  "Tmean.EQ1.SO1", "Tmean.SO1.EQ2", "Tmean.EQ2.SO2",       # Previous-year temperature        
  "Tmean.GU","RH.GU","SWrad.GU"                            # Preseason climate for Greenup dates   
)


#-----------------------------------------------------------
# Load MODIS data
#-----------------------------------------------------------

MODIS.df <- fread(paste(MODIS.dir, "MODIS_drivers_spring_data_preseason.csv", sep="/")) %>% 
  
  dplyr::select(ForestType_Continent, year, s_id, lat, lon, year, LSF2,
                Tmean.13, Tmean.14, Tmean.15,
                all_of(c(response_vars,pred_vars))) %>% 
  
  
  #-----------------------------------------------------------
  # Data preparation
  #-----------------------------------------------------------
  
  # Add mean spring (January to March) temperatures
  mutate(Tmean_spring = rowMeans(
    select(., Tmean.13, Tmean.14, Tmean.15),
    na.rm = TRUE
  )) %>% 

  # Group by pixel (s_id = unique location identifier)
  group_by(s_id) %>%
  
  # Scale each continuous predictor within each pixel
  # This removes across-site variation in mean/variance,
  # ensuring that model coefficients reflect within-site effects over time
  #
  # → Why is this important?
  # Because we're interested in whether interannual variation in predictors
  # (e.g., temperature, RH) explains interannual shifts in phenology,
  # not whether warmer sites leaf out earlier on average.
  
  # In simple or idealized cases, a random-intercept model may effectively estimate 
  # a within-site effect even without centering. However, in more complex settings—especially
  # when predictors differ systematically across groups or are constructed differently across
  # sites—this is no longer guaranteed, and explicit within–between decomposition becomes 
  # necessary for clear interpretation.
  
  # Site-level scaling of variables
  mutate(
    across(
      all_of(c(response_vars, pred_vars, "Tmean_spring")), 
      ~ as.numeric(scale(.x, center = TRUE, scale = TRUE)),
      .names = "{.col}_z")) %>% 
  ungroup() %>% 
  
  # Global scaling of predictor variables
  mutate(
    across(
      all_of(c(response_vars, pred_vars, "Tmean_spring")), 
      ~ as.numeric(scale(.x, center = TRUE, scale = TRUE)),
      .names = "{.col}_z_global"))
```

</details>

<br>

# 2. Run Bayesian hierarchical models

<details>

<summary>Full models</summary>

```{r}
#-----------------------------------------------------------
# Define modeling formulae
#-----------------------------------------------------------

# To account for temporal trends in phenology and potential non-independence of observations across years within each site, 
# we included year as a random slope grouped by timeseries using an uncorrelated random effects structure (year || timeseries). 
# This formulation allows both the baseline phenology (intercept) and the temporal trend (slope of year) to vary across sites, 
# while simplifying model complexity by not estimating the correlation between them.
#
# By doing so, we capture spatial heterogeneity in long-term phenological trends and reduce the risk of inflated fixed-effect 
# estimates caused by unmodeled site-specific trajectories. This specification preserves our ability to detect interannual 
# relationships between climate predictors and leaf-out timing within sites, while controlling for varying temporal baselines 
# across locations in a computationally efficient and statistically robust manner.

# Original with fixed spring temp variable
formula_noSpring <- bf(
  Greenup_z ~ Greenup_prev_z + LSF2 + Tmean.EQ1.SO1_z + Tmean.SO1.EQ2_z + MidGreendown_prev_z + Tmean.EQ2.SO2_z + Tmean_spring_z + RH.GU_z + SWrad.GU_z + (year_z || s_id)
)

# No site-level scaling (global scaling)
formula_noScale <- bf(
  Greenup_z_global ~ Greenup_prev_z_global + LSF2 + Tmean.EQ1.SO1_z_global + Tmean.SO1.EQ2_z_global + MidGreendown_prev_z_global + Tmean.EQ2.SO2_z_global + Tmean.GU_z_global + RH.GU_z_global + SWrad.GU_z_global + (year_z_global || s_id)
)

# no year term
formula_noYear <- bf(
  Greenup_z ~ Greenup_prev_z + LSF2 + Tmean.EQ1.SO1_z + Tmean.SO1.EQ2_z + MidGreendown_prev_z + Tmean.EQ2.SO2_z + Tmean.GU_z + RH.GU_z + SWrad.GU_z + (1 | s_id)
)

#-----------------------------------------------------------
# Setup: model loop across forest types
#-----------------------------------------------------------

ForestType_Continent_list <- unique(MODIS.df$ForestType_Continent)



#===============================================================================================================================



#==================================================
# Mean spring temperature models
#==================================================


#-----------------------------------------------------------
# Define priors
#-----------------------------------------------------------

priors_brm <- c(
  # Fixed effects — weakly informative (small shrinkage)
  set_prior("normal(0, 1)", class = "b"), 
  
  # Intercept — mild regularization (helps model intercept stability)
  set_prior("normal(0, 1)", class = "Intercept"), 
  
   # Residual SD (sigma):
  set_prior("student_t(3, 0, 1)", class = "sigma"),
  
  # General SDs for random effects — conservative, heavy tails
  set_prior("student_t(3, 0, 2.5)", class = "sd"), 
  
  # Specific shrinkage for random slope of year by timeseries
  set_prior("normal(0, 0.5)", class = "sd", group = "s_id", coef = "year_z")
)

#-----------------------------------------------------------
# Loop over each forest type
#-----------------------------------------------------------

models <- list() # Create empty list to store models

for (ft in ForestType_Continent_list) {
  
  cat("Running model for ForestType_Continent:", ft, "\n")
   
  # Create sublist for each forest type
  models[[ft]] <- list()           
    
  # Subset the data for this species and period
  data_sub <- MODIS.df %>% filter(ForestType_Continent == ft)
    
  # Skip if there are too few observations for stable estimation
  if (nrow(data_sub) < 30) {
    cat("Skipping species due to insufficient data:", sp, "\n")
    next
  }
    
  #-----------------------------------------------------------
  # Fit the model using brms and CmdStanR backend
  #-----------------------------------------------------------
    
  models[[ft]] <- brm(
    formula = formula_noSpring,      # The model formula
    data = data_sub,                 # Subset of data for this forest type
    family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
    prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
      
    backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
    #init = 0,                       # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
      
    chains = 4,                      # Number of MCMC chains to run (default is 4)
    cores = 12,                      # Number of CPU cores used in total
    threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
      
    iter = 4000,                     # Total number of iterations per chain (including warmup)
    warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
      
    control = list(
      adapt_delta = 0.95,            # Target acceptance probability for the sampler (higher = fewer divergences)
      max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
    ),
      
    seed = 123                       # Set random seed for reproducibility
  )
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_full_GU_models_noSpring.rds"))



#===============================================================================================================================



#==================================================
# No site-level scaling models
#==================================================


#-----------------------------------------------------------
# Define priors
#-----------------------------------------------------------

priors_brm <- c(
  # Fixed effects — weakly informative (small shrinkage)
  set_prior("normal(0, 1)", class = "b"), 
  
  # Intercept — mild regularization (helps model intercept stability)
  set_prior("normal(0, 1)", class = "Intercept"), 
  
   # Residual SD (sigma):
  set_prior("student_t(3, 0, 1)", class = "sigma"),
  
  # General SDs for random effects — conservative, heavy tails
  set_prior("student_t(3, 0, 2.5)", class = "sd"), 
  
  # Specific shrinkage for random slope of year by timeseries
  set_prior("normal(0, 0.5)", class = "sd", group = "s_id", coef = "year_z_global")
)

#-----------------------------------------------------------
# Loop over each forest type
#-----------------------------------------------------------

models <- list() # Create empty list to store models

for (ft in ForestType_Continent_list) {
  
  cat("Running model for ForestType_Continent:", ft, "\n")
   
  # Create sublist for each forest type
  models[[ft]] <- list()           
    
  # Subset the data for this species and period
  data_sub <- MODIS.df %>% filter(ForestType_Continent == ft)
    
  # Skip if there are too few observations for stable estimation
  if (nrow(data_sub) < 30) {
    cat("Skipping species due to insufficient data:", sp, "\n")
    next
  }
    
  #-----------------------------------------------------------
  # Fit the model using brms and CmdStanR backend
  #-----------------------------------------------------------
    
  models[[ft]] <- brm(
    formula = formula_noScale,       # The model formula
    data = data_sub,                 # Subset of data for this forest type
    family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
    prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
      
    backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
    #init = 0,                       # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
      
    chains = 4,                      # Number of MCMC chains to run (default is 4)
    cores = 12,                      # Number of CPU cores used in total
    threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
      
    iter = 4000,                     # Total number of iterations per chain (including warmup)
    warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
      
    control = list(
      adapt_delta = 0.95,            # Target acceptance probability for the sampler (higher = fewer divergences)
      max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
    ),
      
    seed = 123                       # Set random seed for reproducibility
  )
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_full_GU_models_noScale.rds"))




#===============================================================================================================================



#==================================================
# No-year-term models
#==================================================


#-----------------------------------------------------------
# Define priors
#-----------------------------------------------------------

priors_brm <- c(
  # Fixed effects — weakly informative (small shrinkage)
  set_prior("normal(0, 1)", class = "b"), 
  
  # Intercept — mild regularization (helps model intercept stability)
  set_prior("normal(0, 1)", class = "Intercept"), 
  
   # Residual SD (sigma):
  set_prior("student_t(3, 0, 1)", class = "sigma"),
  
  # General SDs for random effects — conservative, heavy tails
  set_prior("student_t(3, 0, 2.5)", class = "sd")
)

#-----------------------------------------------------------
# Loop over each forest type
#-----------------------------------------------------------

models <- list() # Create empty list to store models

for (ft in ForestType_Continent_list) {
  
  cat("Running model for ForestType_Continent:", ft, "\n")
   
  # Create sublist for each forest type
  models[[ft]] <- list()           
    
  # Subset the data for this species and period
  data_sub <- MODIS.df %>% filter(ForestType_Continent == ft)
    
  # Skip if there are too few observations for stable estimation
  if (nrow(data_sub) < 30) {
    cat("Skipping species due to insufficient data:", sp, "\n")
    next
  }
    
  #-----------------------------------------------------------
  # Fit the model using brms and CmdStanR backend
  #-----------------------------------------------------------
    
  models[[ft]] <- brm(
    formula = formula_noYear,        # The model formula
    data = data_sub,                 # Subset of data for this forest type
    family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
    prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
      
    backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
    #init = 0,                       # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
      
    chains = 4,                      # Number of MCMC chains to run (default is 4)
    cores = 12,                      # Number of CPU cores used in total
    threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
      
    iter = 4000,                     # Total number of iterations per chain (including warmup)
    warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
      
    control = list(
      adapt_delta = 0.95,            # Target acceptance probability for the sampler (higher = fewer divergences)
      max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
    ),
      
    seed = 123                       # Set random seed for reproducibility
  )
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_full_GU_models_noYear.rds"))
```

</details>

<br>

# 3. Reproducibility

<details>

<summary>show</summary>

```{r}
# Datetime
Sys.time()

# Session info
sessionInfo()
```

</details>
