---
title: "Spring phenology tier2 project"
subtitle: "Run Bayesian hierarchical models (MODIS data)"
author: "Constantin Zohner"
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

format:
  html:
    toc: false
    theme: default
    highlight-style: haddock
    df-print: paged
    embed-resources: true

execute:
  #warning: false
  message: false
  echo: true
  output: true

always_allow_html: true
editor: visual
editor_options:
  chunk_output_type: inline
---

<br>

### Driver names

-   MidGreenup_prev: Leaf-out date in the previous year
-   LSF2: Binary indicator of late spring frost event (\<-2°C) after leaf-out in previous year
-   MidGreendown_prev: Leaf senescence date in the previous year
-   Tmean: Daily mean air temperature (°C):
    -   Previous year:
        -   EQ1 = March equinox
        -   SO1 = Summer solstice
        -   EQ2 = September equinox
        -   SO2 = December solstice
    -   From 1-17: 1 = January of previous year, 13 = January of current year
-   RH: Spring relative air humidity (%)
-   SWrad: Spring (preseason) incoming short-wave solar radiation (W/m²)
-   Tmax: Spring (preseason) Daily maximum air temperature (°C)

<br>

# 1. Setup

<details>

<summary>Packages</summary>

```{r, message=F}
library(tidyverse)
library(data.table)
library(car)
library(brms)
#install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
#cmdstanr::install_cmdstan()
library(cmdstanr)
options(brms.backend = "cmdstanr")
```

</details>

<details>

<summary>Directories</summary>

```{r}
#-------------------------------------------------
# Define the base directory
#-------------------------------------------------

base.dir = "/Users/goalie/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/MODIS_Analysis"


#-------------------------------------------------
# Build specific directories from the base
#-------------------------------------------------

# Model results
MODIS.dir = file.path(base.dir, "Analysis_input/MODIS_drivers")

# Figure output
output.dir = file.path(base.dir, "Analysis_output/ModelResults")
```

</details>

<details>

<summary>Data</summary>

```{r}
#-----------------------------------------------------------
# define model variables
#-----------------------------------------------------------

# List of full variables
full_variables = c(
  "year",                                                  # Observation year
  "Greenup","MidGreenup",                                  # Phenology responses     
  "Greenup_prev",  "MidGreenup_prev", "MidGreendown_prev", # Previous-year phenology data      
  "LSF2",                                                  # Previous-year late frost
  "Tmean.EQ1.SO1", "Tmean.SO1.EQ2", "Tmean.EQ2.SO2",       # Previous-year temperature        
  "Tmean.GU","RH.GU","SWrad.GU",                           # Preseason climate for Greenup dates   
  "Tmean.MG","RH.MG","SWrad.MG"                            # Preseason climate for MidGreenup dates     
)

# List of monthly mean temperature variables    
monthly_variables = c(
  "Tmean.8","Tmean.9","Tmean.10","Tmean.11","Tmean.12",           
  "Tmean.13","Tmean.14","Tmean.15","Tmean.16","Tmean.17"
)

# Identify continuous variables to be scaled (excluding LSF2 and monthly variables)
scaled_vars <- setdiff(full_variables, c("LSF2"))


#-----------------------------------------------------------
# Load MODIS data
#-----------------------------------------------------------

MODIS.df <- fread(paste(MODIS.dir, "MODIS_drivers_spring_data_preseason.csv", sep="/")) %>% 
  dplyr::select(ForestType_Continent, year, s_id, lat, lon, 
                all_of(full_variables),
                all_of(monthly_variables)) %>% 
  
  
  #-----------------------------------------------------------
  # Data preparation
  #-----------------------------------------------------------

  mutate(
    
    # Add leaf-out date columns that won't be scaled
    Greenup_DOY    = Greenup,
    MidGreenup_DOY = MidGreenup) %>% 
  
  # Group by pixel (s_id = unique location identifier)
  group_by(s_id) %>%
  
  # Scale each continuous predictor within each pixel
  # This removes across-site variation in mean/variance,
  # ensuring that model coefficients reflect within-site effects over time
  #
  # → Why is this important?
  # Because we're interested in whether interannual variation in predictors
  # (e.g., temperature, RH) explains interannual shifts in phenology,
  # not whether warmer sites leaf out earlier on average.
  #
  # Without this grouping, variables would be scaled globally,
  # and site-level means could dominate effect estimates —
  # falsely inflating signals driven by spatial gradients rather than temporal dynamics. 
  #
  # Example: If warm lowland sites always have earlier leaf-out, a global
  # correlation between Tmean and leaf_out could appear strong,
  # even if within-site interannual correlations are weak or inconsistent.
  mutate(across(all_of(scaled_vars), ~ as.numeric(scale(.x, center = T, scale = T)))) %>%
  
  # Site-level anomalies of predictor variables for monthly models (centered but not scaled)
  mutate(across(all_of(monthly_variables), 
                ~ as.numeric(scale(.x, center = T, scale = F)))) %>%
  
  # Site-level anomalies of response variables for monthly models (centered but not scaled)
  mutate(across(c("Greenup_DOY", "MidGreenup_DOY"), 
                ~ as.numeric(scale(.x, center = T, scale = F)),
                .names = "{.col}_anomaly")) %>%
  
  # Remove group structure
  ungroup()
```

</details>

<br>

# 2. Check multicollinearity using variance inflation factors (VIFs)

<details>

<summary>Full models</summary>

```{r}
#==================================================
# FULL MODELS — Includes all general predictors
#==================================================


#-----------------------------------------------------------
# Greenup dates
#-----------------------------------------------------------

VIF_full_GU.df <- MODIS.df %>%
  group_by(ForestType_Continent) %>%
  group_modify(~ {
    data_subset <- .x
    
    # Construct model formula
    formula <- 
      Greenup ~ Greenup_prev + LSF2 + MidGreendown_prev + 
      Tmean.EQ1.SO1 + Tmean.SO1.EQ2 + Tmean.EQ2.SO2 +
      Tmean.GU + RH.GU + SWrad.GU + year
    
    #--------------------------------------------------
    # Fit model and compute VIFs
    #--------------------------------------------------
    model <- lm(formula, data = data_subset)
    vif_vals <- vif(model)
    
    # Return tidy VIF output
    data.frame(
      Variable = names(vif_vals),
      VIF = as.numeric(vif_vals)
    )
  }) %>%
  ungroup() %>%
  
  # Reshape into wide format: each species gets its own column
  pivot_wider(names_from = ForestType_Continent, values_from = VIF)

# View resulting wide-format VIF table
VIF_full_GU.df


#-----------------------------------------------------------
# Mid-Greenup dates
#-----------------------------------------------------------

VIF_full_MG.df <- MODIS.df %>%
  group_by(ForestType_Continent) %>%
  group_modify(~ {
    data_subset <- .x
    
    # Construct model formula
    formula <- 
      MidGreenup ~ MidGreenup_prev + LSF2 + MidGreendown_prev + 
      Tmean.EQ1.SO1 + Tmean.SO1.EQ2 + Tmean.EQ2.SO2 +
      Tmean.MG + RH.MG + SWrad.MG + year
    
    #--------------------------------------------------
    # Fit model and compute VIFs
    #--------------------------------------------------
    model <- lm(formula, data = data_subset)
    vif_vals <- vif(model)
    
    # Return tidy VIF output
    data.frame(
      Variable = names(vif_vals),
      VIF = as.numeric(vif_vals)
    )
  }) %>%
  ungroup() %>%
  
  # Reshape into wide format: each species gets its own column
  pivot_wider(names_from = ForestType_Continent, values_from = VIF)

# View resulting wide-format VIF table
VIF_full_MG.df
```

</details>

<details>

<summary>Monthly models</summary>

```{r}
#==================================================
# MONTHLY MODELS — Only include monthly temperature
# predictors up to the phenologically relevant month
#==================================================


#-----------------------------------------------------------
# Greenup dates
#-----------------------------------------------------------

VIF_monthly_GU.df <- MODIS.df %>%
  group_by(ForestType_Continent) %>%
  group_map(~ {
    
    data_subset <- .x                  # Subset of data for the current species
    ForestType_Continent_name <- .y$ForestType_Continent   # Current Forest type 
    
    # Skip if too few observations
    if (nrow(data_subset) < 30) return(NULL)
    
    #--------------------------------------------------
    # Determine latest Tmean month to include
    #--------------------------------------------------
    mean_doy <- mean(data_subset$Greenup_DOY, na.rm = TRUE)
    
    # Estimate approximate month from DOY (1 = Jan, ..., 5 = May)
    latest_month <- floor((mean_doy - 1) / 30) + 1
    
    # Only include a month if leaf-out occurs in its second half
    if (mean_doy < ((latest_month - 1) * 30 + 15)) {
      latest_month <- latest_month - 1  # Exclude month if leaf-out too early
    }
    
    # Translate to Tmean variable names: Tmean.8 = August previous year
    latest_tmean <- min(12 + latest_month, 17)
    tmean_vars <- if (latest_tmean >= 8) paste0("Tmean.", 8:latest_tmean) else character(0)
    
    # Include only if we have at least one Tmean variable
    predictors <- c(tmean_vars, "year")
    if (length(predictors) == 0) return(NULL)
    
    #--------------------------------------------------
    # Fit model and compute VIFs
    #--------------------------------------------------
    formula <- as.formula(paste("Greenup_DOY_anomaly ~", paste(predictors, collapse = " + ")))
    
    model <- tryCatch(lm(formula, data = data_subset), error = function(e) return(NULL))
    if (is.null(model)) return(NULL)
    
    vif_vals <- tryCatch(vif(model), error = function(e) return(NULL))
    if (is.null(vif_vals) || length(vif_vals) == 0) return(NULL)
    
    # Return tidy output with species column
    data.frame(
      Variable = names(vif_vals),
      VIF = as.numeric(vif_vals),
      ForestType_Continent = ForestType_Continent_name
    )
  }) %>%
  bind_rows() %>%                     # Combine species-wise results into one data frame
  pivot_wider(                        # Reshape: one column per species
    names_from = ForestType_Continent,
    values_from = VIF
  )

# View resulting wide-format VIF table
VIF_monthly_GU.df


#-----------------------------------------------------------
# Mid-Greenup dates
#-----------------------------------------------------------

VIF_monthly_MG.df <- MODIS.df %>%
  group_by(ForestType_Continent) %>%
  group_map(~ {
    
    data_subset <- .x                  # Subset of data for the current species
    ForestType_Continent_name <- .y$ForestType_Continent   # Current Forest type 
    
    # Skip if too few observations
    if (nrow(data_subset) < 30) return(NULL)
    
    #--------------------------------------------------
    # Determine latest Tmean month to include
    #--------------------------------------------------
    mean_doy <- mean(data_subset$MidGreenup_DOY, na.rm = TRUE)
    
    # Estimate approximate month from DOY (1 = Jan, ..., 5 = May)
    latest_month <- floor((mean_doy - 1) / 30) + 1
    
    # Only include a month if leaf-out occurs in its second half
    if (mean_doy < ((latest_month - 1) * 30 + 15)) {
      latest_month <- latest_month - 1  # Exclude month if leaf-out too early
    }
    
    # Translate to Tmean variable names: Tmean.8 = August previous year
    latest_tmean <- min(12 + latest_month, 17)
    tmean_vars <- if (latest_tmean >= 8) paste0("Tmean.", 8:latest_tmean) else character(0)
    
    # Include only if we have at least one Tmean variable
    predictors <- c(tmean_vars, "year")
    if (length(predictors) == 0) return(NULL)
    
    #--------------------------------------------------
    # Fit model and compute VIFs
    #--------------------------------------------------
    formula <- as.formula(paste("MidGreenup_DOY_anomaly ~", paste(predictors, collapse = " + ")))
    
    model <- tryCatch(lm(formula, data = data_subset), error = function(e) return(NULL))
    if (is.null(model)) return(NULL)
    
    vif_vals <- tryCatch(vif(model), error = function(e) return(NULL))
    if (is.null(vif_vals) || length(vif_vals) == 0) return(NULL)
    
    # Return tidy output with species column
    data.frame(
      Variable = names(vif_vals),
      VIF = as.numeric(vif_vals),
      ForestType_Continent = ForestType_Continent_name
    )
  }) %>%
  bind_rows() %>%                     # Combine species-wise results into one data frame
  pivot_wider(                        # Reshape: one column per species
    names_from = ForestType_Continent,
    values_from = VIF
  )

# View resulting wide-format VIF table
VIF_monthly_MG.df
```

</details>

<br>

# 3. Run Bayesian hierarchical models

<details>

<summary>Full models</summary>

```{r}
#==================================================
# Define priors and formulae
#==================================================


#-----------------------------------------------------------
# Define priors
#-----------------------------------------------------------

priors_brm <- c(
  # Fixed effects — weakly informative (small shrinkage)
  set_prior("normal(0, 1)", class = "b"), 
  
  # Intercept — mild regularization (helps model intercept stability)
  set_prior("normal(0, 1)", class = "Intercept"), 
  
   # Residual SD (sigma):
  set_prior("student_t(3, 0, 1)", class = "sigma"),
  
  # General SDs for random effects — conservative, heavy tails
  set_prior("student_t(3, 0, 2.5)", class = "sd"), 
  
  # Specific shrinkage for random slope of year by timeseries
  set_prior("normal(0, 0.5)", class = "sd", group = "s_id", coef = "year")
)

#-----------------------------------------------------------
# Define modeling formulae
#-----------------------------------------------------------

# To account for temporal trends in phenology and potential non-independence of observations across years within each site, 
# we included year as a random slope grouped by timeseries using an uncorrelated random effects structure (year || timeseries). 
# This formulation allows both the baseline phenology (intercept) and the temporal trend (slope of year) to vary across sites, 
# while simplifying model complexity by not estimating the correlation between them.
#
# By doing so, we capture spatial heterogeneity in long-term phenological trends and reduce the risk of inflated fixed-effect 
# estimates caused by unmodeled site-specific trajectories. This specification preserves our ability to detect interannual 
# relationships between climate predictors and leaf-out timing within sites, while controlling for varying temporal baselines 
# across locations in a computationally efficient and statistically robust manner.

# Greenup models
formula_GU <- bf(
  Greenup ~ Greenup_prev + LSF2 + Tmean.EQ1.SO1 + Tmean.SO1.EQ2 + MidGreendown_prev +
    Tmean.EQ2.SO2 + Tmean.GU + RH.GU + SWrad.GU + (year || s_id)
)

# MidGreenup models
formula_MG <- bf(
  MidGreenup ~ MidGreenup_prev + LSF2 + Tmean.EQ1.SO1 + Tmean.SO1.EQ2 + MidGreendown_prev +
    Tmean.EQ2.SO2 + Tmean.MG + RH.MG + SWrad.MG + (year || s_id)
)

#-----------------------------------------------------------
# Setup: model loop across forest types
#-----------------------------------------------------------

ForestType_Continent_list <- unique(MODIS.df$ForestType_Continent)



#===============================================================================================================================



#==================================================
# Greenup date models
#==================================================


#-----------------------------------------------------------
# Loop over each forest type
#-----------------------------------------------------------

models <- list() # Create empty list to store models

for (ft in ForestType_Continent_list) {
  
  cat("Running model for ForestType_Continent:", ft, "\n")
   
  # Create sublist for each forest type
  models[[ft]] <- list()           
    
  # Subset the data for this species and period
  data_sub <- MODIS.df %>% filter(ForestType_Continent == ft)
    
  # Skip if there are too few observations for stable estimation
  if (nrow(data_sub) < 30) {
    cat("Skipping species due to insufficient data:", sp, "\n")
    next
  }
    
  #-----------------------------------------------------------
  # Fit the model using brms and CmdStanR backend
  #-----------------------------------------------------------
    
  models[[ft]] <- brm(
    formula = formula_GU,            # The model formula
    data = data_sub,                 # Subset of data for this forest type
    family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
    prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
      
    backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
    #init = 0,                        # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
      
    chains = 4,                      # Number of MCMC chains to run (default is 4)
    cores = 12,                      # Number of CPU cores used in total
    threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
      
    iter = 4000,                     # Total number of iterations per chain (including warmup)
    warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
      
    control = list(
      adapt_delta = 0.95,            # Target acceptance probability for the sampler (higher = fewer divergences)
      max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
    ),
      
    seed = 123                       # Set random seed for reproducibility
  )
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_full_GU_models.rds"))



#===============================================================================================================================



#==================================================
# Mid-Greenup date models
#==================================================


#-----------------------------------------------------------
# Loop over each forest type
#-----------------------------------------------------------

models <- list() # Create empty list to store models

for (ft in ForestType_Continent_list) {
  
  cat("Running model for ForestType_Continent:", ft, "\n")
  
  # Create sublist for each forest type
  models[[ft]] <- list()           
  
  # Subset the data for this species and period
  data_sub <- MODIS.df %>% filter(ForestType_Continent == ft)
  
  # Skip if there are too few observations for stable estimation
  if (nrow(data_sub) < 30) {
    cat("Skipping species due to insufficient data:", sp, "\n")
    next
  }
  
  #-----------------------------------------------------------
  # Fit the model using brms and CmdStanR backend
  #-----------------------------------------------------------
  
  models[[ft]] <- brm(
    formula = formula_MG,            # The model formula
    data = data_sub,                 # Subset of data for this forest type
    family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
    prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
    
    backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
    #init = 0,                        # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
    
    chains = 4,                      # Number of MCMC chains to run (default is 4)
    cores = 12,                      # Number of CPU cores used in total
    threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
    
    iter = 4000,                     # Total number of iterations per chain (including warmup)
    warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
    
    control = list(
      adapt_delta = 0.95,            # Target acceptance probability for the sampler (higher = fewer divergences)
      max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
    ),
    
    seed = 123                       # Set random seed for reproducibility
  )
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_full_MG_models.rds"))
```

</details>

<details>

<summary>Monthly models</summary>

```{r}
#==================================================
# Define phenology cut-offs and formulae
#==================================================


#-----------------------------------------------------------
# Calculate average leaf-out DOY per forest type (across all years)
# Used to determine which monthly temperature predictors to include
#-----------------------------------------------------------

ForestType_Continent_leaf_out = MODIS.df %>% 
  group_by(ForestType_Continent) %>% 
  summarise(mean_Greenup    = mean(Greenup_DOY, na.rm = TRUE),
            mean_MidGreenup = mean(MidGreenup_DOY, na.rm = TRUE)) %>%
  arrange(mean_Greenup)
ForestType_Continent_leaf_out

#-----------------------------------------------------------
# Define priors
#-----------------------------------------------------------

priors_brm <- c(
  # Fixed effects — weakly informative (small shrinkage)
  set_prior("normal(0, 1.5)", class = "b"), 
  
  # Intercept — mild regularization (helps model intercept stability)
  set_prior("normal(0, 2)", class = "Intercept"), 
  
   # Residual SD (sigma):
  set_prior("student_t(3, 0, 3)", class = "sigma"),
  
  # General SDs for random effects — conservative, heavy tails
  set_prior("student_t(3, 0, 3)", class = "sd"), 
  
  # Specific shrinkage for random slope of year by timeseries
  set_prior("normal(0, 0.5)", class = "sd", group = "s_id", coef = "year")
)

#-----------------------------------------------------------
# Setup model structure
#-----------------------------------------------------------

ForestType_Continent_list <- unique(MODIS.df$ForestType_Continent)



#===============================================================================================================================



#==================================================
# Greenup date models
#==================================================


#-----------------------------------------------------------
# Loop over each period × species combination
#-----------------------------------------------------------

models <- list()
  
for (ft in ForestType_Continent_list) {
    
  cat("Running model for Forest Type:", ft, "\n")
    
  # Create sublist for each forest type
  models[[ft]] <- list()
    
  # Subset data for this species and period
  data_sub <- MODIS.df %>% filter(ForestType_Continent == ft)
    
  # Skip model fitting if insufficient data
  if (nrow(data_sub) < 30) {
    cat("Skipping species due to insufficient data:", ft, "\n")
    next
  }
    
  #-----------------------------------------------------------
  # Select monthly temperature predictors based on leaf-out timing
  #-----------------------------------------------------------
    
  # Get forest type-specific average leaf-out day
  mean_doy <- ForestType_Continent_leaf_out[ForestType_Continent_leaf_out$ForestType_Continent==ft,]$mean_Greenup
    
  # Convert DOY to approximate month (1 = Jan, 5 = May, etc.)
  latest_month <- floor((mean_doy - 1) / 30) + 1
    
  # Only include monthly temperature if leaf-out occurs in the second half of that month
  if (mean_doy < ((latest_month - 1) * 30 + 15)) {
    latest_month <- latest_month - 1  # Too early → exclude current month
  }
    
  # Convert calendar month to Tmean.X format (Tmean.8 = August prev year)
  latest_tmean <- 12 + latest_month  # e.g., March → Tmean.15
    
  # Dynamically build Tmean term string (e.g., Tmean.8 + Tmean.9 + ...)
  tmean_terms <- paste0("Tmean.", 8:latest_tmean, collapse = " + ")
    
  # Build model formula string including only relevant months + uncorrelated random slopes for year by timeseries
  formula_string <- paste("Greenup_DOY_anomaly ~", tmean_terms, "+ (year || s_id)")
  formula <- bf(as.formula(formula_string)) # Convert to brms formula
    
  #-----------------------------------------------------------
  # Fit the Bayesian model
  #-----------------------------------------------------------
    
  models[[ft]] <- brm(
    formula = formula,               # The model formula, chosen dynamically
    data = data_sub,                 # Subset of data for this species and period
    family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
    prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
      
    backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
    #init = 0,                        # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
      
    chains = 4,                      # Number of MCMC chains to run (default is 4)
    cores = 12,                      # Number of CPU cores used in total
    threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
      
    iter = 4000,                     # Total number of iterations per chain (including warmup)
    warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
      
    control = list(
    adapt_delta = 0.95,              # Target acceptance probability for the sampler (higher = fewer divergences)
      max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
    ),
      
    seed = 123                       # Set random seed for reproducibility
  )
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_monthly_GU_models.rds"))



#===============================================================================================================================



#==================================================
# Mid-Greenup date models
#==================================================


#-----------------------------------------------------------
# Loop over each period × species combination
#-----------------------------------------------------------

models <- list()

for (ft in ForestType_Continent_list) {
  
  cat("Running model for Forest Type:", ft, "\n")
  
  # Create sublist for each forest type
  models[[ft]] <- list()
  
  # Subset data for this species and period
  data_sub <- MODIS.df %>% filter(ForestType_Continent == ft)
  
  # Skip model fitting if insufficient data
  if (nrow(data_sub) < 30) {
    cat("Skipping species due to insufficient data:", ft, "\n")
    next
  }
  
  #-----------------------------------------------------------
  # Select monthly temperature predictors based on leaf-out timing
  #-----------------------------------------------------------
  
  # Get forest type-specific average leaf-out day
  mean_doy <- ForestType_Continent_leaf_out[ForestType_Continent_leaf_out$ForestType_Continent==ft,]$mean_MidGreenup
  
  # Convert DOY to approximate month (1 = Jan, 5 = May, etc.)
  latest_month <- floor((mean_doy - 1) / 30) + 1
  
  # Only include monthly temperature if leaf-out occurs in the second half of that month
  if (mean_doy < ((latest_month - 1) * 30 + 15)) {
    latest_month <- latest_month - 1  # Too early → exclude current month
  }
  
  # Convert calendar month to Tmean.X format (Tmean.8 = August prev year)
  latest_tmean <- 12 + latest_month  # e.g., March → Tmean.15
  
  # Dynamically build Tmean term string (e.g., Tmean.8 + Tmean.9 + ...)
  tmean_terms <- paste0("Tmean.", 8:latest_tmean, collapse = " + ")
  
  # Build model formula string including only relevant months + uncorrelated random slopes for year by timeseries
  formula_string <- paste("MidGreenup_DOY_anomaly ~", tmean_terms, "+ (year || s_id)")
  formula <- bf(as.formula(formula_string)) # Convert to brms formula
  
  #-----------------------------------------------------------
  # Fit the Bayesian model
  #-----------------------------------------------------------
  
  models[[ft]] <- brm(
    formula = formula,               # The model formula, chosen dynamically
    data = data_sub,                 # Subset of data for this species and period
    family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
    prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
    
    backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
    #init = 0,                        # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
    
    chains = 4,                      # Number of MCMC chains to run (default is 4)
    cores = 12,                      # Number of CPU cores used in total
    threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
    
    iter = 4000,                     # Total number of iterations per chain (including warmup)
    warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
    
    control = list(
      adapt_delta = 0.95,              # Target acceptance probability for the sampler (higher = fewer divergences)
      max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
    ),
    
    seed = 123                       # Set random seed for reproducibility
  )
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_monthly_MG_models.rds"))
```

</details>

<br>

# 4. Reproducibility

<details>

<summary>show</summary>

```{r}
# Datetime
Sys.time()

# Session info
sessionInfo()
```

</details>
