---
title: Spring phenology tier2 project
author: Constantin Zohner
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

subtitle: R code generating the MODIS drivers
output:
  html_document:
    highlight: haddock
    toc: false
    use_bookdown: true
    df_print: paged
---

<style>
body
  { counter-reset: source-line 0; }
pre.numberSource code
  { counter-reset: none; }
</style>

<br>

### Driver names
- Tmean: Mean daily temperature
- Tmin: Minimum daily temperature
- Tmax: Maximum daily temperature
- SWrad: Net short-wave radiation
- RH: Relative humidity
- LSF: Late frost (-2°C or -4°C) occurrence in previous year 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  class.source = "numberLines lineAnchors"
  )
```

<br><br>

## 1. Readme, load packages and data
<details><summary>Readme</summary>
```{r, message=F, warning=F,  attr.source='.numberLines'}
################################################################################
# Climate Driver Extraction for MODIS Phenology Time Series
#
# Description:
# This script extracts and summarizes daily climate variables from GLDAS data
# for MODIS-derived phenological events (e.g. green-up) across forest sites.
# The result is a site-year table of seasonal, monthly, and preseason climate
# drivers associated with spring phenology timing.
#
# Input Data:
# - MODIS phenology data (Greenup, MidGreenup, etc.) from MODIS_data_clean.csv
# - Daily climate data (Tmean, Tmin, Tmax, Precip, Qair, SWrad) from GLDAS
# - Daily photoperiod data (DL) from external photoperiod model
#
# Climate variables used:
# - Tmean: Daily mean temperature (°C)
# - Tmin: Daily minimum temperature (°C)
# - Tmax: Daily maximum temperature (°C)
# - Prcp: Daily precipitation (mm/day)
# - Qair: Specific humidity (converted to RH)
# - RH: Relative humidity (derived)
# - SWrad: Shortwave radiation (W/m²)
# - DL: Daylength (photoperiod, hours)
#
# Outputs:
# - CSV: MODIS_drivers_spring_data.csv
#     A site-year table of all extracted and derived variables
#
# Output Variables Include:
# - Preseason means (Tmean.PS.GU10, ..., Tmean.PS.GU120, etc.)
# - Frost event flags (LSF2, LSF4)
# - Seasonal means by solstice and equinox intervals
# - Monthly climate averages (Tmean.1, ..., RH.18)
#
# Parallelization:
# - Script is parallelized using pbmclapply over all unique site-year time series
# - Cores: 12 (adjust as needed)
#
# Error Handling:
# - Failed site-years are caught via tryCatch and logged as RDS files for debugging
#
# Author: Constantin Zohner
# Institute: Crowther Lab, ETH Zürich
# Date created: 9 June 2025
#
# Required R packages: data.table, dplyr, purrr, pbmcapply, zoo, ggplot2
#
# Usage:
# 1. Define input and output paths
# 2. Ensure GLDAS and MODIS data are cleaned and formatted correctly
# 3. Run the script end-to-end or adapt specific components
################################################################################
```

</details>

<details><summary>Packages</summary>
```{r, message=F, warning=F,  attr.source='.numberLines'}
library(data.table)
library(tidyverse)
library(purrr)
library(pbmcapply)
library(ggplot2)
```
</details>

<details><summary>Helper functions</summary>
```{r}
#-------------------------------------------------------------
# Convert specific to relative humidity
#-------------------------------------------------------------

qair2rh <- function(qair, temp, press = 1013.25) {
  es <- 6.112 * exp((17.67 * temp)/(temp + 243.5))
  e <- qair * press / (0.378 * qair + 0.622)
  rh <- e / es
  rh[rh > 1] <- 1
  rh[rh < 0] <- 0
  return(rh)
}
```
</details>

<details><summary>Load and prepare data</summary>
```{r}
################
# Define Paths #
################



base.dir <- "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/MODIS_Analysis"


# 1. Input
##########


#Climate
GLDAS.dir <- file.path(base.dir, "Analysis_input/GLDAS_data")

#Photoperiod
photo.dir <- file.path(base.dir, "Analysis_input/Photoperiod")

#Phenology
MODIS.dir <- file.path(base.dir, "MODIS_data")


# 2. Output
###########


MODIS_drivers.dir <- file.path(base.dir, "/Analysis_input/MODIS_drivers")
Missing.dir       <- file.path(base.dir, "/Analysis_input/MODIS_drivers/Missing_observations")
error_log.dir     <- file.path(base.dir, "/Analysis_input/MODIS_drivers/Error_logs")



##############################################################################################################################################
##############################################################################################################################################



###########################
# Import and prepate data #
###########################


#-------------------------------------------------------------
# Phenology data
#-------------------------------------------------------------

MODIS.df <- fread(file.path(MODIS.dir, "MODIS_data_clean.csv")) %>%
  group_by(s_id) %>%
  mutate(
    # unique pixel-year identifier
    site_year = paste0(s_id,"_",year),
    # Calculate pixel-level mean greenup dates
    Greenup_mean = round(mean(Greenup)),
    MidGreenup_mean = round(mean(MidGreenup))) %>%
  ungroup()


#-------------------------------------------------------------
# Photoperiod
#-------------------------------------------------------------

photo.df <- fread(file.path(photo.dir, "Photoperiod.csv"))


#-------------------------------------------------------------
# Import daily climatic datasets from GLDAS
#-------------------------------------------------------------

# Load core climate variables from files
list.files(GLDAS.dir)
vn <- c(
  "Daily_Mean_Data_Tair_f_inst_red.csv", # daily mean temperature 
  "Daily_Min_Data_Tair_f_inst_red.csv",  # daily minimum temperature
  "Daily_Max_Data_Tair_f_inst_red.csv",  # daily maximum temperature
  "Daily_Data_Rainf_f_tavg_red.csv",     # daily precipitation sum 
  "Daily_Data_Qair_f_inst_red.csv",      # daily mean specific humidity 
  "Daily_Data_Swnet_tavg_red.csv"        # global radiation
)  
# not used: 
# "Daily_Data_SoilMoi10_40cm_inst_red.csv" (soil moisture)

# Read into named list
DataList <- lapply(file.path(GLDAS.dir, vn), fread)


#-------------------------------------------------------------
# Merge data and add names to list
#-------------------------------------------------------------

# Add additional datasets
DataList[[length(DataList) + 1]] <- photo.df
DataList[[length(DataList) + 1]] <- MODIS.df

# Set all names
names(DataList) = c(vn, "photoperiod", "MODIS")
names(DataList)
```
</details>

<br>

## 2. Parallel computing
<details><summary>Create loop</summary>
```{r}
###########################################################
# Loop through each observations using parallel computing #
###########################################################



# Define time series by unique site-year combinations
timeseries_year <- unique(MODIS.df$site_year)

# Parallel processing function
parallelCalc <- function(timeseries_years, DataList) {
 
  #-------------------------------------------------------------
  # extract MODIS data per site x year combination
  #-------------------------------------------------------------
  
  pheno.sub <- DataList[["MODIS"]][DataList[["MODIS"]]$site_year == timeseries_years, ]
  # Skip if doesn't exist
  if (nrow(pheno.sub) == 0) return(NULL)
  
  
  #-------------------------------------------------------------
  # Define the length of the period of interest in calendar units
  #-------------------------------------------------------------
  
  year1 <- as.character(pheno.sub$year - 1)
  year2 <- as.character(pheno.sub$year)
  start_doy <- paste0(year1, "-01-01")
  end_doy <- paste0(year2, "-06-30")
  days <- seq(as.Date(start_doy), as.Date(end_doy), by = "days")
  
  #get number of days in year 1 and year 2
  days_year1 <- length(seq(as.Date(start_doy), as.Date(paste0(year1, "-12-31")), by = "days"))
  days_year2 <- length(seq(as.Date(paste0(year2, "-01-01")), as.Date(end_doy), by = "days"))
  
  
  #-------------------------------------------------------------
  # Load and merge all climate variables
  #-------------------------------------------------------------
  
  tryCatch({
    
    # Function to load climate data
    get_subdata <- function(index, var_index) {
      # Year 1
      sub1 <- DataList[[var_index]][DataList[[var_index]]$lat == pheno.sub$lat &
                                      DataList[[var_index]]$lon == pheno.sub$lon &
                                      DataList[[var_index]]$year == year1, as.character(1:days_year1), with = FALSE]
      # Year 2
      sub2 <- DataList[[var_index]][DataList[[var_index]]$site_year == pheno.sub$site_year, as.character(1:days_year2), with = FALSE]
      if (nrow(sub1) == 0 || nrow(sub2) == 0) return(NULL)
      cbind(sub1, sub2)
    }
    
    # ------------------------------------------------------------------------
    # Short Name        ; Long Name                               ; Unit [d-1]
    # ------------------------------------------------------------------------
    # SWRAD             ; Net short wave radiation flux           ; W m-2
    # PRCP              ; Precipitation rate                      ; mm d-1
    # Qair              ; Specific humidity                       ; kg kg-1
    # TMIN              ; Minimum Air Temperature                 ; degC
    # TMEAN             ; Mean Air Temperature                    ; degC
    # TMAX              ; Maximum Air Temperature                 ; degC
    # ------------------------------------------------------------------------
    
    Tmean <- get_subdata(pheno.sub, 1)
    Tmin  <- get_subdata(pheno.sub, 2)
    Tmax  <- get_subdata(pheno.sub, 3)
    Prcp  <- get_subdata(pheno.sub, 4)
    Qair  <- get_subdata(pheno.sub, 5)
    SWrad <- get_subdata(pheno.sub, 6)
    DL    <- cbind(
      DataList[["photoperiod"]][DataList[["photoperiod"]]$lat == pheno.sub$lat, as.character(1:days_year1), with = FALSE][1],
      DataList[["photoperiod"]][DataList[["photoperiod"]]$lat == pheno.sub$lat, as.character(1:days_year2), with = FALSE][1]
    )
    
    if (any(sapply(list(Tmean, Tmin, Tmax, Prcp, Qair, SWrad, DL), is.null))) {
      write.table(pheno.sub,
                  file = file.path(Missing.dir, paste0(timeseries_years, ".csv")),
                  sep = ",", row.names = FALSE, col.names = TRUE)
      return(NULL)
    }

    
    #-------------------------------------------------------------
    # Table of climate variables
    #-------------------------------------------------------------
    
    daily_vals <- data.frame(
      Year  = lubridate::year(days),
      Month = lubridate::month(days),
      Day   = lubridate::day(days),
      DOY   = 1:length(days),
      Tmin  = as.numeric(Tmin),
      Tmean = as.numeric(Tmean),
      Tmax  = as.numeric(Tmax),
      SWrad = as.numeric(SWrad),
      Prcp  = as.numeric(Prcp),
      Qair  = as.numeric(Qair),
      DL    = as.numeric(DL)
    ) %>%
      # Add relative humidity (%)
      mutate(RH = qair2rh(Qair, Tmean) * 100) %>% 
      dplyr::select(-c(Qair))
     
    
    #-------------------------------------------------------------
    # Generate sub-dataframe to store results
    #-------------------------------------------------------------
    
    factors.sub <- pheno.sub %>% dplyr::select(s_id, year, lat, lon, Biome, LC, ForestType,
                                        Greenup,Greenup_mean, 
                                        MidGreenup,MidGreenup_mean,
                                        Greenup_prev,MidGreenup_prev,MidGreendown_prev)
    
    
    ##############################################################################################################################################
    
    
    #-------------------------------------------------------------
    # Get key dates
    #-------------------------------------------------------------
    
    # Leaf-out date of previous year
    Greenup_prev_DOY = pheno.sub$Greenup_prev
    
    # Row of mean Greenup and MidGreenup date
    Greenup_mean_DOY = days_year1 + pheno.sub$Greenup_mean
    MidGreenup_mean_DOY = days_year1 + pheno.sub$MidGreenup_mean  
    
    # Longest and shortest day of previous year
    solstice1    = which(daily_vals$DL==max(daily_vals$DL))[1] 
    solstice2    = solstice1 + which.min(daily_vals$DL[(solstice1 + 1):nrow(daily_vals)])
    
    # March & Septmeber equinox of previous year
    equinox1     = solstice1 - 93       
    equinox2     = solstice2 - 90      
    
    # May 1 + Sep 30 of previous year (Growing season as in Lei Chen's paper)
    May1         = which(daily_vals$Day==1 & daily_vals$Month==5 & daily_vals$Year==year1)
    Sep30        = which(daily_vals$Day==30 & daily_vals$Month==9 & daily_vals$Year==year1)      
    
    # July 15 (Late frost threshold date)
    Jul15        = which(daily_vals$Day==15 & daily_vals$Month==7 & daily_vals$Year==year1)
    
    
    ##############################################################################################################################################
    
    
    #-------------------------------------------------------------
    # Late spring frost in previous year
    #-------------------------------------------------------------
    
    
    # was there a frost event below -2°C from leaf-out until July 15?
    #-------------------------------------------------------------
    factors.sub$LSF2 = ifelse(any(daily_vals$Tmin[Greenup_prev_DOY:Jul15] < -2, na.rm=T), 1, 0)
    
    # was there a frost event below -4°C from leaf-out until July 15?
    #-------------------------------------------------------------
    factors.sub$LSF4 = ifelse(any(daily_vals$Tmin[Greenup_prev_DOY:Jul15] < -4, na.rm=T), 1, 0)
    

    ##############################################################################################################################################
    
    
    #-------------------------------------------------------------
    # Seasonal drivers
    #-------------------------------------------------------------
    
    
    #define variables
    variable.names <- c("Tmin", "Tmean", "Tmax", "SWrad", "Prcp", "RH")
    for (i in 1:length(variable.names)) {
      variable <- daily_vals[, variable.names[i]]
      # EQ1...March equinox previous year
      # SO1...Summer solstice (~22 June)
      # EQ2...September equinox previous year
      # SO2...Winter solstice (~22 Dec)
      # GS...period from May 1 to September 30
      factors.sub[[paste0(variable.names[i], ".GS")]] <- mean(variable[May1:Sep30])
      factors.sub[[paste0(variable.names[i], ".EQ1.SO1")]] <- mean(variable[equinox1:solstice1])
      factors.sub[[paste0(variable.names[i], ".SO1.EQ2")]] <- mean(variable[solstice1:equinox2])
      factors.sub[[paste0(variable.names[i], ".EQ2.SO2")]] <- mean(variable[equinox2:solstice2])
    }
    
    
    ##############################################################################################################################################
    
    
    #----------------------------------------------------------------------------
    # Monthly climate means: assign months 1–12 (year1) and 13–18 (year2 Jan–Jun)
    #----------------------------------------------------------------------------
    
    
    # Filter to relevant months
    monthly_vals <- daily_vals %>%
      filter((Year == as.integer(year1) & Month %in% 1:12) |
               (Year == as.integer(year2) & Month %in% 1:6)) %>%
      mutate(month_index = ifelse(Year == as.integer(year1), Month, Month + 12))
    
    # Loop over variables and compute per-month means
    for (var in c("Tmin", "Tmean", "Tmax", "SWrad", "RH")) {
      
      # Monthly mean for this variable
      var_monthly <- monthly_vals %>%
        group_by(month_index) %>%
        summarise(mean_val = mean(.data[[var]], na.rm = TRUE)) %>%
        complete(month_index = 1:18, fill = list(mean_val = NA)) %>%
        arrange(month_index)
      
      # Add as columns
      col_names <- paste0(var, ".", 1:18)
      values <- as.list(setNames(var_monthly$mean_val, col_names))
      factors.sub <- bind_cols(factors.sub, as_tibble_row(values))
    }
    
    
    ##############################################################################################################################################
    
    
    #-------------------------------------------------------------
    # Preseason means for each climate driver
    #-------------------------------------------------------------
    
    
    #get preseason length vector (10 to 120 days with 10-day steps)
    preseason.lengths = seq(10, 120, 10)
    
    
    #-------------------------------------------------------------
    # Calculate preseason means before Greenup
    #-------------------------------------------------------------
    
    for (len in preseason.lengths) {
      if ((Greenup_mean_DOY - len) >= 1) {
        idx <- (Greenup_mean_DOY - len):Greenup_mean_DOY
        factors.sub[[paste0("Tmean.PS.GU", len)]] <- mean(daily_vals$Tmean[idx], na.rm = TRUE)
        factors.sub[[paste0("Tmax.PS.GU", len)]]  <- mean(daily_vals$Tmax[idx],  na.rm = TRUE)
        factors.sub[[paste0("Tmin.PS.GU", len)]]  <- mean(daily_vals$Tmin[idx],  na.rm = TRUE)
        factors.sub[[paste0("SWrad.PS.GU", len)]] <- mean(daily_vals$SWrad[idx], na.rm = TRUE)
        factors.sub[[paste0("RH.PS.GU", len)]]    <- mean(daily_vals$RH[idx],    na.rm = TRUE)
        }
      }
    
    #-------------------------------------------------------------
    # Calculate preseason means before MidGreenup
    #-------------------------------------------------------------
    
    for (len in preseason.lengths) {
      if ((MidGreenup_mean_DOY - len) >= 1) {
        idx <- (MidGreenup_mean_DOY - len):MidGreenup_mean_DOY
        factors.sub[[paste0("Tmean.PS.MG", len)]] <- mean(daily_vals$Tmean[idx], na.rm = TRUE)
        factors.sub[[paste0("Tmax.PS.MG", len)]]  <- mean(daily_vals$Tmax[idx],  na.rm = TRUE)
        factors.sub[[paste0("Tmin.PS.MG", len)]]  <- mean(daily_vals$Tmin[idx],  na.rm = TRUE)
        factors.sub[[paste0("SWrad.PS.MG", len)]] <- mean(daily_vals$SWrad[idx], na.rm = TRUE)
        factors.sub[[paste0("RH.PS.MG", len)]]    <- mean(daily_vals$RH[idx],    na.rm = TRUE)
      }
    }
    
    
    ##############################################################################################################################################
    
    
    return(factors.sub)
    
  }, error = function(e) {
    saveRDS(list(pheno = pheno.sub, error = e$message),
            file = file.path(error_log.dir, paste0("error_", timeseries_years, ".rds")))
    return(NULL)
  })
}
```
</details>

<details><summary>Run loop and safe data</summary>
```{r}
#-------------------------------------------------------------
# Run in parallel
#-------------------------------------------------------------

outputlist <- pbmclapply(timeseries_year, function(y) parallelCalc(y, DataList), mc.cores = 12, mc.preschedule = T)
climate.factors.df <- rbindlist(outputlist, fill = TRUE)


#-------------------------------------------------------------
# Save results
#-------------------------------------------------------------

write.csv(climate.factors.df, file.path(MODIS_drivers.dir, "MODIS_drivers_spring_data.csv"))
```
</details>

<br>

## 3. Reproducibility
<details><summary>show code</summary>
```{r}
## date time
Sys.time()

## session info
sessionInfo()
```