---
title: "Spring phenology tier2 project"
subtitle: "Bayesian model figures (MODIS data)"
author: "Constantin Zohner"
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

format:
  html:
    toc: false
    theme: default
    highlight-style: haddock
    df-print: paged
    embed-resources: true

execute:
  #warning: false
  message: false
  echo: true
  output: true

always_allow_html: true
editor: visual
editor_options:
  chunk_output_type: inline
---

<br>

### Models

-   Full_GU: Multivariate Greenup model
-   Full_MG: Multivariate MidGreenup model
-   Monthly_GU: Greenup models including monthly mean temperatures
-   Monthly_MG: MidGreenup models including monthly mean temperatures

### Variable names

#### Preseason

##### (start and end dates vary across time series; start date = fitted; end date = time series-level mean leaf-out)

-   Tmean/max: Preseason mean/maximum daily air temperature
-   SWrad: Preseason short-wave radiation
-   RH: Preseason relative humidity

#### Fixed periods

-   Tmean.GS: Growing-season air temperature (May 1 to September 30 of preceding year)
-   Tmean.EQ1.SO1: Mean daily air temperature (March equinox to summer solstice of preceding year)
-   Tmean.SO1.EQ2: Mean daily air temperature (Summer solstice to September equinox of preceding year)
-   Tmean.EQ2.SO2: Mean daily air temperature (September equinox to winter solstice of preceding year)
-   Tmean.8-17: Mean monthly air temperature from August of previous year to May of current year

#### Phenology

-   (Mid)Greenup_prev: Leaf-out date of preceding year
-   MidGreendown_prev: Leaf senescence date of preceding year

#### Late frosts in preceding year

-   LSF2 / LSF4: Binary indicators of late spring frost events occurring after leaf-out in the previous year:
    -   LSF2: Set to 1 if temperature dropped below –2°C
    -   LSF4: Set to 1 if temperature dropped below –4°C

<br>

# 1. Setup

<details>

<summary>Packages</summary>

```{r, message=F}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggpattern)
library(brms)
library(posterior)
library(rstan)
library(pracma)
```

</details>

<details>

<summary>Functions</summary>

```{r, message=F}
#-----------------------------------------------------------
# Extract key convergence diagnostics from a brmsfit model
#-----------------------------------------------------------

extract_diagnostics <- function(model, ForestType_Continent) {
  draws <- as_draws_df(model)
  diag_summary <- summarise_draws(draws)
  data.frame(
    ForestType_Continent = ForestType_Continent,
    max_rhat = max(diag_summary$rhat, na.rm = TRUE),               # Rhat convergence statistic
    min_ess_bulk = min(diag_summary$ess_bulk, na.rm = TRUE),       # Minimum effective sample size (bulk)
    min_ess_tail = min(diag_summary$ess_tail, na.rm = TRUE),       # Minimum effective sample size (tail)
    n_divergent = sum(model$diagnostic_summary$divergent__),       # Total number of divergent transitions
    treedepth_exceeded = sum(model$diagnostic_summary$treedepth__) # Total number of transitions exceeding tree depth
  )
}


#-----------------------------------------------------------
# Interpolation function
#-----------------------------------------------------------

lin_interp = function(x, y, length.out=100) {
  approx(x, y, xout=seq(min(x), max(x), length.out=length.out))$y
}
```

</details>

<details>

<summary>Plot theme</summary>

```{r, message=F}
plotTheme1 = theme(legend.position   = "none",
                   legend.background = element_rect(fill=NA, linewidth=0.5, linetype="solid"),
                   legend.text       = element_text(color="black"),
                   panel.grid.major  = element_line(color = "grey90"),
                   panel.background  = element_blank(),
                   panel.border      = element_rect(colour = "black", fill=NA),
                   axis.line         = element_line(color = "black"),
                   axis.text         = element_text(colour = "black"),
                   strip.background  = element_rect(fill=NA),
                   strip.text        = element_text(colour = 'black'),
                   plot.title        = element_text(face="bold"))
```

</details>

<details>

<summary>Directories</summary>

```{r}
#-------------------------------------------------
# Define the base directory
#-------------------------------------------------

base.dir = "/Users/goalie/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/MODIS_Analysis/Analysis_output"


#-------------------------------------------------
# Build specific directories from the base
#-------------------------------------------------

# Model results
model.dir = file.path(base.dir, "ModelResults")

# Figure output
output.dir = file.path(base.dir, "ModelFigures")
```

</details>

<details>

<summary>Data</summary>

```{r}
#----------------------------------------------------
# Full: Multivariate model
#----------------------------------------------------

model.full.GU    = readRDS(file.path(model.dir, "brms_full_GU_models.rds")) # Greenup
model.full.MG    = readRDS(file.path(model.dir, "brms_full_MG_models.rds")) # MidGreenup

#----------------------------------------------------
# Full: Multivariate model alternative (student distribution + response centered but not scaled)
#----------------------------------------------------

model.fullSC.GU    = readRDS(file.path(model.dir, "brms_full_GU_models_studentCenter.rds")) # Greenup
model.fullSC.MG    = readRDS(file.path(model.dir, "brms_full_MG_models_studentCenter.rds")) # MidGreenup

#----------------------------------------------------
# Monthly: Models including monthly mean temperatures
#----------------------------------------------------

model.monthly.GU = readRDS(file.path(model.dir, "brms_monthly_GU_models.rds")) # Greenup
model.monthly.MG = readRDS(file.path(model.dir, "brms_monthly_MG_models.rds")) # MidGreenup
```

</details>

<br>

# 2. Model diagnostics

<details>

<summary>Full GU model</summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.full.GU, function(ForestType_Continent_list, ForestType_Continent) {
    extract_diagnostics(ForestType_Continent_list, ForestType_Continent = ForestType_Continent)
  }) %>%
  # Tidy and reorder variables
  mutate(ForestType_Continent = factor(ForestType_Continent, levels = rev(sort(unique(ForestType_Continent)))))  # alphabetical order


#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Calculate summary statistics within each group
  summarise(
    prop_rhat_ok = mean(max_rhat < 1.01),        # Proportion of models with Rhat < 1.02 (indicates good convergence)
    prop_ess_bulk_ok = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 200 (good sampling in the posterior bulk)
    prop_ess_tail_ok = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 200 (good sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),        # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded), # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = max_rhat)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  labs(
    x = "",
    y = "Maximum Rhat (Model convergence)",
    color = "Model"
  ) +
  coord_flip() +
  plotTheme1+
  theme(legend.position = "top")

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = min_ess_bulk)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "", y = "Minimum Bulk ESS") +
  coord_flip() +
  plotTheme1

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = min_ess_tail)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "", y = "Minimum Tail ESS") +
  coord_flip() +
  plotTheme1 

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_full_GU_model.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_full_GU_model.plot, file="diagnostics_full_GU_model.pdf", path=output.dir,
       width=6, height=6)

diagnostics_full_GU_model.plot 
```

</details>

<details>

<summary>Full MG model</summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.full.MG, function(ForestType_Continent_list, ForestType_Continent) {
    extract_diagnostics(ForestType_Continent_list, ForestType_Continent = ForestType_Continent)
  }) %>%
  # Tidy and reorder variables
  mutate(ForestType_Continent = factor(ForestType_Continent, levels = rev(sort(unique(ForestType_Continent)))))  # alphabetical order


#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Calculate summary statistics
  summarise(
    prop_rhat_ok = mean(max_rhat < 1.01),        # Proportion of models with Rhat < 1.01 (indicates safe convergence)
    prop_ess_bulk_ok = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 400 (ideal sampling in the posterior bulk)
    prop_ess_tail_ok = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 400 (ideal sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),        # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded), # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = max_rhat)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  labs(
    x = "",
    y = "Maximum Rhat (Model convergence)",
    color = "Model"
  ) +
  coord_flip() +
  plotTheme1+
  theme(legend.position = "top")

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = min_ess_bulk)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "", y = "Minimum Bulk ESS") +
  coord_flip() +
  plotTheme1

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = min_ess_tail)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "", y = "Minimum Tail ESS") +
  coord_flip() +
  plotTheme1

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_full_MG_model.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_full_MG_model.plot , file="diagnostics_full_MG_model.pdf", path=output.dir,
       width=6, height=6)

diagnostics_full_MG_model.plot 
```

</details>

<details>

<summary>Full GU studentCenter model</summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.fullSC.GU, function(ForestType_Continent_list, ForestType_Continent) {
    extract_diagnostics(ForestType_Continent_list, ForestType_Continent = ForestType_Continent)
  }) %>%
  # Tidy and reorder variables
  mutate(ForestType_Continent = factor(ForestType_Continent, levels = rev(sort(unique(ForestType_Continent)))))  # alphabetical order


#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Calculate summary statistics within each group
  summarise(
    prop_rhat_ok = mean(max_rhat < 1.01),        # Proportion of models with Rhat < 1.02 (indicates good convergence)
    prop_ess_bulk_ok = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 200 (good sampling in the posterior bulk)
    prop_ess_tail_ok = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 200 (good sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),        # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded), # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = max_rhat)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  labs(
    x = "",
    y = "Maximum Rhat (Model convergence)",
    color = "Model"
  ) +
  coord_flip() +
  plotTheme1+
  theme(legend.position = "top")

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = min_ess_bulk)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "", y = "Minimum Bulk ESS") +
  coord_flip() +
  plotTheme1

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = min_ess_tail)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "", y = "Minimum Tail ESS") +
  coord_flip() +
  plotTheme1 

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_full_GU_SC_model.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_full_GU_SC_model.plot, file="diagnostics_full_GU_studentCenter_model.pdf", path=output.dir,
       width=6, height=6)

diagnostics_full_GU_SC_model.plot 
```

</details>

<details>

<summary>Full MG studentCenter model</summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.fullSC.MG, function(ForestType_Continent_list, ForestType_Continent) {
    extract_diagnostics(ForestType_Continent_list, ForestType_Continent = ForestType_Continent)
  }) %>%
  # Tidy and reorder variables
  mutate(ForestType_Continent = factor(ForestType_Continent, levels = rev(sort(unique(ForestType_Continent)))))  # alphabetical order


#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Calculate summary statistics
  summarise(
    prop_rhat_ok = mean(max_rhat < 1.01),        # Proportion of models with Rhat < 1.01 (indicates safe convergence)
    prop_ess_bulk_ok = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 400 (ideal sampling in the posterior bulk)
    prop_ess_tail_ok = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 400 (ideal sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),        # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded), # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = max_rhat)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  labs(
    x = "",
    y = "Maximum Rhat (Model convergence)",
    color = "Model"
  ) +
  coord_flip() +
  plotTheme1+
  theme(legend.position = "top")

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = min_ess_bulk)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "", y = "Minimum Bulk ESS") +
  coord_flip() +
  plotTheme1

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = min_ess_tail)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "", y = "Minimum Tail ESS") +
  coord_flip() +
  plotTheme1

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_full_MG_SC_model.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_full_MG_SC_model.plot , file="diagnostics_full_MG_studentCenter_model.pdf", path=output.dir,
       width=6, height=6)

diagnostics_full_MG_SC_model.plot 
```

</details>

<details>

<summary>Monthly GU model</summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.monthly.GU, function(ForestType_Continent_list, ForestType_Continent) {
    extract_diagnostics(ForestType_Continent_list, ForestType_Continent = ForestType_Continent)
  }) %>%
  # Tidy and reorder variables
  mutate(ForestType_Continent = factor(ForestType_Continent, levels = rev(sort(unique(ForestType_Continent)))))  # alphabetical order


#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Calculate summary statistics
  summarise(
    prop_rhat_ok = mean(max_rhat < 1.01),        # Proportion of models with Rhat < 1.01 (indicates safe convergence)
    prop_ess_bulk_ok = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 400 (ideal sampling in the posterior bulk)
    prop_ess_tail_ok = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 400 (ideal sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),        # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded), # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = max_rhat)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  labs(
    x = "",
    y = "Maximum Rhat (Model convergence)",
    color = "Model"
  ) +
  coord_flip() +
  plotTheme1+
  theme(legend.position = "top")

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = min_ess_bulk)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "", y = "Minimum Bulk ESS") +
  coord_flip() +
  plotTheme1 +
  theme(axis.text.y = element_text(face = "italic"))

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = min_ess_tail)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "", y = "Minimum Tail ESS") +
  coord_flip() +
  plotTheme1

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_monthly_GU_model.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_monthly_GU_model.plot, file="diagnostics_monthly_GU_model.pdf", path=output.dir,
       width=6, height=6)

diagnostics_monthly_GU_model.plot
```

</details>

<details>

<summary>Monthly MG model</summary>

```{r}
##################################################################
## Prepare Diagnostics — Rhat, ESS bulk, ESS tail across models ##
##################################################################



#-----------------------------------------------------------
# 1. Flatten model diagnostics into one data frame
#-----------------------------------------------------------

diagnostics.df <- imap_dfr(model.monthly.MG, function(ForestType_Continent_list, ForestType_Continent) {
    extract_diagnostics(ForestType_Continent_list, ForestType_Continent = ForestType_Continent)
  }) %>%
  # Tidy and reorder variables
  mutate(ForestType_Continent = factor(ForestType_Continent, levels = rev(sort(unique(ForestType_Continent)))))  # alphabetical order


#-----------------------------------------------------------
# 2. Summarize convergence diagnostics across models
#-----------------------------------------------------------

summary_stats <- diagnostics.df %>%
  # Calculate summary statistics
  summarise(
    prop_rhat_ok = mean(max_rhat < 1.01),        # Proportion of models with Rhat < 1.01 (indicates safe convergence)
    prop_ess_bulk_ok = mean(min_ess_bulk > 400), # Proportion of models with bulk ESS > 400 (ideal sampling in the posterior bulk)
    prop_ess_tail_ok = mean(min_ess_tail > 400), # Proportion of models with tail ESS > 400 (ideal sampling in the posterior tails)
    n_divergent_total = sum(n_divergent),        # Total number of divergent transitions across all models in the group
    n_treedepth_total = sum(treedepth_exceeded), # Total number of transitions that exceeded the maximum tree depth
    .groups = "drop" # Remove grouping structure in the resulting data frame
  )

# Display the summary statistics table
summary_stats



#################################################################
## Visual Diagnostics — Rhat, ESS bulk, ESS tail across models ##
#################################################################



#-----------------------------------------------------------
# Plot A: Maximum Rhat (Convergence diagnostic)
# - Rhat < 1.01: safe convergence
# - Rhat > 1.05: problematic convergence
#-----------------------------------------------------------

Rhat = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = max_rhat)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 1.01, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 1.05, linetype = "solid", color = "red") +
  labs(
    x = "",
    y = "Maximum Rhat (Model convergence)",
    color = "Model"
  ) +
  coord_flip() +
  plotTheme1+
  theme(legend.position = "top")

#-----------------------------------------------------------
# Plot B: Bulk ESS (Effective sample size for bulk of posterior)
# - ESS < 400: potentially problematic
# - ESS < 100: critical
#-----------------------------------------------------------

Bulk = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = min_ess_bulk)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # absolute minimum threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "ForestType_Continent", y = "Minimum Bulk ESS") +
  coord_flip() +
  plotTheme1

#-----------------------------------------------------------
# Plot C: Tail ESS (Effective sample size for distribution tails)
# - Same thresholds as Bulk ESS
#-----------------------------------------------------------

Tail = ggplot(diagnostics.df, aes(x = ForestType_Continent, y = min_ess_tail)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 100, linetype = "solid", color = "red") +    # critical threshold
  geom_hline(yintercept = 400, linetype = "dashed", color = "black") + # ideal minimum
  labs(x = "ForestType_Continent", y = "Minimum Tail ESS") +
  coord_flip() +
  plotTheme1

#-----------------------------------------------------------
# Merge and safe plots
#-----------------------------------------------------------

#define plot layouts
layout <- "
A
B
C"

#Merge plots
diagnostics_monthly_MG_model.plot = Rhat + Bulk + Tail +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(diagnostics_monthly_MG_model.plot, file="diagnostics_monthly_MG_model.pdf", path=output.dir,
       width=6, height=6)

diagnostics_monthly_MG_model.plot
```

</details>

<br>

# 3. Model results - Extract posterior stats

<details>

<summary>Full GU model</summary>

```{r, warning=FALSE}
#-----------------------------------------------------------
# 1. Loop through nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.full.GU, function(model_obj, ForestType_Continent_name) {
    if (inherits(model_obj, "brmsfit")) {
      broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
        mutate(
          ForestType_Continent = ForestType_Continent_name,
          .before = term
        )
    } else {
      NULL  # skip if not a brms model
    }
  }) %>% 
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
    ForestType_Continent = factor(ForestType_Continent, 
                                  levels = rev(sort(unique(ForestType_Continent)))), # alphabetical order
    term = plyr::revalue(term, warn_missing=F,
                         c("Greenup_prev"      = "Greenup",
                           "LSF2"              = "Late frost",
                           "Tmean.EQ1.SO1"     = "Temp Q2",
                           "Tmean.SO1.EQ2"     = "Temp Q3",
                           "MidGreendown_prev" = "Greendown",
                           "Tmean.EQ2.SO2"     = "Temp Q4",
                           "SWrad.GU"          = "Radiation",
                           "Tmean.GU"          = "Temperature",
                           "RH.GU"             = "Humidity")))


#-----------------------------------------------------------
# 2. Summarize data and create plotting helpers
#-----------------------------------------------------------

# Define consistent term order
term.order <- c("Greenup","Late frost",
                "Temp Q2","Temp Q3","Greendown","Temp Q4",
                "Humidity","Radiation","Temperature")

# Summarize data
summary_model.df <- model.df %>%
  group_by(term) %>%
  summarise(
    mean_estimate = mean(estimate, na.rm = TRUE),
    # 100% of the individual forest types' 95% CIs fall within this range
    lower = min(conf.low, na.rm = TRUE),
    upper = max(conf.high,na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(term = factor(term, levels = term.order))

# get the number of terms
terms <- unique(model.df$term)
n_terms <- length(terms)

# Create numeric positions halfway between rows (used for spacing lines between terms)
grid_lines <- seq(1.5, n_terms - 0.5, by = 1)

# Define dodge & jitter
dodge <- position_dodge(width = 0.6)


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_full_GU_model.plot = ggplot(model.df, aes(x = term)) +
  
  # Horizontal reference line
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  
  # Summary error bars with dashed for early, solid for late
  geom_errorbar(data = summary_model.df,
                aes(x = term, y = mean_estimate, ymin = lower, ymax = upper),
                position = dodge, width = 0.3, linewidth = 0.5,
                inherit.aes = FALSE, color = "black") +
  
  # Forest type-level error bars
  geom_errorbar(aes(y = estimate, ymin = conf.low, ymax = conf.high, color = ForestType_Continent),
                position = dodge, width = 0.5, linewidth = 0.4) +
  
  # Forest type-level points
  geom_point(aes(y = estimate, color = ForestType_Continent),
             position = dodge, size = .8) +
  
  scale_color_manual(values = c("#e31a1c", "#33a02c", "#1f78b4")) +
  
  # Flip axes and label
  coord_flip(ylim = c(-0.66, 0.66)) +
  xlab("") + ylab("Standardized effect (posterior mean ± 95% CI)") +
  
  # Theme
  plotTheme1 +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 0),
    axis.text.y = element_text(face = "plain"),
    panel.grid.major = element_blank(),
  ) +
  
  # add horizontal lines at those positions
  geom_vline(xintercept = grid_lines, linetype = "solid", color = "grey80")

#save plot as .pdf
ggsave(results_full_GU_model.plot, file="results_full_GU_model.pdf", path=output.dir,
       width=7, height=3.5)

results_full_GU_model.plot
```

</details>

<details>

<summary>Full MG model</summary>

```{r, warning=FALSE}
#-----------------------------------------------------------
# 1. Loop through nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.full.MG, function(model_obj, ForestType_Continent_name) {
    if (inherits(model_obj, "brmsfit")) {
      broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
        mutate(
          ForestType_Continent = ForestType_Continent_name,
          .before = term
        )
    } else {
      NULL  # skip if not a brms model
    }
  }) %>% 
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
    ForestType_Continent = factor(ForestType_Continent, 
                                  levels = rev(sort(unique(ForestType_Continent)))), # alphabetical order
    term = plyr::revalue(term, warn_missing=F,
                         c("MidGreenup_prev"   = "Mid-Greenup",
                           "LSF2"              = "Late frost",
                           "Tmean.EQ1.SO1"     = "Temp Q2",
                           "Tmean.SO1.EQ2"     = "Temp Q3",
                           "MidGreendown_prev" = "Greendown",
                           "Tmean.EQ2.SO2"     = "Temp Q4",
                           "SWrad.MG"          = "Radiation",
                           "Tmean.MG"          = "Temperature",
                           "RH.MG"             = "Humidity")))


#-----------------------------------------------------------
# 2. Summarize data and create plotting helpers
#-----------------------------------------------------------

# Define consistent term order
term.order <- c("Mid-Greenup","Late frost",
                "Temp Q2","Temp Q3","Greendown","Temp Q4",
                "Humidity","Radiation","Temperature")

# Summarize data
summary_model.df <- model.df %>%
  group_by(term) %>%
  summarise(
    mean_estimate = mean(estimate, na.rm = TRUE),
    # 100% of the individual forest types' 95% CIs fall within this range
    lower = min(conf.low, na.rm = TRUE),
    upper = max(conf.high,na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(term = factor(term, levels = term.order))

# get the number of terms
terms <- unique(model.df$term)
n_terms <- length(terms)

# Create numeric positions halfway between rows (used for spacing lines between terms)
grid_lines <- seq(1.5, n_terms - 0.5, by = 1)

# Define dodge & jitter
dodge <- position_dodge(width = 0.6)


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_full_MG_model.plot = ggplot(model.df, aes(x = term)) +
  
  # Horizontal reference line
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  
  # Summary error bars with dashed for early, solid for late
  geom_errorbar(data = summary_model.df,
                aes(x = term, y = mean_estimate, ymin = lower, ymax = upper),
                position = dodge, width = 0.3, linewidth = 0.5,
                inherit.aes = FALSE, color = "black") +
  
  # Forest type-level error bars
  geom_errorbar(aes(y = estimate, ymin = conf.low, ymax = conf.high, color = ForestType_Continent),
                position = dodge, width = 0.5, linewidth = 0.4) +
  
  # Forest type-level points
  geom_point(aes(y = estimate, color = ForestType_Continent),
             position = dodge, size = .8) +
  
  scale_color_manual(values = c("#e31a1c","#33a02c","#1f78b4")) +
  
  # Flip axes and label
  coord_flip(ylim = c(-0.66, 0.66)) +
  xlab("") + ylab("Standardized effect (posterior mean ± 95% CI)") +
  
  # Theme
  plotTheme1 +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 0),
    axis.text.y = element_text(face = "plain"),
    panel.grid.major = element_blank(),
  ) +
  
  # add horizontal lines at those positions
  geom_vline(xintercept = grid_lines, linetype = "solid", color = "grey80")

#save plot as .pdf
ggsave(results_full_MG_model.plot, file="results_full_MG_model.pdf", path=output.dir,
       width=7, height=3.5)

results_full_MG_model.plot
```

</details>

<details>

<summary>Full GU studentCenter model</summary>

```{r, warning=FALSE}
#-----------------------------------------------------------
# 1. Loop through nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.fullSC.GU, function(model_obj, ForestType_Continent_name) {
    if (inherits(model_obj, "brmsfit")) {
      broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
        mutate(
          ForestType_Continent = ForestType_Continent_name,
          .before = term
        )
    } else {
      NULL  # skip if not a brms model
    }
  }) %>% 
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
    ForestType_Continent = factor(ForestType_Continent, 
                                  levels = rev(sort(unique(ForestType_Continent)))), # alphabetical order
    term = plyr::revalue(term, warn_missing=F,
                         c("Greenup_prev_z"      = "Greenup",
                           "LSF2"                = "Late frost",
                           "Tmean.EQ1.SO1_z"     = "Temp Q2",
                           "Tmean.SO1.EQ2_z"     = "Temp Q3",
                           "MidGreendown_prev_z" = "Greendown",
                           "Tmean.EQ2.SO2_z"     = "Temp Q4",
                           "SWrad.GU_z"          = "Radiation",
                           "Tmean.GU_z"          = "Temperature",
                           "RH.GU_z"             = "Humidity")))


#-----------------------------------------------------------
# 2. Summarize data and create plotting helpers
#-----------------------------------------------------------

# Define consistent term order
term.order <- c("Greenup","Late frost",
                "Temp Q2","Temp Q3","Greendown","Temp Q4",
                "Humidity","Radiation","Temperature")

# Summarize data
summary_model.df <- model.df %>%
  group_by(term) %>%
  summarise(
    mean_estimate = mean(estimate, na.rm = TRUE),
    # 100% of the individual forest types' 95% CIs fall within this range
    lower = min(conf.low, na.rm = TRUE),
    upper = max(conf.high,na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(term = factor(term, levels = term.order))

# get the number of terms
terms <- unique(model.df$term)
n_terms <- length(terms)

# Create numeric positions halfway between rows (used for spacing lines between terms)
grid_lines <- seq(1.5, n_terms - 0.5, by = 1)

# Define dodge & jitter
dodge <- position_dodge(width = 0.6)


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_full_GU_SC_model.plot = ggplot(model.df, aes(x = term)) +
  
  # Horizontal reference line
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  
  # Summary error bars with dashed for early, solid for late
  geom_errorbar(data = summary_model.df,
                aes(x = term, y = mean_estimate, ymin = lower, ymax = upper),
                position = dodge, width = 0.3, linewidth = 0.5,
                inherit.aes = FALSE, color = "black") +
  
  # Forest type-level error bars
  geom_errorbar(aes(y = estimate, ymin = conf.low, ymax = conf.high, color = ForestType_Continent),
                position = dodge, width = 0.5, linewidth = 0.4) +
  
  # Forest type-level points
  geom_point(aes(y = estimate, color = ForestType_Continent),
             position = dodge, size = .8) +
  
  scale_color_manual(values = c("#e31a1c", "#33a02c", "#1f78b4")) +
  
  # Flip axes and label
  coord_flip(ylim = c(-5, 5)) +
  xlab("") + ylab("Standardized effect (posterior mean ± 95% CI)") +
  
  # Theme
  plotTheme1 +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 0),
    axis.text.y = element_text(face = "plain"),
    panel.grid.major = element_blank(),
  ) +
  
  # add horizontal lines at those positions
  geom_vline(xintercept = grid_lines, linetype = "solid", color = "grey80")

#save plot as .pdf
ggsave(results_full_GU_SC_model.plot, file="results_full_GU_studentCenter_model.pdf", path=output.dir,
       width=7, height=3.5)

results_full_GU_SC_model.plot
```

</details>

<details>

<summary>Full MG studentCenter model</summary>

```{r, warning=FALSE}
#-----------------------------------------------------------
# 1. Loop through nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.fullSC.MG, function(model_obj, ForestType_Continent_name) {
    if (inherits(model_obj, "brmsfit")) {
      broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
        mutate(
          ForestType_Continent = ForestType_Continent_name,
          .before = term
        )
    } else {
      NULL  # skip if not a brms model
    }
  }) %>% 
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
    ForestType_Continent = factor(ForestType_Continent, 
                                  levels = rev(sort(unique(ForestType_Continent)))), # alphabetical order
    term = plyr::revalue(term, warn_missing=F,
                         c("MidGreenup_prev_z"   = "Mid-Greenup",
                           "LSF2"                = "Late frost",
                           "Tmean.EQ1.SO1_z"     = "Temp Q2",
                           "Tmean.SO1.EQ2_z"     = "Temp Q3",
                           "MidGreendown_prev_z" = "Greendown",
                           "Tmean.EQ2.SO2_z"     = "Temp Q4",
                           "SWrad.MG_z"          = "Radiation",
                           "Tmean.MG_z"          = "Temperature",
                           "RH.MG_z"             = "Humidity")))


#-----------------------------------------------------------
# 2. Summarize data and create plotting helpers
#-----------------------------------------------------------

# Define consistent term order
term.order <- c("Mid-Greenup","Late frost",
                "Temp Q2","Temp Q3","Greendown","Temp Q4",
                "Humidity","Radiation","Temperature")

# Summarize data
summary_model.df <- model.df %>%
  group_by(term) %>%
  summarise(
    mean_estimate = mean(estimate, na.rm = TRUE),
    # 100% of the individual forest types' 95% CIs fall within this range
    lower = min(conf.low, na.rm = TRUE),
    upper = max(conf.high,na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(term = factor(term, levels = term.order))

# get the number of terms
terms <- unique(model.df$term)
n_terms <- length(terms)

# Create numeric positions halfway between rows (used for spacing lines between terms)
grid_lines <- seq(1.5, n_terms - 0.5, by = 1)

# Define dodge & jitter
dodge <- position_dodge(width = 0.6)


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_full_MG_SC_model.plot = ggplot(model.df, aes(x = term)) +
  
  # Horizontal reference line
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  
  # Summary error bars with dashed for early, solid for late
  geom_errorbar(data = summary_model.df,
                aes(x = term, y = mean_estimate, ymin = lower, ymax = upper),
                position = dodge, width = 0.3, linewidth = 0.5,
                inherit.aes = FALSE, color = "black") +
  
  # Forest type-level error bars
  geom_errorbar(aes(y = estimate, ymin = conf.low, ymax = conf.high, color = ForestType_Continent),
                position = dodge, width = 0.5, linewidth = 0.4) +
  
  # Forest type-level points
  geom_point(aes(y = estimate, color = ForestType_Continent),
             position = dodge, size = .8) +
  
  scale_color_manual(values = c("#e31a1c","#33a02c","#1f78b4")) +
  
  # Flip axes and label
  coord_flip(ylim = c(-5, 5)) +
  xlab("") + ylab("Standardized effect (posterior mean ± 95% CI)") +
  
  # Theme
  plotTheme1 +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 0),
    axis.text.y = element_text(face = "plain"),
    panel.grid.major = element_blank(),
  ) +
  
  # add horizontal lines at those positions
  geom_vline(xintercept = grid_lines, linetype = "solid", color = "grey80")

#save plot as .pdf
ggsave(results_full_MG_SC_model.plot, file="results_full_MG_studentCenter_model.pdf", path=output.dir,
       width=7, height=3.5)

results_full_MG_SC_model.plot
```

</details>

<details>

<summary>Monthly GU model</summary>

```{r, warning=F}
#-----------------------------------------------------------
# 1. Loop through 2-level nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.monthly.GU, function(model_obj, ForestType_Continent_name) {
  if (inherits(model_obj, "brmsfit")) {
    broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
      mutate(
        ForestType_Continent = ForestType_Continent_name,
        .before = term
        )
    } else {
      NULL  # skip if not a brms model
      }
  }) %>%
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
      ForestType_Continent = factor(ForestType_Continent, levels = rev(sort(unique(ForestType_Continent)))), # alphabetical order
      term = str_remove(term, "[.]"),
      term = readr::parse_number(term))


#-----------------------------------------------------------
# 2. Interpolation of monthly estimates 
#-----------------------------------------------------------

#create identifier
ForestType = unique(model.df$ForestType_Continent)

#create interpolation dataframe
df.interp = data.frame()
df.AUC = data.frame()

#loop over variable x equation x species vector
for (type in ForestType){

  #subset table
  df.sub = model.df  %>% 
    filter(ForestType_Continent == type) 
  
  # Interpolate data
  created.interp = lin_interp(df.sub$term, df.sub$term)
  score.interp   = lin_interp(df.sub$term, df.sub$estimate)
  df.interp.sub  = data.frame(created=created.interp, score=score.interp)
  
  # Make a grouping variable for each pos/neg segment
  cat.rle        = rle(df.interp.sub$score < 0)
  df.interp.sub  = df.interp.sub %>%
    mutate(group = rep.int(1:length(cat.rle$lengths),  times=cat.rle$lengths),
           ForestType_Continent = unique(df.sub$ForestType_Continent))
  #rbind sub dataframes 
  df.interp = rbind(df.interp, df.interp.sub)
  
  #get Area under curve (%)
  df.AUC.sub = df.interp.sub %>%
    mutate(positive = ifelse(score<0, 0, score),
           negative = ifelse(score>0, 0, score))%>%
    summarise(sum.pos = trapz(created, positive), 
              sum.neg = abs(trapz(created, negative)))%>%
    mutate(percent.neg = round(sum.neg/(sum.pos+sum.neg)*100),
           percent.pos = round(sum.pos/(sum.pos+sum.neg)*100),
           ForestType_Continent = unique(df.sub$ForestType_Continent))
  
  #rbind sub dataframes 
  df.AUC = rbind(df.AUC, df.AUC.sub)
}


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_monthly_GU_model.plot = ggplot() + 
  
  geom_area(data = df.interp, aes(x = created, y = score, fill=score>0, group=group)) + 
  
  scale_fill_manual(values = c('#F21A00', '#3B9AB2'))+
  
  geom_point(data=model.df, aes(x=term, y=estimate), size=0.2)+
  
  geom_errorbar(data=model.df, aes(x=term, ymin=conf.low, ymax=conf.high), width=.2,
                position=position_dodge(.9)) +
  
  geom_hline(yintercept=0)+
  
  geom_vline(xintercept=12.3, linewidth=2, alpha=0.4)+
  
  geom_text(data = df.AUC, mapping = aes(x = -Inf, y = -Inf, hjust = -0.2, vjust = -1.5,
                                         label = paste0(percent.pos,'% / ',percent.neg, '%')))+
  
  coord_cartesian(xlim=c(8.3,15.7))+
  
  xlab("") + ylab("St (days °C-1)")+
  
  scale_x_continuous(breaks = seq(8,16,by=1),
                     labels = c('Aug','Sep','Oct','Nov','Dec','Jan','Feb','Mar','Apr'))+
  
  plotTheme1 + 
  
  facet_wrap(~ForestType_Continent, ncol=1) +

  theme(
    strip.text = element_blank(),
    strip.background = element_blank() 
    )
  
results_monthly_GU_model.plot
```

</details>

<details>

<summary>Monthly MG model</summary>

```{r, warning=F}
#-----------------------------------------------------------
# 1. Loop through 2-level nested model list
#-----------------------------------------------------------

model.df <- purrr::imap_dfr(model.monthly.MG, function(model_obj, ForestType_Continent_name) {
  if (inherits(model_obj, "brmsfit")) {
    broom.mixed::tidy(model_obj, effects = "fixed", conf.int = TRUE, conf.level = 0.95) %>%
      mutate(
        ForestType_Continent = ForestType_Continent_name,
        .before = term
        )
    } else {
      NULL  # skip if not a brms model
      }
  }) %>%
  # Tidy
  filter(!term == "(Intercept)") %>% 
  dplyr::select(-c(component,effect)) %>% 
  # Tidy, reorder and rename variables
  mutate(
      ForestType_Continent = factor(ForestType_Continent, levels = rev(sort(unique(ForestType_Continent)))), # alphabetical order
      term = str_remove(term, "[.]"),
      term = readr::parse_number(term))


#-----------------------------------------------------------
# 2. Interpolation of monthly estimates 
#-----------------------------------------------------------

#create identifier
ForestType = unique(model.df$ForestType_Continent)

#create interpolation dataframe
df.interp = data.frame()
df.AUC = data.frame()

#loop over variable x equation x species vector
for (type in ForestType){

  #subset table
  df.sub = model.df  %>% 
    filter(ForestType_Continent == type) 
  
  # Interpolate data
  created.interp = lin_interp(df.sub$term, df.sub$term)
  score.interp   = lin_interp(df.sub$term, df.sub$estimate)
  df.interp.sub  = data.frame(created=created.interp, score=score.interp)
  
  # Make a grouping variable for each pos/neg segment
  cat.rle        = rle(df.interp.sub$score < 0)
  df.interp.sub  = df.interp.sub %>%
    mutate(group = rep.int(1:length(cat.rle$lengths),  times=cat.rle$lengths),
           ForestType_Continent = unique(df.sub$ForestType_Continent))
  #rbind sub dataframes 
  df.interp = rbind(df.interp, df.interp.sub)
  
  #get Area under curve (%)
  df.AUC.sub = df.interp.sub %>%
    mutate(positive = ifelse(score<0, 0, score),
           negative = ifelse(score>0, 0, score))%>%
    summarise(sum.pos = trapz(created, positive), 
              sum.neg = abs(trapz(created, negative)))%>%
    mutate(percent.neg = round(sum.neg/(sum.pos+sum.neg)*100),
           percent.pos = round(sum.pos/(sum.pos+sum.neg)*100),
           ForestType_Continent = unique(df.sub$ForestType_Continent))
  
  #rbind sub dataframes 
  df.AUC = rbind(df.AUC, df.AUC.sub)
}


#-----------------------------------------------------------
# 3. Plot data
#-----------------------------------------------------------

results_monthly_MG_model.plot = ggplot() + 
  
  geom_area(data = df.interp, aes(x = created, y = score, fill=score>0, group=group)) + 
  
  scale_fill_manual(values = c('#F21A00', '#3B9AB2'))+
  
  geom_point(data=model.df, aes(x=term, y=estimate), size=0.2)+
  
  geom_errorbar(data=model.df, aes(x=term, ymin=conf.low, ymax=conf.high), width=.2,
                position=position_dodge(.9)) +
  
  geom_hline(yintercept=0)+
  
  geom_vline(xintercept=12.3, linewidth=2, alpha=0.4)+
  
  geom_text(data = df.AUC, mapping = aes(x = -Inf, y = -Inf, hjust = -0.2, vjust = -1.5,
                                         label = paste0(percent.pos,'% / ',percent.neg, '%')))+
  
  coord_cartesian(xlim=c(8.3,16.7))+
  
  xlab("") + ylab("")+
  
  scale_x_continuous(breaks = seq(8,17,by=1),
                     labels = c('Aug','Sep','Oct','Nov','Dec','Jan','Feb','Mar','Apr','May'))+
  
  plotTheme1 + 
  
  facet_wrap(~ForestType_Continent, ncol=1, strip.position = "right")

results_monthly_MG_model.plot


#-----------------------------------------------------------
# 4. Merge and safe Greenup and Mid-Greenup plots
#-----------------------------------------------------------

#define plot layouts
layout <- "AB"

#Merge plots
results_monthly_model.plot = results_monthly_GU_model.plot+ results_monthly_MG_model.plot +
  plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#save plots as .pdf
ggsave(results_monthly_model.plot , file="results_monthly_model_MODIS.pdf", path=output.dir,
       width=6.5, height=6)

results_monthly_model.plot
```

</details>

<br>

# 4. Reproducibility

<details>

<summary>show</summary>

```{r}
# Datetime
Sys.time()

# Session info
sessionInfo()
```

</details>
