---
title: Spring phenology tier2 project
author: Constantin Zohner
date: "last updated 15 March 2025"

subtitle: R code generating the drivers
output:
  html_document:
    highlight: haddock
    toc: false
    use_bookdown: true
    df_print: paged
---

<style>
body
  { counter-reset: source-line 0; }
pre.numberSource code
  { counter-reset: none; }
</style>

<br>

### Driver names
- Tmean: Mean daily temperature
- Tmin: Minimum daily temperature
- Tmax: Maximum daily temperature
- SWrad: Net short-wave radiation
- RH: Relative humidity
- Chill: Chill days
- GDD: Growing-Degree days
- LSF: Late frost (-2°C or -4°C) occurrence in previous year 

### Species names
- Deciduous trees: Aesculus hippocastanum, Alnus glutinosa, Betula pendula, Fagus sylvatica, Fraxinus excelsior, Quercus robur
- Deciduous shrub: Ribes uva-crispa
- Deciduous needleleaf: Larix decidua
- Evergreen needleleaf: Picea abies, Pinus sylvestris

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  class.source = "numberLines lineAnchors"
  )
```

<br><br>

```{r, message=FALSE, warning=FALSE, attr.source='.numberLines'}
#####################
# Required packages #
#####################



require(data.table)
require(dplyr)
require(pbmcapply)
require(zoo)
require(ggplot2)



##############################################################################################################################################
##############################################################################################################################################



################
# Define Paths #
################



# 1. Input
##########


#Climate
EOBS_path  = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_input/EOBS_data"

#Photoperiod
photo_path  = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_input/Photoperiod"

#Phenology
PEPcsv_path = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/PEP_data"


# 2. Output
###########

PEP_drivers_path  = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_input/PEP_drivers_final_spring/Individual_files"
PEP_drivers_path2 = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_input/PEP_drivers_final_spring/Merged_file"
PEP_drivers_path3 = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_input/PEP_drivers_final_spring/Missing_observations"



##############################################################################################################################################
##############################################################################################################################################



#################
## Import data ##
#################



## Phenology data
#################

PEP.df <- fread(paste(PEPcsv_path, "PEP725_data_clean.csv", sep="/")) 


## Photoperiod
##############

photo.df = fread(paste(photo_path, "Photoperiod.csv", sep="/"))


## Import daily climatic datasets from EOB-S
############################################

#define climate variables
list.files(EOBS_path)
vn <- c("tg_ens_mean_0.1deg_reg_v30.0e.csv", # daily mean temperature 
        "tn_ens_mean_0.1deg_reg_v30.0e.csv", # daily minimum temperature
        "tx_ens_mean_0.1deg_reg_v30.0e.csv", # daily maximum temperature
        "rr_ens_mean_0.1deg_reg_v30.0e.csv", # daily precipitation sum 
        "hu_ens_mean_0.1deg_reg_v30.0e.csv", # daily mean relative humidity 
        "qq_ens_mean_0.1deg_reg_v29.0e.csv") # global radiation
# not used: 
# "pp_ens_mean_0.1deg_reg_v30.0e.csv" daily mean sea level pressure 

#create empty list
DataList <- replicate(length(vn),data.frame())
#loop through climate variables
for(i in 1:length(vn)) {
  #read data
  data = fread(paste0(EOBS_path, "/", vn[i]))
  #add table to list
  DataList[[i]] = data }
#add names to list
names(DataList)=vn



##############################################################################################################################################
##############################################################################################################################################



######################
## Helper functions ##
######################



################################
# Chilling and GDD calculation #
################################


# Chilling units
Chill.fun = function(Temp, threshold_low, threshold_up){
  if(between(Temp,threshold_low, threshold_up)) {CU=1} else {CU=0}
  return(CU)
}
#Example
temperature = c(-20:25)
response=sapply(temperature, Chill.fun, threshold_low=-10, threshold_up=10)
plot(response~temperature, type="l", xlab="Temperature (Celsius)", ylab="Chilling response", col="blue")

# Forcing units
GDD.fun = function(Temp, threshold_low){
  if(Temp >= threshold_low) {GDD=Temp-threshold_low} else {GDD=0}
  return(GDD)
}
#Example
temperature = c(-10:25)
response=sapply(temperature, GDD.fun, threshold_low=5)
plot(response~temperature, type="l", xlab="Temperature (Celsius)", ylab="Forcing response", col="red")


#####################
# Photoperiod model #
#####################


# Function for photoperiod sensitivity with different k and p values
photoperiod_response <- function(DL, k, p) {
  DL_adjusted <- pmax(DL - 6, 0)  # Shift DL so that it starts at 6 hours instead of 0
  f_DL <- (log(1 + k * DL_adjusted) / log(1 + k * (20 - 6)))^p  # Normalize to DL = 20
  f_DL <- pmin(f_DL, 1)  # Cap at 1 from DL = 20 to 24
  return(f_DL)
}

# Sample data
DL_values <- seq(0, 24, by=1)  # Photoperiod values from 8h to 24h

# Compute GDD factor for all cases
DL_factor <- photoperiod_response(DL_values, 1, 1)

# Combine results into a dataframe
df <- data.frame(DayLength = DL_values,
                 DL_factor = DL_factor)

# Plot results
ggplot(df, aes(x = DayLength)) +
  geom_line(aes(y = DL_factor)) +
  labs(x = "Photoperiod (hours)", y = "Photoperiod factor",
       title = "Effect of Photoperiod on GDD Accumulation") +
  theme_bw()


##################
# Chilling model #
##################


# Function for chilling dependency scaling factor
chilling_response <- function(Chilling, requirement) {
  factor <- pmin(Chilling / requirement, 1)  # chilling dependency reaches 1 at X chill days
  return(factor)
}
# Chilling = actual chilling unit 
# requirement = total chilling requirement (in days), i.e., after how many chill days is chilling completely fulfilled?

# Sample data
Chill_values <- seq(0, 180, by=1)  # Photoperiod values from 8h to 24h

# Compute GDD factor for all cases
Chill_factor_high <- chilling_response(Chill_values, 150)
Chill_factor_low <- chilling_response(Chill_values, 75)

# Combine results into a dataframe
df <- data.frame(Chilling          = Chill_values,
                 Chill_factor_high = Chill_factor_high,
                 Chill_factor_low = Chill_factor_low)

# Plot results
ggplot(df, aes(x = Chilling)) +
  geom_line(aes(y = Chill_factor_high), color="blue", linetype="dashed") +
  geom_line(aes(y = Chill_factor_low), color="darkblue", linetype="dotted") +
  labs(x = "Chilling days", y = "Chilling factor",
       title = "Effect of Chilling on GDD Accumulation") +
  theme_bw()


#####################################
# Function to compute adjusted GDDs #
#####################################


adjust_gdd <- function(GDD, Chilling, requirement=150, include_photo = F, DL = NULL, k = 1, p = 1) {
  f_Chill <- chilling_response(Chilling, requirement)
  
  if (include_photo) {
    if (is.null(DL)) stop("DL must be provided when include_photo is TRUE")
    f_Photo <- photoperiod_response(DL, k, p)
    GDD_adjusted <- GDD * f_Chill * f_Photo
  } else {
    GDD_adjusted <- GDD * f_Chill
  }
  
  return(GDD_adjusted)
}


#################################
# Replace NA with neighbor mean #
#################################


replace_na_with_neighbor_mean <- function(x) {
  if (!is.numeric(x)) return(x)  # Ensure only numeric columns are processed
  
  na_indices <- which(is.na(x))  # Identify NA indices
  known_indices <- which(!is.na(x))  # Identify non-NA values
  
  if (length(known_indices) == 0) {
    return(x)  # If all values are NA, return as is
  }
  
  for (i in na_indices) {
    # Find nearest non-NA before and after
    before_idx <- ifelse(any(known_indices < i), max(known_indices[known_indices < i]), NA)
    after_idx <- ifelse(any(known_indices > i), min(known_indices[known_indices > i]), NA)
    
    if (!is.na(before_idx) & !is.na(after_idx)) {
      # Compute mean of the surrounding values
      x[i] <- mean(c(x[before_idx], x[after_idx]), na.rm = TRUE)
    } else if (!is.na(before_idx)) {
      x[i] <- x[before_idx]  # If NA is at the end, take the previous value
    } else if (!is.na(after_idx)) {
      x[i] <- x[after_idx]  # If NA is at the beginning, take the next value
    }
  }
  
  return(x)
}


############################################
# Stacking of hourly temperatures (chillR) #
############################################


stack_hourly_temps = function (weather = NULL, latitude = 50, hour_file = NULL, keep_sunrise_sunset = FALSE) 
{
  if (length(latitude) > 1) 
    stop("'latitude' has more than one element")
  if (!is.numeric(latitude)) 
    stop("'latitude' is not numeric")
  if (latitude > 90 | latitude < (-90)) 
    warning("'latitude' is usually between -90 and 90")
  if (is.null(weather) & (!is.null(hour_file))) 
    weather <- hour_file
  if ((length(names(weather)) == 2) & ("weather" %in% names(weather)) & 
      ("QC" %in% names(weather))) {
    if (!keep_sunrise_sunset) 
      THourly <- make_hourly_temps(latitude, weather$weather)
    if (keep_sunrise_sunset) 
      THourly <- make_hourly_temps(latitude, weather$weather, 
                                   keep_sunrise_sunset = TRUE)
    QC <- weather$QC
  }
  if ((length(names(weather)) > 24) & ("Hour_23" %in% names(weather))) {
    THourly <- weather
    QC <- NA
  }
  if ((length(names(weather)) > 2) & ("Tmax" %in% names(weather))) {
    if (!keep_sunrise_sunset) 
      THourly <- make_hourly_temps(latitude, weather)
    if (keep_sunrise_sunset) 
      THourly <- make_hourly_temps(latitude, weather, keep_sunrise_sunset = TRUE)
    QC <- NA
  }
  if (is.null(weather) & is.null(hour_file)) {
    tt = NA
    QC = NA
  }
  else {
    preserve_columns <- colnames(THourly)[1:(ncol(THourly) - 
                                               24)]
    THourly[, "index"] <- row(THourly)[, 1]
    ss <- stack(THourly, select = c(paste("Hour_", 0:23, 
                                          sep = "")))
    colnames(ss) <- c("Temp", "Hour")
    ss$Hour <- as.numeric(sapply(strsplit(as.character(ss$Hour), 
                                          "_"), "[", 2))
    for (pc in preserve_columns) {
      ss[, pc] <- rep(THourly[, pc], 24)
    }
    tt <- ss[with(ss, order(Year, JDay, Hour)), ]
    tt <- tt[c(preserve_columns, "Hour", "Temp")]
  }
  return(list(hourtemps = tt, QC = QC))
}


#####################################
# Make hourly temperatures (chillR) #
#####################################


make_hourly_temps = function (latitude, year_file, keep_sunrise_sunset = FALSE) 
{
    if (missing(latitude)) 
        stop("'latitude' not specified")
    if (length(latitude) > 1) 
        stop("'latitude' has more than one element")
    if (!is.numeric(latitude)) 
        stop("'latitude' is not numeric")
    if (latitude > 90 | latitude < (-90)) 
        warning("'latitude' is usually between -90 and 90")
    year_file <- year_file[which(!is.na(year_file$Tmin) & !is.na(year_file$Tmax)), 
        ]
    if (!"JDay" %in% colnames(year_file)) 
        year_file[, "JDay"] <- strptime(paste(year_file$Month, 
            "/", year_file$Day, "/", year_file$Year, sep = ""), 
            "%m/%d/%Y")$yday + 1
    preserve_columns <- colnames(year_file)
    Day_times <- daylength(latitude = latitude, JDay = c(year_file$JDay[1] - 
        1, year_file$JDay, year_file$JDay[nrow(year_file)] + 
        1))
    Day_times$Sunrise[which(Day_times$Sunrise == 99)] <- 0
    Day_times$Sunrise[which(Day_times$Sunrise == -99)] <- 12
    Day_times$Sunset[which(Day_times$Sunset == 99)] <- 24
    Day_times$Sunset[which(Day_times$Sunset == -99)] <- 12
    year_file$Sunrise <- Day_times$Sunrise[2:(length(Day_times$Sunrise) - 
        1)]
    year_file$Sunset <- Day_times$Sunset[2:(length(Day_times$Sunset) - 
        1)]
    year_file$Daylength <- Day_times$Daylength[2:(length(Day_times$Daylength) - 
        1)]
    year_file$prev_Sunset <- Day_times$Sunset[1:(length(Day_times$Sunset) - 
        2)]
    year_file$next_Sunrise <- Day_times$Sunrise[3:length(Day_times$Sunrise)]
    year_file$prev_max <- year_file$Tmax[c(NA, 1:(nrow(year_file) - 
        1))]
    year_file$next_min <- year_file$Tmin[c(2:nrow(year_file), 
        NA)]
    year_file$prev_min <- year_file$Tmin[c(NA, 1:(nrow(year_file) - 
        1))]
    year_file$Tsunset <- year_file$Tmin + (year_file$Tmax - year_file$Tmin) * 
        sin((pi * (year_file$Sunset - year_file$Sunrise)/(year_file$Daylength + 
            4)))
    year_file$prev_Tsunset <- year_file$prev_min + (year_file$prev_max - 
        year_file$prev_min) * sin((pi * (year_file$Daylength)/(year_file$Daylength + 
        4)))
    colnum <- ncol(year_file) + 1
    hourcol <- c(colnum:(colnum + 23))
    for (hour in 0:23) {
        hourcount <- hour + 1
        no_riseset <- which(year_file$Daylength %in% c(0, 24, 
            -99))
        year_file[no_riseset, colnum + hour] <- ((year_file$Tmax + 
            year_file$Tmin)/2)[no_riseset]
        c_morn <- which(hour <= year_file$Sunrise)
        if (1 %in% c_morn) 
            if (!length(c_morn) == 1) 
                c_morn <- c_morn[2:length(c_morn)]
            else c_morn <- c()
        c_day <- which(hour > year_file$Sunrise & hour <= year_file$Sunset)
        c_eve <- which(hour >= year_file$Sunset)
        if (nrow(year_file) %in% c_eve) 
            c_eve <- c_eve[1:(length(c_eve) - 1)]
        year_file[c_morn, colnum + hour] <- year_file$prev_Tsunset[c_morn] - 
            ((year_file$prev_Tsunset[c_morn] - year_file$Tmin[c_morn])/log(pmax(1, 
                24 - (year_file$prev_Sunset[c_morn] - year_file$Sunrise[c_morn]))) * 
                log(hour + 24 - year_file$prev_Sunset[c_morn] + 
                  1))
        year_file[c_day, colnum + hour] <- year_file$Tmin[c_day] + 
            (year_file$Tmax[c_day] - year_file$Tmin[c_day]) * 
                sin((pi * (hour - year_file$Sunrise[c_day])/(year_file$Daylength[c_day] + 
                  4)))
        year_file[c_eve, colnum + hour] <- year_file$Tsunset[c_eve] - 
            ((year_file$Tsunset[c_eve] - year_file$next_min[c_eve])/log(24 - 
                (year_file$Sunset[c_eve] - year_file$next_Sunrise[c_eve]) + 
                1) * log(hour - year_file$Sunset[c_eve] + 1))
    }
    colnames(year_file)[(ncol(year_file) - 23):(ncol(year_file))] <- c(paste("Hour_", 
        0:23, sep = ""))
    if (!keep_sunrise_sunset) 
        year_file <- year_file[, c(preserve_columns, paste("Hour_", 
            0:23, sep = ""))]
    if (keep_sunrise_sunset) 
        year_file <- year_file[, c(preserve_columns, "Sunrise", 
            "Sunset", "Daylength", paste("Hour_", 0:23, sep = ""))]
    year_file[1, (ncol(year_file) - 23):(ncol(year_file))][which(is.na(year_file[1, 
        (ncol(year_file) - 23):(ncol(year_file))]))] <- year_file[1, 
        "Tmin"]
    year_file[nrow(year_file), (ncol(year_file) - 23):(ncol(year_file))][which(is.na(year_file[nrow(year_file), 
        (ncol(year_file) - 23):(ncol(year_file))]))] <- year_file[nrow(year_file), 
        "Tmin"]
    return(year_file)
}


############################################################
# Compute sunrise and sunset times, and daylength (chillR) #
############################################################


daylength = function (latitude, JDay, notimes.as.na = FALSE) 
{
    if (missing(latitude)) 
        stop("'latitude' not specified")
    if (missing(JDay)) 
        stop("'JDay' not specified")
    if (!isTRUE(all(is.numeric(JDay)))) 
        stop("'JDay' contains non-numeric values")
    if (length(latitude) > 1) 
        stop("'latitude' has more than one element")
    if (!is.numeric(latitude)) 
        stop("'latitude' is not numeric")
    if (latitude > 90 | latitude < (-90)) 
        warning("'latitude' is usually between -90 and 90")
    Gamma <- 2 * pi/365 * ((JDay) - 1)
    Delta <- 180/pi * (0.006918 - 0.399912 * cos(Gamma) + 0.070257 * 
        sin(Gamma) - 0.006758 * cos(2 * Gamma) + 0.000907 * sin(2 * 
        (Gamma)) - 0.002697 * cos(3 * (Gamma)) + 0.00148 * sin(3 * 
        (Gamma)))
    CosWo <- (sin(-0.8333/360 * 2 * pi) - sin(latitude/360 * 
        2 * pi) * sin(Delta/360 * 2 * pi))/(cos(latitude/360 * 
        2 * pi) * cos(Delta/360 * 2 * pi))
    normal_days <- which(CosWo >= -1 & CosWo <= 1)
    Sunrise <- rep(-99, length(CosWo))
    Sunrise[normal_days] <- 12 - acos(CosWo[normal_days])/(15/360 * 
        2 * pi)
    Sunset <- rep(-99, length(CosWo))
    Sunset[normal_days] <- 12 + acos(CosWo[normal_days])/(15/360 * 
        2 * pi)
    Daylength <- Sunset - Sunrise
    Daylength[which(CosWo > 1)] <- 0
    Daylength[which(CosWo < (-1))] <- 24
    Sunrise[which(Daylength == 24)] <- 99
    Sunset[which(Daylength == 24)] <- 99
    if (notimes.as.na) {
        Sunrise[which(Sunrise %in% c(-99, 99))] <- NA
        Sunset[which(Sunset %in% c(-99, 99))] <- NA
    }
    Sunset[which(is.na(JDay))] <- NA
    Sunrise[which(is.na(JDay))] <- NA
    Daylength[which(is.na(JDay))] <- NA
    return(list(Sunrise = Sunrise, Sunset = Sunset, Daylength = Daylength))
}



##############################################################################################################################################
##############################################################################################################################################



##############################
## Merge datasets into list ##
##############################



# Identifier 1 (all site x year combinations)
PEP.df$site_year = paste0(PEP.df$s_id,"_",PEP.df$year)

# Identifier 2 (all timeseries x year combinations)
PEP.df$ts_yr     = paste0(PEP.df$timeseries,"_",PEP.df$year)
timeseries_year  = unique(PEP.df$ts_yr)

# add PEP data (+plant functional type label) and photoperiod to list
DataList[[7]] = photo.df
DataList[[8]] = PEP.df

rm(photo.df, data, PEP.df)
names(DataList)=c(vn,"photoperiod","PEP")
names(DataList)




#############################################################
## Loop through each observations using parallel computing ##
#############################################################



parallelCalc <- function(timeseries_years){ 

  # Subset input data by time-point
  #################################
  
  #phenology data
  pheno.sub  <- DataList[[8]][which(DataList[[8]]$ts_yr==timeseries_years),]
  
  # Define the length of the period of interest in calendar units
  year1     <- as.character(pheno.sub$year-1)
  year2     <- as.character(pheno.sub$year)
  start_doy <- paste(year1,"-01-01", sep="") 
  end_doy   <- paste(year2,"-06-30", sep="")
  days      <- seq(as.Date(start_doy), as.Date(end_doy), by="days")
  
  #get number of days in year 1 and year 2
  days_year1 <- length(seq(as.Date(start_doy), as.Date(paste(year1,"-12-31", sep="")), by="days"))
  days_year2 <- length(seq(as.Date(paste0(year2,"-01-01")), as.Date(end_doy), by="days"))
  
  
  # ------------------------------------------------------------------------
  # Short Name        ; Long Name                               ; Unit [d-1]
  # ------------------------------------------------------------------------
  # SWRAD             ; Surface shortwave downwelling radiation ; W m-2
  # PRCP              ; Precipitation rate                      ; mm d-1
  # RH                ; Relative humidity                       ; %
  # TMIN              ; Minimum Air Temperature                 ; degC
  # TMEAN             ; Mean Air Temperature                    ; degC
  # TMAX              ; Maximum Air Temperature                 ; degC
  # ------------------------------------------------------------------------
  
  
  #Tmean
  T_mean.sub2 <- DataList[[1]][which(DataList[[1]]$site_year==pheno.sub$site_year),]%>%
    dplyr::select(as.character(1:days_year2))
  
  #--------------------------------------
  #Skip timeseries for which there is: 
  if (
    # 1) no climate data for current year and/or
    nrow(T_mean.sub2)==0 |
    # 2) no climate data for previous year
    nrow(DataList[[1]][which(DataList[[1]]$s_id==pheno.sub$s_id & DataList[[1]]$year==year1),])==0
  ) {
    # safe table
    write.table(pheno.sub, file=paste0(PEP_drivers_path3, '/', timeseries_years, '.csv'), 
                sep=',', row.names = F, col.names = T)
  } else {
    #--------------------------------------
    
    T_mean.sub1 <- DataList[[1]][which(DataList[[1]]$s_id==pheno.sub$s_id & DataList[[1]]$year==year1),]%>% 
      dplyr::select(as.character(1:days_year1))
    T_mean.sub  <- cbind(T_mean.sub1, T_mean.sub2)
    
    #Tmin
    T_min.sub1 <- DataList[[2]][which(DataList[[2]]$s_id==pheno.sub$s_id & DataList[[2]]$year==year1),]%>% 
      dplyr::select(as.character(1:days_year1))
    T_min.sub2 <- DataList[[2]][which(DataList[[2]]$site_year==pheno.sub$site_year),]%>%
      dplyr::select(as.character(1:days_year2))
    T_min.sub  <- cbind(T_min.sub1, T_min.sub2)
    
    #Tmax
    T_max.sub1 <- DataList[[3]][which(DataList[[3]]$s_id==pheno.sub$s_id & DataList[[3]]$year==year1),]%>% 
      dplyr::select(as.character(1:days_year1))
    T_max.sub2 <- DataList[[3]][which(DataList[[3]]$site_year==pheno.sub$site_year),]%>%
      dplyr::select(as.character(1:days_year2))
    T_max.sub  <- cbind(T_max.sub1, T_max.sub2)
    
    #Prcp
    prcp.sub1 <- DataList[[4]][which(DataList[[4]]$s_id==pheno.sub$s_id & DataList[[4]]$year==year1),]%>% 
      dplyr::select(as.character(1:days_year1))
    prcp.sub2 <- DataList[[4]][which(DataList[[4]]$site_year==pheno.sub$site_year),]%>%
      dplyr::select(as.character(1:days_year2))
    prcp.sub  <- cbind(prcp.sub1, prcp.sub2)
    
    #Qair
    rh.sub1 <- DataList[[5]][which(DataList[[5]]$s_id==pheno.sub$s_id & DataList[[5]]$year==year1),]%>% 
      dplyr::select(as.character(1:days_year1))
    rh.sub2 <- DataList[[5]][which(DataList[[5]]$site_year==pheno.sub$site_year),]%>%
      dplyr::select(as.character(1:days_year2))
    rh.sub  <- cbind(rh.sub1, rh.sub2)
    
    #SWrad
    swrad.sub1 <- DataList[[6]][which(DataList[[6]]$s_id==pheno.sub$s_id & DataList[[6]]$year==year1),]%>% 
      dplyr::select(as.character(1:days_year1))
    swrad.sub2 <- DataList[[6]][which(DataList[[6]]$site_year==pheno.sub$site_year),]%>%
      dplyr::select(as.character(1:days_year2))
    swrad.sub  <- cbind(swrad.sub1, swrad.sub2)
    
    #Photo
    photo.sub1 <- DataList[[7]][which(DataList[[7]]$lat==pheno.sub$lat),][1]%>% 
      dplyr::select(as.character(1:days_year1))
    photo.sub2 <- DataList[[7]][which(DataList[[7]]$lat==pheno.sub$lat),][1]%>%
      dplyr::select(as.character(1:days_year2))
    photo.sub  <- cbind(photo.sub1, photo.sub2)
    
    
    ##############################################################################################################################################
    
    
    ###############################
    # Create table of daily climate
    ###############################
    
    
    # Generate sub-dataframe to store results
    factors.sub <- pheno.sub %>% 
      dplyr::select(s_id,species,timeseries,year,lat,lon,alt,alt_dem,
                    leaf_out,leaf_out_prev,leaf_out_mean,leaf_out_sd,leaf_off,n.years,n.years.reduced)
    factors.sub = as.data.frame(factors.sub) 
    
    #create table
    daily_vals <- data.frame(Year     = lubridate::year(as.Date(days,origin=days[1])),
                             Month    = lubridate::month(as.Date(days,origin=days[1])),
                             Day      = lubridate::day(as.Date(days,origin=days[1])),
                             DOY      = as.numeric(names(T_mean.sub)),
                             Tmin     = as.numeric(T_min.sub), 
                             Tmean    = as.numeric(T_mean.sub), 
                             Tmax     = as.numeric(T_max.sub), 
                             SWrad    = as.numeric(swrad.sub),
                             Prcp     = as.numeric(prcp.sub), 
                             RH       = as.numeric(rh.sub),
                             DL       = as.numeric(photo.sub)) %>% 
      #replace NAs with neighbor mean
      mutate(across(c(Tmin, Tmean, Tmax, SWrad, Prcp, RH, DL), replace_na_with_neighbor_mean))
    
    
    ##############################################################################################################################################
    
    
    ######################
    # Add chilling and GDD
    ######################
    
    
    # Get hourly values
    hourly_vals = stack_hourly_temps(daily_vals, latitude=pheno.sub$lat)$hourtemps
    
    # Chilling and Forcing hours
    Chill_days = hourly_vals %>%
      #add Chilling and Forcing units
      mutate(Chill = sapply(Temp, Chill.fun, threshold_low=-10, threshold_up=10),
             GDD   = sapply(Temp, GDD.fun,   threshold_low=5)) %>%
      #summarise hourly to daily values
      group_by(Year, Month, Day) %>%
      summarise(Chill = mean(Chill, na.rm=T),
                GDD   = mean(GDD, na.rm=T))%>%
      ungroup() %>%
      #order
      dplyr::select(Chill, GDD) 
    
    # define November first as start date of chilling accumulation
    start_row <-  which(daily_vals$Month==11 & daily_vals$Day==1)  
    
    # Cbind
    daily_vals = cbind(daily_vals, Chill_days) %>%
      mutate(
        #continous count from Jan 1
        id = row_number(),
        # Sum chilling days from 1 November
        Chill_sum = ifelse(id < start_row, 0, cumsum(ifelse(id >= start_row, Chill, 0))),
        # Add chilling/photoperiod-adjusted GDDs
        GDD.chill.lo = adjust_gdd(GDD=GDD, Chilling=Chill_sum, requirement=75),
        GDD.chill.hi = adjust_gdd(GDD=GDD, Chilling=Chill_sum, requirement=150),
        GDD.chill.DL = adjust_gdd(GDD=GDD, Chilling=Chill_sum, requirement=150, include_photo = T, DL = DL)) %>% 
      #order
      dplyr::select(id, Year, Month, Day, DOY, Tmin, Tmean, Tmax, everything()) 
    
    
    ##############################################################################################################################################
    
    
    #####################
    # Get important dates
    #####################
    
    
    #row of leaf-out date
    DOY_out      = which(daily_vals$DOY==pheno.sub$leaf_out & daily_vals$Year==year2)   
    
    #leaf-out date of the previous year
    DOY_out_prev = pheno.sub$leaf_out_prev
    
    #row of mean leaf-out date
    DOY_out_mean = which(daily_vals$DOY==pheno.sub$leaf_out_mean & daily_vals$Year==year2)   
    
    #longest day of year
    solstice1    = which(daily_vals$DL==max(daily_vals$DL))[1] 
    
    #shortest day of year
    solstice2    = which(daily_vals$DL==min(daily_vals$DL))[1] 
    
    #March equinox previous year
    equinox1     = solstice1 - 93       
    
    #September equinox
    equinox2     = solstice2 - 90      
    
    #March equinox
    equinox3     = solstice2 + 89                                    
    
    #May 1
    May1         = which(daily_vals$Day==1 & daily_vals$Month==5 & daily_vals$Year==year1)
    
    #Sep 30
    Sep30        = which(daily_vals$Day==30 & daily_vals$Month==9 & daily_vals$Year==year1)      
    
    #Jul 15
    Jul15        = which(daily_vals$Day==15 & daily_vals$Month==7 & daily_vals$Year==year1)      
    
    
    ##############################################################################################################################################
    
    
    ####################################
    # Late spring frost in previous year
    ####################################
    
    
    # was there a frost event below -2°C from leaf-out until July 15?
    factors.sub$LSF2 = ifelse(any(daily_vals$Tmin[DOY_out_prev:Jul15] < -2, na.rm=T), 1, 0)
    
    # was there a frost event below -4°C from leaf-out until July 15?
    factors.sub$LSF4 = ifelse(any(daily_vals$Tmin[DOY_out_prev:Jul15] < -4, na.rm=T), 1, 0)
    
    
    ##############################################################################################################################################
    
    
    #######################################
    # Get drivers before and after solstice
    #######################################
    
    
    #clean daily data table
    daily_vals = daily_vals %>%
      dplyr::select(-c(DL, Chill_sum))
    
    #define variables
    variable.names = c("Tmin","Tmean","Tmax","SWrad","Prcp","RH","Chill")
    
    for(i in 1:length(variable.names)) {
      
      #choose variable (daily values)
      variable = daily_vals[,variable.names[i]]
      
      # Name variables
      ################
      
      # Seasonal
      ##########
      
      # EQ1...March equinox previous year
      # SO1...Summer solstice (~22 June)
      # EQ2...September equinox previous year
      # SO2...Winter solstice (~22 Dec)
      # EQ3...March equinox
      # LO...leaf-out date
      # GS...period from May 1 to September 30
      
      varname.GS          <- paste(variable.names[i], "GS",    sep=".")
      
      varname.EQ1.SO1     <- paste(variable.names[i], "EQ1.SO1",    sep=".")
      varname.EQ1.EQ2     <- paste(variable.names[i], "EQ1.EQ2",    sep=".")
      varname.EQ1.SO2     <- paste(variable.names[i], "EQ1.SO2",    sep=".")
      varname.EQ1.EQ3     <- paste(variable.names[i], "EQ1.EQ3",    sep=".")
      varname.EQ1.LO      <- paste(variable.names[i], "EQ1.LO",     sep=".")
      
      varname.SO1.EQ2     <- paste(variable.names[i], "SO1.EQ2",    sep=".")
      varname.SO1.SO2     <- paste(variable.names[i], "SO1.SO2",    sep=".")
      varname.SO1.EQ3     <- paste(variable.names[i], "SO1.EQ3",    sep=".")
      varname.SO1.LO      <- paste(variable.names[i], "SO1.LO",     sep=".")
      
      varname.EQ2.SO2     <- paste(variable.names[i], "EQ2.SO2",    sep=".")
      varname.EQ2.EQ3     <- paste(variable.names[i], "EQ2.EQ3",    sep=".")
      varname.EQ2.LO      <- paste(variable.names[i], "EQ2.LO",     sep=".")
      varname.EQ2.LOsum   <- paste(variable.names[i], "EQ2.LOsum",  sep=".")
      
      varname.SO2.EQ3     <- paste(variable.names[i], "SO2.EQ3",    sep=".")
      varname.SO2.LO      <- paste(variable.names[i], "SO2.LO",     sep=".")
      
      varname.LO.SO1      <- paste(variable.names[i], "LO.SO1",    sep=".")
      varname.LO.EQ2      <- paste(variable.names[i], "LO.EQ2",    sep=".")
      varname.LO.SO2      <- paste(variable.names[i], "LO.SO2",    sep=".")
      varname.LO.EQ3      <- paste(variable.names[i], "LO.EQ3",    sep=".")
      varname.LO.LO       <- paste(variable.names[i], "LO.LO",     sep=".")
      
      # Solstice
      ##########
      
      # solstice1...sum of 40 to 10 days before solstice
      # solstice2...sum of 30 to 0 days before solstice
      # solstice3...sum of 20 days before to 10 days after solstice
      # solstice4...sum of 10 days before to 20 days after solstice
      # solstice5...sum of 0 to 30 days after solstice
      # solstice6...sum of 10 to 40 days after solstice
      varname.solstice1   <- paste(variable.names[i], "solstice1", sep=".")
      varname.solstice2   <- paste(variable.names[i], "solstice2", sep=".")
      varname.solstice3   <- paste(variable.names[i], "solstice3", sep=".")
      varname.solstice4   <- paste(variable.names[i], "solstice4", sep=".")
      varname.solstice5   <- paste(variable.names[i], "solstice5", sep=".")
      varname.solstice6   <- paste(variable.names[i], "solstice6", sep=".")
      
      
      # Extract drivers
      #################
      
      factors.sub = factors.sub %>%
        mutate(
          #seasonal
          !!varname.GS          := mean(variable[May1:Sep30]),
          
          !!varname.EQ1.SO1     := mean(variable[equinox1:solstice1]),
          !!varname.EQ1.EQ2     := mean(variable[equinox1:equinox2]),
          !!varname.EQ1.SO2     := mean(variable[equinox1:solstice2]),
          !!varname.EQ1.EQ3     := mean(variable[equinox1:equinox3]),
          !!varname.EQ1.LO      := mean(variable[equinox1:DOY_out_mean]),
          
          !!varname.SO1.EQ2     := mean(variable[solstice1:equinox2]),
          !!varname.SO1.SO2     := mean(variable[solstice1:solstice2]),
          !!varname.SO1.EQ3     := mean(variable[solstice1:equinox3]),
          !!varname.SO1.EQ2     := mean(variable[solstice1:equinox2]),
          !!varname.SO1.LO      := mean(variable[solstice1:DOY_out_mean]),
          
          !!varname.EQ2.SO2     := mean(variable[equinox2:solstice2]),
          !!varname.EQ2.EQ3     := mean(variable[equinox2:equinox3]),
          !!varname.EQ2.LO      := mean(variable[equinox2:DOY_out_mean]),
          !!varname.EQ2.LOsum   := sum(variable[equinox2:DOY_out_mean]),
          
          !!varname.SO2.EQ3     := mean(variable[solstice2:equinox3]),
          !!varname.SO2.LO      := mean(variable[solstice2:DOY_out_mean]),
          
          !!varname.LO.SO1      := sum(variable[DOY_out_prev:solstice1]),
          !!varname.LO.EQ2      := sum(variable[DOY_out_prev:equinox2]),
          !!varname.LO.SO2      := sum(variable[DOY_out_prev:solstice2]),
          !!varname.LO.EQ3      := sum(variable[DOY_out_prev:equinox3]),
          !!varname.LO.LO       := sum(variable[DOY_out_prev:DOY_out_mean]),
          
          #solstice
          !!varname.solstice1   := mean(variable[(solstice2-39):(solstice2-10)]),
          !!varname.solstice2   := mean(variable[(solstice2-29):solstice2]),
          !!varname.solstice3   := mean(variable[(solstice2-19):(solstice2+10)]),
          !!varname.solstice4   := mean(variable[(solstice2-9):(solstice2+20)]),
          !!varname.solstice5   := mean(variable[(solstice2+1):(solstice2+30)]),
          !!varname.solstice6   := mean(variable[(solstice2+11):(solstice2+40)]) )
    }
    
    
    ##############################################################################################################################################
    
    
    #######################################################
    # Add accumulated GDDs from winter solstice to leaf-out
    #######################################################
    
    
    factors.sub = factors.sub %>%
      mutate(
        GDD          = sum(daily_vals$GDD[solstice2:DOY_out]),
        GDD.chill.lo = sum(daily_vals$GDD.chill.lo[solstice2:DOY_out]),
        GDD.chill.hi = sum(daily_vals$GDD.chill.hi[solstice2:DOY_out]),
        GDD.chill.DL = sum(daily_vals$GDD.chill.DL[solstice2:DOY_out]))
    
    
    ##############################################################################################################################################
    
    
    ################################
    # Calculate the monthly averages
    ################################
    
    
    #get means and sums
    variable.names = c("Tmin", "Tmean", "Tmax", "SWrad", "RH")
    monthly_means  = data.frame(daily_vals %>% 
                                  group_by(Year,Month) %>% 
                                  summarize_at(variable.names, mean, na.rm = TRUE))
    
    #Transform data
    monthly_means = as.data.frame(t(monthly_means))
    
    #Add to table
    #############
    
    #loop through variables
    for(i in 1:length(variable.names)) {
      #select  variable
      MONTHLY.DF = monthly_means[variable.names[i],]
      #add column names
      names(MONTHLY.DF)=paste0(row.names(MONTHLY.DF), '.', c(1:18))
      #cbind with table
      factors.sub = cbind(factors.sub, MONTHLY.DF)
    }
    
    
    ##############################################################################################################################################
    
    
    ############################
    # Get preseason temperatures
    ############################
    
    
    ## Calculate the average preseason temperatures (15 to 120 days prior to mean leaf-out date)
    
    #get preseason length vector (10 to 120 days with 10-day steps)
    preseason.lengths = seq(10, 120, 10)
    
    #loop through preseasons
    for(preseason.length in preseason.lengths) {
      #name columns
      preseason.Tmean  <- paste0("Tmean.PS",  preseason.length)
      preseason.Tmax   <- paste0("Tmax.PS",   preseason.length)
      preseason.Tmin   <- paste0("Tmin.PS",   preseason.length)
      preseason.SWrad  <- paste0("SWrad.PS",  preseason.length)
      preseason.RH     <- paste0("RH.PS",     preseason.length)
      #add columns to table
      factors.sub = factors.sub %>%
        mutate(!!preseason.Tmean   := mean(daily_vals$Tmean[(DOY_out_mean-preseason.length):DOY_out_mean]),
               !!preseason.Tmax    := mean(daily_vals$Tmax[(DOY_out_mean-preseason.length):DOY_out_mean]),
               !!preseason.Tmin    := mean(daily_vals$Tmin[(DOY_out_mean-preseason.length):DOY_out_mean]),
               !!preseason.SWrad   := mean(daily_vals$SWrad[(DOY_out_mean-preseason.length):DOY_out_mean]),
               !!preseason.RH      := mean(daily_vals$RH[(DOY_out_mean-preseason.length):DOY_out_mean]) )
    }
    
    
    ##############################################################################################################################################
    
    
    # Safe the table   
    write.table(factors.sub, file=paste0(PEP_drivers_path, '/', timeseries_years, '.csv'), sep=',', row.names =F, col.names = T)
    
  }
}



##############################################################################################################################################
##############################################################################################################################################



#initialize the loop
outputlist <- pbmclapply(timeseries_year[1:5], parallelCalc, mc.cores=12, mc.preschedule=F)

#check how many files there are
length(list.files(path=PEP_drivers_path, pattern='.csv'))
#safe number of files
length.df=data.frame(n.files=length(list.files(path=PEP_drivers_path, pattern='.csv')) )
write.table(length.df, file=paste0(PEP_drivers_path2, '/n.files.subset.csv'), sep=',', row.names =F, col.names = T)

#Rbind files
climate.factors.table = rbindlist(lapply(list.files(path = PEP_drivers_path), 
                                         function(n) fread(file.path(PEP_drivers_path, n))))



##############################################################################################################################################
##############################################################################################################################################



###################
## Safe the data ##
###################



#Safe table
write.csv(climate.factors.table, paste(PEP_drivers_path2, "pep_drivers_spring_data_subset.csv", sep="/"))

#Remove individual files
do.call(file.remove, list(list.files(PEP_drivers_path, 
                                     full.names = TRUE)))

do.call(file.remove, list(list.files(PEP_drivers_path3, 
                                     full.names = TRUE)))



#####################
## Reproducibility ##	
#####################



## date time
Sys.time()

## session info
sessionInfo()



##############################################################################################################################################
#############################################################THE END##########################################################################
##############################################################################################################################################
```