---
title: Spring phenology tier2 project
author: Constantin Zohner
date: "last updated 5 June 2025"

subtitle: Format and clean MODIS data
output:
  html_document:
    highlight: haddock
    toc: false
    use_bookdown: true
    df_print: paged
    fig_caption: yes
---

<style>
body
  { counter-reset: source-line 0; }
pre.numberSource code
  { counter-reset: none; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  class.source = "numberLines lineAnchors"
  )
```

<br><br>

## 1. Load packages, set directories, and add functions

<details><summary>get packages</summary>
```{r, mesage=F, warning=F}
require(data.table)
require(tidyverse)
require(raster)
require(sf)
require(ggplot2)
require(maps)
require(terra)
```
</details>

<details><summary>set directories</summary>
```{r}
# paths
MODISraw.dir  = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/MODIS_Analysis/MODIS_data/MODIS_annual"
MODISout.dir  = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/MODIS_Analysis/MODIS_data"
biome.dir     = "/Users/consti/Desktop/PhD/Data/Global_layers/Ecology_layers/biomes"
LandCover.dir = "/Users/consti/Desktop/PhD/Data/Global_layers/Ecology_layers/MODIS_LandCover/Maps"
figure.dir    = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/MODIS_Analysis/Analysis_output/LocationMaps"
```
</details>

<details><summary>add functions</summary>
```{r}
# Define a function to extract quality scores from the QA column
UnpackDetailedQA <- function(x){
  bits  <- as.integer(intToBits(x))  # Convert integer to binary bits
  quals <- sapply(seq(1, 16, by=2), function(i) sum(bits[i:(i+1)] * 2^c(0, 1)))[1:7]
  return(quals) }
```
</details>

<br>

## 2. Load, merge, and revise MODIS phenology data

<details><summary>show code</summary>
```{r}
# Phenology data
################

MODIS.df = list.files(
  path = MODISraw.dir, # Define the directory containing CSV files
  pattern = "*.csv", # Look for all CSV files
  full.names = TRUE  # Use full file paths
  ) %>% 
  purrr::map_df(~ {
    df <- fread(.x)  # Read each CSV file into a dataframe
    
    # Extract the year from the filename (e.g., "Phenology_Data_for_Year_2001.csv")
    year    <- str_extract(basename(.x), "\\d{4}")
    df$year <- as.integer(year)  # Convert the extracted year to an integer
    
    # Apply QA unpacking function to each value in the QA Detailed column
    if ("QA_Overall_1" %in% names(df)) {  # Check if column exists
      qa_matrix <- t(sapply(df$QA_Overall_1, UnpackDetailedQA)) # Unpack QA values
      colnames(qa_matrix) <- c("Greenup_QA", "MidGreenup_QA", "Maturity_QA", "Peak_QA",
                               "Senescence_QA", "MidGreendown_QA", "Dormancy_QA")
      df <- cbind(df, qa_matrix)  # Add extracted QA scores to the dataframe
    }
    
    return(df) # Return the processed dataframe
  }) %>% 
  
  # Remove unwanted columns: second cycle data, specific phenological stages, and QA columns
  dplyr::select(-matches("2"), # Remove all columns containing "2" in their names (second cycle)
                -QA_Overall_1, -NumCycles, # Remove overall QA and cycle count
                -Peak_1, -Maturity_1, -Dormancy_1, # Remove specific phenological stages
                -Peak_QA, -Maturity_QA, -Dormancy_QA) %>% # Remove corresponding QA scores
  
  # Rename columns by removing "_1" (ensuring consistency in naming)
  rename_with(~ gsub("_1", "", .x)) %>% 
  
  # Rename spatial columns: x → lon, y → lat
  rename(lon = x, lat = y) %>%
  
  # Remove duplicate records based on unique (lon, lat, year) combinations
  distinct(lon, lat, year, .keep_all = T) 

# Print first row of the updated MODIS dataset
MODIS.df[1,]
```
</details>

<br>

## 3. Add biome and land cover info

<details><summary>show code</summary>
```{r}
# Biome layer
#############

biome.layers.list <- list.files(
  path=paste(biome.dir), # Directory containing biome raster files
  pattern='*.tif',       # Look for all .tif files (adjust if using .asc)
  full.names=TRUE      # Use full file paths
) 

# Load the first biome raster
biome.layer <- rast(biome.layers.list[1])

# Assign categorical colors
biome_colors <- terrain.colors(n = length(unique(values(biome.layer))))

# Plot with categorical colors
plot(as.factor(biome.layer), col = biome_colors, main = "Biome Classification", legend = TRUE)


# Land cover layer (PFT classification)
#######################################

LC.layers.list <- list.files(
  path=paste(LandCover.dir), # Directory containing biome raster files
  pattern='*.tif',       # Look for all .tif files (adjust if using .asc)
  full.names=TRUE      # Use full file paths
) 
LC.layers.list
# see MCD12_User_Guide_V61.pdf for details
#-----------------------------------------
# MODIS_LandCover_2010_WGS84_0.5deg.tif → Land Cover Type 1 LC Type1 Annual IGBP classification
# MODIS_PFT_2010_WGS84_0.5deg.tif       → Land Cover Type 5 LC Type5 Annual PFT classification
#-----------------------------------------

# Define the full file path
pft_file <- "/Users/consti/Desktop/PhD/Data/Global_layers/Ecology_layers/MODIS_LandCover/Maps/MODIS_PFT_2010_WGS84_0.5deg.tif"

# Load the raster
LC.layer <- rast(pft_file)

# Assign categorical colors
LC_colors <- terrain.colors(n = length(unique(values(LC.layer))))

# Plot with categorical colors
plot(as.factor(LC.layer), col = biome_colors, main = "Land Cover Classification", legend = TRUE)


# Ensure MODIS has correct lat/lon column names before conversion
if (!("lon" %in% names(MODIS.df) & "lat" %in% names(MODIS.df))) {
  stop("MODIS dataset must contain 'lon' and 'lat' columns")
}

# Convert MODIS and LC dataframes to a spatial object (sf format)
MODIS_sf <- st_as_sf(MODIS.df, coords = c("lon", "lat"), crs = crs(biome.layer))
LC_sf    <- st_as_sf(MODIS.df, coords = c("lon", "lat"), crs = crs(LC.layer))

# Extract biome and LC values at MODIS locations
MODIS.df$Biome <- terra::extract(biome.layer, MODIS_sf)[,2]  
MODIS.df$LC    <- terra::extract(LC.layer, MODIS_sf)[,2]  

# Create a new column "ForestType" based on Biome and Land Cover (LC)
MODIS.df$ForestType <- with(MODIS.df, case_when(
  
  LC == 4 & Biome %in% c(4,5,8,10) ~ "Temperate deciduous broadleaf",
  LC == 4 & Biome %in% c(6,11)     ~ "Boreal deciduous broadleaf",
  LC == 3                          ~ "Deciduous Needleleaf",
  
  TRUE ~ NA_character_  # Assign NA to all other cases
))

# Filter MODIS data to include only specific biome classifications
#--------------------------------------
# Biome Codes:
# 4  → Temperate Broadleaf/Mixed Forest
# 5  → Temperate Conifer Forest
# 8  → Temperate Grassland/Savanna
# 10 → Montane Grassland/Shrublands
# 6  → Boreal Forest
# 11 → Tundra
#--------------------------------------

# Filter MODIS PFT data to include only specific land cover classifications
#--------------------------------------
# Land cover Codes:
# 3  → Deciduous Needleleaf Trees
# 4  → Deciduous Broadleaf Trees
#--------------------------------------

MODIS.df <- MODIS.df %>% 
  filter(Biome %in% c(4,5,6,8,10,11),
         LC %in% c(3, 4),
         lat>0)

# Print first row of the updated MODIS dataset
MODIS.df[1,]
```
</details>

<br>

## 4. Format and clean MODIS data

<details><summary>show code</summary>
```{r}
# Filter Greenup data
MODIS.df <- MODIS.df %>%
  filter(Greenup > 0 & MidGreenup > 0 &  # Ensure positive values
         Greenup < MidGreenup &  # Greenup before MidGreenup
         MidGreenup < Senescence &  # MidGreenup before Senescence
         Senescence < MidGreendown &  # Senescence before MidGreendown
         Greenup < 173 &  # Greenup occurs before the summer solstice
         MidGreenup < 213 &  # MidGreenup before August 1
         Senescence > 173 &  # Senescence occurs after the summer solstice
         MidGreendown > 213 & # MidGreendown after August 1
         Senescence < 365 & MidGreendown < 365)  # Both Senescence & MidGreendown within the year

# Create a dataset for previous year's leaf-out and senescence data
MODISprev.df <- MODIS.df %>%
  
  # Select relevant columns
  dplyr::select(lon, lat, year, Greenup, MidGreenup, Senescence, MidGreendown) %>%
  
  # Shift year forward to match previous year's phenology with the current year
  mutate(year = year + 1) %>%
  
  # Rename phenology columns to indicate previous year values
  rename(Greenup_prev = Greenup,
         MidGreenup_prev = MidGreenup,
         Senescence_prev = Senescence,
         MidGreendown_prev = MidGreendown)

# Merge previous year data with current data using LEFT JOIN (to avoid data loss)
MODIS.df <- left_join(MODIS.df, 
                       MODISprev.df, 
                       by = c("lon", "lat", "year")) %>%
  
  # create pixel identifier
  mutate(s_id = paste(lat, lon, sep ="_")) %>% 
  
  # filter rows lacking phenology data for previous year
  filter(!is.na(MidGreenup_prev),
         !is.na(Greenup_prev),
         !is.na(MidGreendown_prev)) %>%
  
  # Group by location (lon, lat)
  group_by(s_id) %>%

  # Count valid Greenup years per location BEFORE filtering
  mutate(n.years = sum(!is.na(Greenup))) %>% 

  # Remove pixels with fewer than 10 years of valid data
  filter(n.years >= 10) %>%  
  
  ungroup()

# Print first row of the updated MODIS dataset
MODIS.df[1,]
```
</details>

<br>

## 5. Sample sizes

<details><summary>show code</summary>
```{r}
# Total observations
nrow(MODIS.df)

# How many pixels in total?
n_distinct(MODIS.df$s_id)

# Time span 
range(MODIS.df$year)
paste0("Minimum timeseries length = ",   min(MODIS.df$n.years, na.rm = T),
       "; Maximum timeseries length = ", max(MODIS.df$n.years, na.rm = T))

# Histograms
hist(MODIS.df$n.years, xlab="Number of years", main="Time series length", breaks = 30)
hist(MODIS.df$year, xlab="Year", main="Temporal distribution of data", breaks = 30)

# Latitudinal and longitudinal gradients
range(MODIS.df$lon)
range(MODIS.df$lat)
hist(MODIS.df$lon, xlab="Longitude", main="Longitudinal gradient", breaks = 100)
hist(MODIS.df$lat, xlab="Latitude",  main="Latitudinal gradient",  breaks = 100)


# Distribution of phenology metrics
###################################

# Define all phenology metrics
metrics <- c("Greenup", "MidGreenup", "Senescence", "MidGreendown")

for (metric in metrics) {
  print(paste("Summary for", metric))
  print(mean(MODIS.df[[metric]], na.rm = TRUE))
  print(sd(MODIS.df[[metric]], na.rm = TRUE))
  print(range(MODIS.df[[metric]], na.rm = TRUE))
  
  # Histogram for overall distribution
  hist(MODIS.df[[metric]], 
       xlab = paste(metric, "date"), 
       main = paste(metric, "gradient"), 
       breaks = 100)
  
  # Histogram by Forest Type using ggplot2
  p = ggplot(MODIS.df, aes(x = .data[[metric]])) +
  geom_histogram(bins = 100, fill = "steelblue", color = "black", alpha = 0.7) +
  xlab(paste(metric, "Date")) +
  ylab("Frequency") +
  ggtitle(paste(metric, "Gradient by Forest Type")) +
  facet_wrap(~ ForestType, scales = "free_y") +
  theme_classic()
  
  print(p)
}


# Forest type-level statistics
##############################

# Create summary dataframe per Forest Type
sample.size = MODIS.df %>%
  group_by(ForestType) %>% # Group by Forest Type
  summarize(n.time.series  = n_distinct(lat, lon), # Count unique spatial locations (pixels)
            n.observations = n()) # Count total observations

# Display sample size summary
data.frame(sample.size)

# Total time series count
sum(sample.size$n.time.series)

# Convert wide to long format for visualization
sample.size <- sample.size %>%
  pivot_longer(cols = c(n.time.series, n.observations), # Columns to pivot
               names_to = "Metric",                     # New column for variable names
               values_to = "Count")                     # New column for values

# Bar Plot for Sample Size
ggplot(sample.size, aes(x = factor(ForestType), y = Count, fill = Metric)) +  
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  facet_wrap(~Metric, nrow=2, scales="free_y") +
  scale_fill_manual(values = c("#1f78b4", "#33a02c")) +
  labs(x = "Forest Type", y = "Count", fill = "Metric") +
  theme_classic()

# Temporal coverage per Forest Type
ggplot(MODIS.df, aes(x = year)) +
  geom_histogram(position = 'identity', binwidth = 1, fill = "steelblue", color = "black") +
  facet_wrap(~ForestType, ncol=5, scales = "free_y") +
  theme_classic() +
  labs(title = "Temporal Coverage per Forest Type", x = "Year", y = "Count")


# Location maps
###############

# Create base map with world borders
mapWorld <- borders("world", colour="gray90", fill="gray90") 

# Initialize ggplot with map
mp <- ggplot() + mapWorld

# Filter unique stations (one per pixel)
MODIS_unique <- MODIS.df[!duplicated(MODIS.df[, c("lat", "lon")]), ]


# Add points for stations, with Biome-based coloring
mp_all <- mp + 
  geom_point(data = MODIS_unique, 
             aes(x = lon, y = lat, color = factor(Biome)),  # Color mapped correctly
             size = 0.3, shape = 15) +
  coord_cartesian(ylim = range(MODIS.df$lat), 
                  xlim = range(MODIS.df$lon)) +
  xlab("Longitude") + 
  ylab("Latitude") +
  theme_classic() +
  theme(legend.position = c(0.05, 0.25)) +
  #scale_color_manual(values = c("#e41a1c","#377eb8","#4daf4a","#984ea3")) +  # Optional: Custom biome colors
  labs(color = "Biome Type")  # Legend label

# Save plot as PDF
ggsave("AllLocations_biomes.pdf", plot = mp_all, path = figure.dir, width = 16, height = 4)

# Display the map
mp_all


# Add points for stations, with PFT-based coloring
mp_PFT <- mp + 
  geom_point(data = MODIS_unique, 
             aes(x = lon, y = lat, color = factor(LC)),  # Color mapped correctly
             size = 0.3, shape = 15) +
  coord_cartesian(ylim = range(MODIS.df$lat), 
                  xlim = range(MODIS.df$lon)) +
  xlab("Longitude") + 
  ylab("Latitude") +
  theme_classic() +
  theme(legend.position = c(0.05, 0.25)) +
  scale_color_manual(values = c("#4daf4a","#377eb8")) +  # Optional: Custom biome colors
  labs(color = "PFT")  # Legend label

# Save plot as PDF
ggsave("AllLocations_PFT.pdf", plot = mp_PFT, path = figure.dir, width = 16, height = 4)

# Display the map
mp_PFT


# Add points for stations, with forest type-based coloring
mp_type <- mp + 
  geom_point(data = MODIS_unique, 
             aes(x = lon, y = lat, color = ForestType),  # Color mapped correctly
             size = 0.3, shape = 15) +
  coord_cartesian(ylim = range(MODIS.df$lat), 
                  xlim = range(MODIS.df$lon)) +
  xlab("Longitude") + 
  ylab("Latitude") +
  theme_classic() +
  theme(legend.position = c(0.08, 0.25)) +
  scale_color_manual(values = c("#e41a1c","#377eb8","#4daf4a")) +  # Optional: Custom biome colors
  labs(color = "PFT")  # Legend label

# Save plot as PDF
ggsave("AllLocations_forest_type.pdf", plot = mp_type, path = figure.dir, width = 16, height = 4)

# Display the map
mp_type
```
</details>

<br>

## 6. Safe final tables

<details><summary>show code</summary>
```{r}
# Order and select relevant columns
MODIS.df = MODIS.df %>% 
  dplyr::select(year, s_id, lat, lon, Biome, LC, ForestType,
                Greenup, Greenup_QA, MidGreenup, MidGreenup_QA,
                Greenup_prev, MidGreenup_prev, Senescence_prev, MidGreendown_prev)

# Print the first row to check
MODIS.df[1,]

# Save the cleaned MODIS table
write.csv(MODIS.df, paste(MODISout.dir, "MODIS_data_clean.csv", sep="/"), row.names = FALSE)  

# Extract unique coordinates
MODIS_data_coordinates = MODIS.df %>% 
  distinct(s_id, lat, lon, .keep_all = F)

# Save the unique coordinates table
write.csv(MODIS_data_coordinates, paste(MODISout.dir, "MODIS_coordinates.csv", sep="/"), row.names = FALSE)  
```
</details>

<br>

## 7. Reproducibility 

<details><summary>show code</summary>
```{r}
## datetime
Sys.time()

## session info
sessionInfo()
```
</details>