---
title: "Spring phenology tier2 project"
subtitle: "Run Bayesian hierarchical models (PEP725 data)"
author: "Constantin Zohner"
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

format:
  html:
    toc: false
    theme: default
    highlight-style: haddock
    df-print: paged
    embed-resources: true

execute:
  #warning: false
  message: false
  echo: true
  output: true

always_allow_html: true
editor: visual
editor_options:
  chunk_output_type: inline
---

<br>

### Driver names

-   leaf_out_prev: Leaf-out date in the previous year
-   LSF2: Binary indicator of late spring frost event (\<-2°C) after leaf-out in previous year
-   leaf_off: Leaf senescence date in the previous year
-   Tmean: Daily mean air temperature (°C):
    -   Previous year:
        -   EQ1 = March equinox
        -   SO1 = Summer solstice
        -   EQ2 = September equinox
        -   SO2 = December solstice
    -   From 1-17: 1 = January of previous year, 13 = January of current year
-   RH: Spring relative air humidity (%)
-   SWrad: Spring (preseason) incoming short-wave solar radiation (W/m²)
-   Tmax: Spring (preseason) Daily maximum air temperature (°C)

<br>

# 1. Setup

<details>

<summary>Packages</summary>

```{r, message=F}
library(tidyverse)
library(data.table)
library(car)
library(brms)
#install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
#cmdstanr::install_cmdstan()
library(cmdstanr)
options(brms.backend = "cmdstanr")
```

</details>

<details>

<summary>Directories</summary>

```{r}
#-------------------------------------------------
# Define the base directory
#-------------------------------------------------

base.dir = "/Users/goalie/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis"


#-------------------------------------------------
# Build specific directories from the base
#-------------------------------------------------

# PEP data
PEP.dir = file.path(base.dir, "Analysis_input/PEP_drivers_final_spring/Merged_file")

# Output
output.dir = file.path(base.dir, "Analysis_output/ModelResults/Bayesian_models")
```

</details>

<details>

<summary>Data</summary>

```{r}
#################
## Import data ##
#################



#-----------------------------------------------------------
# define model variables
#-----------------------------------------------------------

full_variables = c("year", "leaf_out", "leaf_out_prev", "leaf_off",
                   "LSF2",
                   "Tmean.EQ1.SO1", "Tmean.SO1.EQ2", "Tmean.EQ2.SO2",
                   "Tmax","RH","SWrad")

monthly_variables = c("Tmean.8","Tmean.9","Tmean.10","Tmean.11","Tmean.12",
                      "Tmean.13","Tmean.14","Tmean.15","Tmean.16","Tmean.17")


#-----------------------------------------------------------
# Load full data
#-----------------------------------------------------------

PEP.df <- fread(paste(PEP.dir, "pep_drivers_spring_data_clean_preseason.csv", sep="/")) %>% 
  dplyr::select(species, year, timeseries, lat, lon, alt_dem, 
                all_of(full_variables),
                all_of(monthly_variables)) %>% 
  mutate(leaf_out_DOY = leaf_out,
         period = "full")


#-----------------------------------------------------------
# Reduced data including all leaf-off observations
#-----------------------------------------------------------

PEPred.df <- PEP.df %>% 
  filter(!is.na(leaf_off)) 



##############################################################################################################################################
##############################################################################################################################################



####################
## Filter periods ##
####################



#-----------------------------------------------------------
# Define periods
#-----------------------------------------------------------

min.year.full = min(PEP.df$year) # Earliest year
min.year.full
max.year.full = max(PEP.df$year) # Latest year
max.year.full
year.center.full = round(median(c(min.year.full, max.year.full))) # central year
year.center.full
year.center.full-(min.year.full-1) #length of each period
max.year.full-year.center.full


#-----------------------------------------------------------
# Early data full
#-----------------------------------------------------------

PEP_early.df <- PEP.df %>% 
  mutate(period = "early") %>% 
  filter(!alt_dem>600,
         year>=min.year.full, 
         year<=year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()


#-----------------------------------------------------------
# Early data reduced
#-----------------------------------------------------------

PEPred_early.df <- PEPred.df %>% 
  mutate(period = "early") %>% 
  filter(!alt_dem>600,
         year>=min.year.full, 
         year<=year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()


#-----------------------------------------------------------
# Late data full
#-----------------------------------------------------------

PEP_late.df <- PEP.df %>% 
  mutate(period = "late") %>% 
  filter(!alt_dem>600,
         year>year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()


#-----------------------------------------------------------
# Late data reduced
#-----------------------------------------------------------

PEPred_late.df <- PEPred.df %>% 
  mutate(period = "late") %>% 
  filter(!alt_dem>600,
         year>year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()



##############################################################################################################################################
##############################################################################################################################################



###############################
## Merge early and late data ##
###############################



#-----------------------------------------------------------
# Reduced data
#-----------------------------------------------------------

PEPred.df = bind_rows(PEPred.df, PEPred_early.df, PEPred_late.df) %>% 
  filter(!species %in% c("Larix decidua"))
# Clean up
rm(PEPred_early.df, PEPred_late.df)


#-----------------------------------------------------------
# Full data
#-----------------------------------------------------------

PEP.df = bind_rows(PEP.df, PEP_early.df, PEP_late.df) %>% 
  # Remove all species that are already present in PEPred.df 
  # ensures that species with valid 'leaf_off' data are handled exclusively via PEPred.df
  filter(!species %in% unique(PEPred.df$species))
# Clean up
rm(PEP_early.df, PEP_late.df)


#-----------------------------------------------------------
# Merge and scale continuous variables
#-----------------------------------------------------------

# Identify continuous variables to be scaled (excluding LSF2 and monthly variables)
scaled_vars <- setdiff(full_variables, c("LSF2"))

PEP.df = bind_rows(PEP.df, PEPred.df) %>%

  # Group by observation period and site (i.e., timeseries = unique site-location combination)
  group_by(period, timeseries) %>%

  # Scale each continuous predictor within each site and period group
  # This removes across-site variation in mean/variance,
  # ensuring that model coefficients reflect *within-site* effects over time
  #
  # → Why is this important?
  # Because we're interested in whether *interannual variation* in predictors
  # (e.g., temperature, RH) explains *interannual shifts* in phenology,
  # not whether warmer sites leaf out earlier on average.
  #
  # Without this grouping, variables would be scaled globally,
  # and site-level means could dominate effect estimates —
  # falsely inflating signals driven by spatial gradients rather than temporal dynamics.
  #
  # Example: If warm lowland sites always have earlier leaf-out, a global
  # correlation between Tmean and leaf_out could appear strong,
  # even if within-site interannual correlations are weak or inconsistent.

  mutate(across(all_of(scaled_vars), ~ as.numeric(scale(.x, center = T, scale = T)))) %>%
  
  # Site-level anomalies of variables for monthly models (centered but not scaled)
  mutate(across(all_of(monthly_variables), ~ as.numeric(scale(.x, center = T, scale = F)))) %>%
  mutate(leaf_out_anomaly = scale(leaf_out_DOY, center = T, scale = F)) %>%

  # Remove group structure
  ungroup()

# Clean up
rm(PEPred.df)
```

</details>

<br>

# 2. Check multicollinearity using variance inflation factors (VIFs)

<details>

<summary>Full models</summary>

```{r}
#==================================================
# FULL MODELS — Includes all general predictors
# Adds 'leaf_off' if available for the species
#==================================================

VIF_full.df <- PEP.df %>%
  filter(period != "full") %>% 
  group_by(species) %>%
  group_modify(~ {
    data_subset <- .x
    
    # Determine if 'leaf_off' is usable for this species
    include_leaf_off <- any(!is.na(data_subset$leaf_off))
    
    # Conditionally construct model formula
    formula <- if (include_leaf_off) {
      leaf_out ~ leaf_out_prev + LSF2 + leaf_off + 
        Tmean.EQ1.SO1 + Tmean.SO1.EQ2 + Tmean.EQ2.SO2 +
        Tmax + RH + SWrad + year
    } else {
      leaf_out ~ leaf_out_prev + LSF2 +
        Tmean.EQ1.SO1 + Tmean.SO1.EQ2 + Tmean.EQ2.SO2 +
        Tmax + RH + SWrad + year
    }
    
    #--------------------------------------------------
    # Fit model and compute VIFs
    #--------------------------------------------------
    model <- lm(formula, data = data_subset)
    vif_vals <- vif(model)
    
    # Return tidy VIF output
    data.frame(
      Variable = names(vif_vals),
      VIF = as.numeric(vif_vals)
    )
  }) %>%
  ungroup() %>%
  
  # Reshape into wide format: each species gets its own column
  pivot_wider(names_from = species, values_from = VIF)

# View resulting wide-format VIF table
VIF_full.df
```

</details>

<details>

<summary>Monthly models</summary>

```{r}
#==================================================
# MONTHLY MODELS — Only include monthly temperature
# predictors up to the phenologically relevant month
#==================================================

VIF_monthly.df <- PEP.df %>%
  filter(period == "full") %>% 
  group_by(species) %>%
  group_map(~ {
    
    data_subset <- .x                  # Subset of data for the current species
    species_name <- .y$species         # Current species name
    
    # Skip if too few observations
    if (nrow(data_subset) < 30) return(NULL)
    
    #--------------------------------------------------
    # Determine latest Tmean month to include
    #--------------------------------------------------
    mean_doy <- mean(data_subset$leaf_out_DOY, na.rm = TRUE)
    
    # Estimate approximate month from DOY (1 = Jan, ..., 5 = May)
    latest_month <- floor((mean_doy - 1) / 30) + 1
    
    # Only include a month if leaf-out occurs in its second half
    if (mean_doy < ((latest_month - 1) * 30 + 15)) {
      latest_month <- latest_month - 1  # Exclude month if leaf-out too early
    }
    
    # Translate to Tmean variable names: Tmean.8 = August previous year
    latest_tmean <- min(12 + latest_month, 17)
    tmean_vars <- if (latest_tmean >= 8) paste0("Tmean.", 8:latest_tmean) else character(0)
    
    # Include only if we have at least one Tmean variable
    predictors <- c(tmean_vars, "year")
    if (length(predictors) == 0) return(NULL)
    
    #--------------------------------------------------
    # Fit model and compute VIFs
    #--------------------------------------------------
    formula <- as.formula(paste("leaf_out_anomaly ~", paste(predictors, collapse = " + ")))
    
    model <- tryCatch(lm(formula, data = data_subset), error = function(e) return(NULL))
    if (is.null(model)) return(NULL)
    
    vif_vals <- tryCatch(vif(model), error = function(e) return(NULL))
    if (is.null(vif_vals) || length(vif_vals) == 0) return(NULL)
    
    # Return tidy output with species column
    data.frame(
      Variable = names(vif_vals),
      VIF = as.numeric(vif_vals),
      species = species_name
    )
  }) %>%
  bind_rows() %>%                     # Combine species-wise results into one data frame
  pivot_wider(                        # Reshape: one column per species
    names_from = species,
    values_from = VIF
  )

# View resulting wide-format VIF table
VIF_monthly.df
```

</details>

<br>

# 3. Run Bayesian hierarchical models

<details>

<summary>Full models</summary>

```{r}
#-----------------------------------------------------------
# Define priors
#-----------------------------------------------------------

priors_brm <- c(
  # Fixed effects — weakly informative (small shrinkage)
  set_prior("normal(0, 1)", class = "b"), 
  
  # Intercept — mild regularization (helps model intercept stability)
  set_prior("normal(0, 1)", class = "Intercept"), 
  
  # Residual SD (sigma):
  set_prior("student_t(3, 0, 1)", class = "sigma"),
  
  # General SDs for random effects — conservative, heavy tails
  set_prior("student_t(3, 0, 2.5)", class = "sd"), 
  
  # Specific shrinkage for random slope of year by timeseries
  set_prior("normal(0, 0.5)", class = "sd", group = "timeseries", coef = "year")
)

#-----------------------------------------------------------
# Define modeling formulae
#-----------------------------------------------------------

# To account for temporal trends in phenology and potential non-independence of observations across years within each site, 
# we included year as a random slope grouped by timeseries using an uncorrelated random effects structure (year || timeseries). 
# This formulation allows both the baseline phenology (intercept) and the temporal trend (slope of year) to vary across sites, 
# while simplifying model complexity by not estimating the correlation between them.
#
# By doing so, we capture spatial heterogeneity in long-term phenological trends and reduce the risk of inflated fixed-effect 
# estimates caused by unmodeled site-specific trajectories. This specification preserves our ability to detect interannual 
# relationships between climate predictors and leaf-out timing within sites, while controlling for varying temporal baselines 
# across locations in a computationally efficient and statistically robust manner.

# Model without leaf_off (used if no data for that variable)
formula_nolo <- bf(
  leaf_out ~ leaf_out_prev + LSF2 + Tmean.EQ1.SO1 + Tmean.SO1.EQ2 + 
    Tmean.EQ2.SO2 + Tmax + RH + SWrad + (year || timeseries)
)

# Model including leaf_off (used if data is available for that species)
formula_lo <- bf(
  leaf_out ~ leaf_out_prev + LSF2 + Tmean.EQ1.SO1 + Tmean.SO1.EQ2 + leaf_off + 
    Tmean.EQ2.SO2 + Tmax + RH + SWrad + (year || timeseries)
)

#-----------------------------------------------------------
# Setup: model loop across phenological periods and species
#-----------------------------------------------------------

periods <- c("early", "late")          # Define observation periods
species_list <- unique(PEP.df$species) # Extract species names
models <- list()                       # Create empty list to store models

#-----------------------------------------------------------
# Loop over each period × species combination
#-----------------------------------------------------------

for (p in periods) {
  models[[p]] <- list()                # Create sublist for each period
  
  for (sp in species_list) {
    cat("Running model for Period:", p, "| Species:", sp, "\n")
    
    # Subset the data for this species and period
    data_sub <- PEP.df %>% filter(period == p, species == sp)
    
    # Skip if there are too few observations for stable estimation
    if (nrow(data_sub) < 30) {
      cat("Skipping species due to insufficient data:", sp, "\n")
      next
    }
    
    #-----------------------------------------------------------
    # Use different formula depending on whether 'leaf_off' is available
    #-----------------------------------------------------------
    
    include_leaf_off <- any(!is.na(data_sub$leaf_off))
    formula <- if (include_leaf_off) formula_lo else formula_nolo
    
    #-----------------------------------------------------------
    # Fit the model using brms and CmdStanR backend
    #-----------------------------------------------------------
    
    models[[p]][[sp]] <- brm(
      formula = formula,               # The model formula, chosen dynamically depending on whether 'leaf_off' is available
      data = data_sub,                 # Subset of data for this species and period
      family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
      prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
      
      backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
      #init = 0,                        # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
      
      chains = 4,                      # Number of MCMC chains to run (default is 4)
      cores = 12,                      # Number of CPU cores used in total
      threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
      
      iter = 4000,                     # Total number of iterations per chain (including warmup)
      warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
      
      control = list(
        adapt_delta = 0.95,            # Target acceptance probability for the sampler (higher = fewer divergences)
        max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
      ),
      
      seed = 123                       # Set random seed for reproducibility
    )
  }
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_full_models.rds"))
```

</details>

<details>

<summary>Monthly models</summary>

```{r}
#-----------------------------------------------------------
# Calculate average leaf-out DOY per species (across all years/periods)
# Used to determine which monthly temperature predictors to include
#-----------------------------------------------------------

species_leaf_out = PEP.df %>% 
  filter(period == "full") %>% 
  group_by(species) %>% 
  summarise(mean_leaf_out = mean(leaf_out_DOY, na.rm = TRUE)) %>%
  arrange(mean_leaf_out)
species_leaf_out

#-----------------------------------------------------------
# Define priors
#-----------------------------------------------------------

priors_brm <- c(
  # Fixed effects (slopes, in days/°C):
  # weakly informative (small shrinkage)
  # One shared prior for all T predictors. SD = 1.5 comes from:
  #   target ≈ 3 days change per 1-SD(x) and median SD(x) ≈ 2 °C ⇒ 3/2 ≈ 1.5 days/°C.
  # This is weakly-informative: it comfortably covers ~±5 days/°C (≈ ±3.3 SDs),
  # while gently shrinking implausibly large effects given sd_y ≈ 9 days.
  set_prior("normal(0, 1.5)", class = "b"), 
  
  # Intercept — mild regularization (helps model intercept stability)
  set_prior("normal(0, 2)", class = "Intercept"), 
  
  # Residual SD (sigma):
  set_prior("student_t(3, 0, 3)", class = "sigma"),
  
  # General SDs for random effects — conservative, heavy tails
  set_prior("student_t(3, 0, 3)", class = "sd"), 
  
  # Specific shrinkage for random slope of year by timeseries
  set_prior("normal(0, 0.5)", class = "sd", group = "timeseries", coef = "year")
)

#-----------------------------------------------------------
# Setup model structure
#-----------------------------------------------------------

periods <- c("full")
species_list <- unique(PEP.df$species)
models <- list()

#-----------------------------------------------------------
# Loop over each period × species combination
#-----------------------------------------------------------

for (p in periods) {
  models[[p]] <- list()
  
  for (sp in species_list) {
    cat("Running model for Period:", p, "| Species:", sp, "\n")
    
    # Subset data for this species and period
    data_sub <- PEP.df %>% filter(period == p, species == sp)
    
    # Skip model fitting if insufficient data
    if (nrow(data_sub) < 30) {
      cat("Skipping species due to insufficient data:", sp, "\n")
      next
    }
    
    #-----------------------------------------------------------
    # Select monthly temperature predictors based on leaf-out timing
    #-----------------------------------------------------------
    
    # Get species-specific average leaf-out day
    mean_doy <- species_leaf_out[species_leaf_out$species==sp,]$mean_leaf_out
    
    # Convert DOY to approximate month (1 = Jan, 5 = May, etc.)
    latest_month <- floor((mean_doy - 1) / 30) + 1
    
    # Only include monthly temperature if leaf-out occurs in the second half of that month
    if (mean_doy < ((latest_month - 1) * 30 + 15)) {
      latest_month <- latest_month - 1  # Too early → exclude current month
    }
    
    # Convert calendar month to Tmean.X format (Tmean.8 = August prev year)
    latest_tmean <- 12 + latest_month  # e.g., March → Tmean.15
    
    # Dynamically build Tmean term string (e.g., Tmean.8 + Tmean.9 + ...)
    tmean_terms <- paste0("Tmean.", 8:latest_tmean, collapse = " + ")
    
    # Build model formula string including only relevant months + uncorrelated random slopes for year by timeseries
    formula_string <- paste("leaf_out_anomaly ~", tmean_terms, "+ (year || timeseries)")
    formula <- bf(as.formula(formula_string)) # Convert to brms formula
    
    #-----------------------------------------------------------
    # Fit the Bayesian model
    #-----------------------------------------------------------
    
    models[[p]][[sp]] <- brm(
      formula = formula,               # The model formula, chosen dynamically
      data = data_sub,                 # Subset of data for this species and period
      family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
      prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
      
      backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
      #init = 0,                        # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
      
      chains = 4,                      # Number of MCMC chains to run (default is 4)
      cores = 12,                      # Number of CPU cores used in total
      threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
      
      iter = 4000,                     # Total number of iterations per chain (including warmup)
      warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
      
      control = list(
        adapt_delta = 0.95,            # Target acceptance probability for the sampler (higher = fewer divergences)
        max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
      ),
      
      seed = 123                       # Set random seed for reproducibility
    )
  }
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_monthly_models.rds"))
```

</details>

<br>

# 4. Reproducibility

<details>

<summary>show</summary>

```{r}
# Datetime
Sys.time()

# Session info
sessionInfo()
```

</details>
