---
title: "Spring phenology tier2 project"
subtitle: "Run Bayesian hierarchical models using student distribution and centered response(PEP725 data)"
author: "Constantin Zohner"
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

format:
  html:
    toc: false
    theme: default
    highlight-style: haddock
    df-print: paged
    embed-resources: true

execute:
  #warning: false
  message: false
  echo: true
  output: true

always_allow_html: true
editor: visual
editor_options:
  chunk_output_type: inline
---

<br>

### Methods

To isolate interannual, within-site drivers of spring phenology while avoiding spurious associations driven by shared temporal trends or between-site differences, all continuous climate predictors were centered and scaled within site, such that coefficients reflect responses to typical site-level anomalies. Green-up dates were centered within site to remove differences in mean phenological timing among sites.

In the main analyses (see 3.3_Bayesian_models_MODIS.html), green-up anomalies were additionally scaled within site, allowing effect sizes to be interpreted as relative sensitivities (in units of within-site phenological variability) and ensuring equal weighting of sites when comparing the relative importance of climate drivers.

As a robustness check, we refitted all models using unscaled (centered-only) green-up anomalies and a Student-t error distribution, which down-weights occasional extreme observations and provides inference in original phenological units (days). The consistency of results across these complementary model formulations indicates that our conclusions are not driven by site-specific differences in phenological variance or by a small number of anomalous observations.

### Driver names

-   leaf_out_prev: Leaf-out date in the previous year
-   LSF2: Binary indicator of late spring frost event (\<-2°C) after leaf-out in previous year
-   leaf_off: Leaf senescence date in the previous year
-   Tmean: Daily mean air temperature (°C):
    -   Previous year:
        -   EQ1 = March equinox
        -   SO1 = Summer solstice
        -   EQ2 = September equinox
        -   SO2 = December solstice
    -   From 1-17: 1 = January of previous year, 13 = January of current year
-   RH: Spring relative air humidity (%)
-   SWrad: Spring (preseason) incoming short-wave solar radiation (W/m²)
-   Tmax: Spring (preseason) Daily maximum air temperature (°C)

<br>

# 1. Setup

<details>

<summary>Packages</summary>

```{r, message=F}
library(tidyverse)
library(data.table)
library(car)
library(brms)
#install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
#cmdstanr::install_cmdstan()
library(cmdstanr)
options(brms.backend = "cmdstanr")
```

</details>

<details>

<summary>Directories</summary>

```{r}
#-------------------------------------------------
# Define the base directory
#-------------------------------------------------

base.dir = "/Users/goalie/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis"


#-------------------------------------------------
# Build specific directories from the base
#-------------------------------------------------

# PEP data
PEP.dir = file.path(base.dir, "Analysis_input/PEP_drivers_final_spring/Merged_file")

# Output
output.dir = file.path(base.dir, "Analysis_output/ModelResults/Bayesian_models")
```

</details>

<details>

<summary>Data</summary>

```{r}
#################
## Import data ##
#################



#-----------------------------------------------------------
# define model variables
#-----------------------------------------------------------

# Response
response_vars = c("leaf_out")

# List of full variables
pred_vars = c("year", 
              "leaf_out_prev", "leaf_off",
              "Tmean.EQ1.SO1", "Tmean.SO1.EQ2", "Tmean.EQ2.SO2",
              "Tmax","RH","SWrad")


#-----------------------------------------------------------
# Load full data
#-----------------------------------------------------------

PEP.df <- fread(paste(PEP.dir, "pep_drivers_spring_data_clean_preseason.csv", sep="/")) %>% 
  dplyr::select(species, year, timeseries, lat, lon, alt_dem, 
                LSF2,
                all_of(c(response_vars,pred_vars))) %>% 
  mutate(period = "full")


#-----------------------------------------------------------
# Reduced data including all leaf-off observations
#-----------------------------------------------------------

PEPred.df <- PEP.df %>% 
  filter(!is.na(leaf_off)) 



##############################################################################################################################################
##############################################################################################################################################



####################
## Filter periods ##
####################



#-----------------------------------------------------------
# Define periods
#-----------------------------------------------------------

min.year.full = min(PEP.df$year) # Earliest year
min.year.full
max.year.full = max(PEP.df$year) # Latest year
max.year.full
year.center.full = round(median(c(min.year.full, max.year.full))) # central year
year.center.full
year.center.full-(min.year.full-1) #length of each period
max.year.full-year.center.full


#-----------------------------------------------------------
# Early data full
#-----------------------------------------------------------

PEP_early.df <- PEP.df %>% 
  mutate(period = "early") %>% 
  filter(!alt_dem>600,
         year>=min.year.full, 
         year<=year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()


#-----------------------------------------------------------
# Early data reduced
#-----------------------------------------------------------

PEPred_early.df <- PEPred.df %>% 
  mutate(period = "early") %>% 
  filter(!alt_dem>600,
         year>=min.year.full, 
         year<=year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()


#-----------------------------------------------------------
# Late data full
#-----------------------------------------------------------

PEP_late.df <- PEP.df %>% 
  mutate(period = "late") %>% 
  filter(!alt_dem>600,
         year>year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()


#-----------------------------------------------------------
# Late data reduced
#-----------------------------------------------------------

PEPred_late.df <- PEPred.df %>% 
  mutate(period = "late") %>% 
  filter(!alt_dem>600,
         year>year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()



##############################################################################################################################################
##############################################################################################################################################



###############################
## Merge early and late data ##
###############################



#-----------------------------------------------------------
# Reduced data
#-----------------------------------------------------------

PEPred.df = bind_rows(PEPred.df, PEPred_early.df, PEPred_late.df) %>% 
  filter(!species %in% c("Larix decidua"))
# Clean up
rm(PEPred_early.df, PEPred_late.df)


#-----------------------------------------------------------
# Full data
#-----------------------------------------------------------

PEP.df = bind_rows(PEP.df, PEP_early.df, PEP_late.df) %>% 
  # Remove all species that are already present in PEPred.df 
  # ensures that species with valid 'leaf_off' data are handled exclusively via PEPred.df
  filter(!species %in% unique(PEPred.df$species))
# Clean up
rm(PEP_early.df, PEP_late.df)


#-----------------------------------------------------------
# Merge and scale continuous variables
#-----------------------------------------------------------

PEP.df = bind_rows(PEP.df, PEPred.df) %>%

  # Group by observation period and site (i.e., timeseries = unique site-location combination)
  group_by(period, timeseries) %>%
  
  # Scale each continuous predictor within each pixel
  # This removes across-site variation in mean/variance,
  # ensuring that model coefficients reflect within-site effects over time
  #
  # → Why is this important?
  # Because we're interested in whether interannual variation in predictors
  # (e.g., temperature, RH) explains interannual shifts in phenology,
  # not whether warmer sites leaf out earlier on average.
  
  # In simple or idealized cases, a random-intercept model may effectively estimate 
  # a within-site effect even without centering. However, in more complex settings—especially
  # when predictors differ systematically across groups or are constructed differently across
  # sites—this is no longer guaranteed, and explicit within–between decomposition becomes 
  # necessary for clear interpretation.

  # Site-level anomalies of response variable (centered but not scaled)
  mutate(
    across(
      all_of(response_vars), 
      ~ .x - mean(.x, na.rm = TRUE),
      .names = "{.col}_anom")) %>%
  
  # Site-level scaling of predictor variables
  mutate(
    across(
      all_of(pred_vars), 
      ~ as.numeric(scale(.x, center = TRUE, scale = TRUE)),
      .names = "{.col}_z"))
  
# Clean up
rm(PEPred.df)
```

</details>

<br>

# 2. Check multicollinearity using variance inflation factors (VIFs)

<details>

<summary>Full models</summary>

```{r}
#==================================================
# FULL MODELS — Includes all general predictors
# Adds 'leaf_off' if available for the species
#==================================================

VIF_full.df <- PEP.df %>%
  filter(period != "full") %>% 
  group_by(species) %>%
  group_modify(~ {
    data_subset <- .x
    
    # Determine if 'leaf_off' is usable for this species
    include_leaf_off <- any(!is.na(data_subset$leaf_off_z))
    
    # Conditionally construct model formula
    formula <- if (include_leaf_off) {
      leaf_out_anom ~ leaf_out_prev_z + LSF2 + leaf_off_z + 
        Tmean.EQ1.SO1_z + Tmean.SO1.EQ2_z + Tmean.EQ2.SO2_z +
        Tmax_z + RH_z + SWrad_z + year_z
    } else {
      leaf_out_anom ~ leaf_out_prev_z + LSF2 +
        Tmean.EQ1.SO1_z + Tmean.SO1.EQ2_z + Tmean.EQ2.SO2_z +
        Tmax_z + RH_z + SWrad_z + year_z
    }
    
    #--------------------------------------------------
    # Fit model and compute VIFs
    #--------------------------------------------------
    model <- lm(formula, data = data_subset)
    vif_vals <- vif(model)
    
    # Return tidy VIF output
    data.frame(
      Variable = names(vif_vals),
      VIF = as.numeric(vif_vals)
    )
  }) %>%
  ungroup() %>%
  
  # Reshape into wide format: each species gets its own column
  pivot_wider(names_from = species, values_from = VIF)

# View resulting wide-format VIF table
VIF_full.df
```

</details>

<br>

# 3. Run Bayesian hierarchical models

<details>

<summary>Full models</summary>

```{r}
#-----------------------------------------------------------
# Define priors
#-----------------------------------------------------------

priors_brm <- c(
  # Fixed effects — weakly informative (small shrinkage)
  set_prior("normal(0, 1)", class = "b"), 
  
  # Intercept — mild regularization (helps model intercept stability)
  set_prior("normal(0, 1)", class = "Intercept"), 
  
  # Residual SD (sigma):
  set_prior("student_t(3, 0, 1)", class = "sigma"),
  
  # General SDs for random effects — conservative, heavy tails
  set_prior("student_t(3, 0, 2.5)", class = "sd"), 
  
  # Specific shrinkage for random slope of year by timeseries
  set_prior("normal(0, 0.5)", class = "sd", group = "timeseries", coef = "year_z"),
  
  # Degrees of freedom for Student-t (controls tail heaviness)
  set_prior("gamma(2, 0.1)", class = "nu")
)

#-----------------------------------------------------------
# Define modeling formulae
#-----------------------------------------------------------

# To account for temporal trends in phenology and potential non-independence of observations across years within each site, 
# we included year as a random slope grouped by timeseries using an uncorrelated random effects structure (year || timeseries). 
# This formulation allows both the baseline phenology (intercept) and the temporal trend (slope of year) to vary across sites, 
# while simplifying model complexity by not estimating the correlation between them.
#
# By doing so, we capture spatial heterogeneity in long-term phenological trends and reduce the risk of inflated fixed-effect 
# estimates caused by unmodeled site-specific trajectories. This specification preserves our ability to detect interannual 
# relationships between climate predictors and leaf-out timing within sites, while controlling for varying temporal baselines 
# across locations in a computationally efficient and statistically robust manner.

# Model without leaf_off (used if no data for that variable)
formula_nolo <- bf(
  leaf_out_anom ~ leaf_out_prev_z + LSF2 + Tmean.EQ1.SO1_z + Tmean.SO1.EQ2_z + 
    Tmean.EQ2.SO2_z + Tmax_z + RH_z + SWrad_z + (year_z || timeseries)
)

# Model including leaf_off (used if data is available for that species)
formula_lo <- bf(
  leaf_out_anom ~ leaf_out_prev_z + LSF2 + Tmean.EQ1.SO1_z + Tmean.SO1.EQ2_z + leaf_off_z + 
    Tmean.EQ2.SO2_z + Tmax_z + RH_z + SWrad_z + (year_z || timeseries)
)

#-----------------------------------------------------------
# Setup: model loop across phenological periods and species
#-----------------------------------------------------------

periods <- c("early", "late")          # Define observation periods
species_list <- unique(PEP.df$species) # Extract species names
models <- list()                       # Create empty list to store models

#-----------------------------------------------------------
# Loop over each period × species combination
#-----------------------------------------------------------

for (p in periods) {
  models[[p]] <- list()                # Create sublist for each period
  
  for (sp in species_list) {
    cat("Running model for Period:", p, "| Species:", sp, "\n")
    
    # Subset the data for this species and period
    data_sub <- PEP.df %>% filter(period == p, species == sp)
    
    # Skip if there are too few observations for stable estimation
    if (nrow(data_sub) < 30) {
      cat("Skipping species due to insufficient data:", sp, "\n")
      next
    }
    
    #-----------------------------------------------------------
    # Use different formula depending on whether 'leaf_off' is available
    #-----------------------------------------------------------
    
    include_leaf_off <- any(!is.na(data_sub$leaf_off_z))
    formula <- if (include_leaf_off) formula_lo else formula_nolo
    
    #-----------------------------------------------------------
    # Fit the model using brms and CmdStanR backend
    #-----------------------------------------------------------
    
    models[[p]][[sp]] <- brm(
      formula = formula,               # The model formula, chosen dynamically depending on whether 'leaf_off' is available
      data = data_sub,                 # Subset of data for this species and period
      #family = gaussian(),            # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
      family  = student(),             # ← THIS IS THE KEY CHANGE
      prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
      
      backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
      #init = 0,                       # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
      
      chains = 4,                      # Number of MCMC chains to run (default is 4)
      cores = 12,                      # Number of CPU cores used in total
      threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
      
      iter = 4000,                     # Total number of iterations per chain (including warmup)
      warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
      
      control = list(
        adapt_delta = 0.95,            # Target acceptance probability for the sampler (higher = fewer divergences)
        max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
      ),
      
      seed = 123                       # Set random seed for reproducibility
    )
  }
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_full_models_studentCenter.rds"))
```

</details>

<br>

# 4. Reproducibility

<details>

<summary>show</summary>

```{r}
# Datetime
Sys.time()

# Session info
sessionInfo()
```

</details>
