---
title: "Spring phenology tier2 project"
subtitle: "Run Bayesian hierarchical models using alternate formulations (PEP725 data)"
author: "Constantin Zohner"
date: "Last run: `r format(Sys.Date(), '%d %B %Y')`"

format:
  html:
    toc: false
    theme: default
    highlight-style: haddock
    df-print: paged
    embed-resources: true

execute:
  #warning: false
  message: false
  echo: true
  output: true

always_allow_html: true
editor: visual
editor_options:
  chunk_output_type: inline
---

<br>

### Methods

To illustrate how common modelling choices can influence inference about phenological drivers, we repeated the analyses under four simplified specifications. We (i) represented spring temperature using a fixed February–April window rather than an empirically identified preseason; (ii) fit models without within-site centering of climate predictors, thereby pooling spatial and temporal variation; and (iii) omitted the year term, reducing accommodation of interannual structure. These alternatives reflect approaches frequently used in phenological analyses. Across both ground and satellite data, relaxing these controls generally weakened the apparent dominance of spring temperature and increased the apparent importance of secondary or lagged predictors, consistent with signal dilution, spatial–temporal confounding, and overconfident inference when clustering is ignored.

### Driver names

-   leaf_out_prev: Leaf-out date in the previous year
-   LSF2: Binary indicator of late spring frost event (\<-2°C) after leaf-out in previous year
-   leaf_off: Leaf senescence date in the previous year
-   Tmean: Daily mean air temperature (°C):
    -   Previous year:
        -   EQ1 = March equinox
        -   SO1 = Summer solstice
        -   EQ2 = September equinox
        -   SO2 = December solstice
    -   From 1-17: 1 = January of previous year, 13 = January of current year
-   RH: Spring relative air humidity (%)
-   SWrad: Spring (preseason) incoming short-wave solar radiation (W/m²)
-   Tmax: Spring (preseason) Daily maximum air temperature (°C)

<br>

# 1. Setup

<details>

<summary>Packages</summary>

```{r, message=F}
library(tidyverse)
library(data.table)
library(car)
library(brms)
#install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
#cmdstanr::install_cmdstan()
library(cmdstanr)
options(brms.backend = "cmdstanr")
```

</details>

<details>

<summary>Directories</summary>

```{r}
#-------------------------------------------------
# Define the base directory
#-------------------------------------------------

base.dir = "/Users/goalie/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis"


#-------------------------------------------------
# Build specific directories from the base
#-------------------------------------------------

# PEP data
PEP.dir = file.path(base.dir, "Analysis_input/PEP_drivers_final_spring/Merged_file")

# Output
output.dir = file.path(base.dir, "Analysis_output/ModelResults/Bayesian_models")
```

</details>

<details>

<summary>Data</summary>

```{r}
#################
## Import data ##
#################



#-----------------------------------------------------------
# define model variables
#-----------------------------------------------------------

# Response
response_vars = c("leaf_out")

# List of full variables
pred_vars = c("year", 
              "leaf_out_prev", "leaf_off",
              "Tmean.EQ1.SO1", "Tmean.SO1.EQ2", "Tmean.EQ2.SO2",
              "Tmax","RH","SWrad")


#-----------------------------------------------------------
# Load full data
#-----------------------------------------------------------

PEP.df <- fread(paste(PEP.dir, "pep_drivers_spring_data_clean_preseason.csv", sep="/")) %>% 
  dplyr::select(species, year, timeseries, lat, lon, alt_dem, 
                LSF2, Tmax.14, Tmax.15, Tmax.16,
                all_of(c(response_vars,pred_vars))) %>% 
  mutate(period = "full")


#-----------------------------------------------------------
# Reduced data including all leaf-off observations
#-----------------------------------------------------------

PEPred.df <- PEP.df %>% 
  filter(!is.na(leaf_off)) 



##############################################################################################################################################
##############################################################################################################################################



####################
## Filter periods ##
####################



#-----------------------------------------------------------
# Define periods
#-----------------------------------------------------------

min.year.full = min(PEP.df$year) # Earliest year
min.year.full
max.year.full = max(PEP.df$year) # Latest year
max.year.full
year.center.full = round(median(c(min.year.full, max.year.full))) # central year
year.center.full
year.center.full-(min.year.full-1) #length of each period
max.year.full-year.center.full


#-----------------------------------------------------------
# Early data full
#-----------------------------------------------------------

PEP_early.df <- PEP.df %>% 
  mutate(period = "early") %>% 
  filter(!alt_dem>600,
         year>=min.year.full, 
         year<=year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()


#-----------------------------------------------------------
# Early data reduced
#-----------------------------------------------------------

PEPred_early.df <- PEPred.df %>% 
  mutate(period = "early") %>% 
  filter(!alt_dem>600,
         year>=min.year.full, 
         year<=year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()


#-----------------------------------------------------------
# Late data full
#-----------------------------------------------------------

PEP_late.df <- PEP.df %>% 
  mutate(period = "late") %>% 
  filter(!alt_dem>600,
         year>year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()


#-----------------------------------------------------------
# Late data reduced
#-----------------------------------------------------------

PEPred_late.df <- PEPred.df %>% 
  mutate(period = "late") %>% 
  filter(!alt_dem>600,
         year>year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()



##############################################################################################################################################
##############################################################################################################################################



###############################
## Merge early and late data ##
###############################



#-----------------------------------------------------------
# Reduced data
#-----------------------------------------------------------

PEPred.df = bind_rows(PEPred.df, PEPred_early.df, PEPred_late.df) %>% 
  filter(!species %in% c("Larix decidua"))
# Clean up
rm(PEPred_early.df, PEPred_late.df)


#-----------------------------------------------------------
# Full data
#-----------------------------------------------------------

PEP.df = bind_rows(PEP.df, PEP_early.df, PEP_late.df) %>% 
  # Remove all species that are already present in PEPred.df 
  # ensures that species with valid 'leaf_off' data are handled exclusively via PEPred.df
  filter(!species %in% unique(PEPred.df$species))
# Clean up
rm(PEP_early.df, PEP_late.df)


#-----------------------------------------------------------
# Merge and scale continuous variables
#-----------------------------------------------------------

PEP.df = bind_rows(PEP.df, PEPred.df) %>%

  # Add mean spring (January to March) temperatures
  mutate(Tmax_spring = rowMeans(
    select(., Tmax.14, Tmax.15, Tmax.16),
    na.rm = TRUE
  )) %>% 
  
  # Group by observation period and site (i.e., timeseries = unique site-location combination)
  group_by(period, timeseries) %>%
  
  # Scale each continuous predictor within each pixel
  # This removes across-site variation in mean/variance,
  # ensuring that model coefficients reflect within-site effects over time
  #
  # → Why is this important?
  # Because we're interested in whether interannual variation in predictors
  # (e.g., temperature, RH) explains interannual shifts in phenology,
  # not whether warmer sites leaf out earlier on average.
  
  # In simple or idealized cases, a random-intercept model may effectively estimate 
  # a within-site effect even without centering. However, in more complex settings—especially
  # when predictors differ systematically across groups or are constructed differently across
  # sites—this is no longer guaranteed, and explicit within–between decomposition becomes 
  # necessary for clear interpretation.

  # Site-level scaling of variable
  mutate(
    across(
      all_of(c(response_vars, pred_vars, "Tmax_spring")), 
      ~ as.numeric(scale(.x, center = TRUE, scale = TRUE)),
      .names = "{.col}_z")) %>%
   ungroup() %>% 
  
  # Global scaling of variables
  mutate(
    across(
      all_of(c(response_vars, pred_vars, "Tmax_spring")), 
      ~ as.numeric(scale(.x, center = TRUE, scale = TRUE)),
      .names = "{.col}_z_global"))
  
# Clean up
rm(PEPred.df)
```

</details>

<br>

# 2. Run Bayesian hierarchical models

<details>

<summary>No Spring model</summary>

```{r}
#-----------------------------------------------------------
# Define priors
#-----------------------------------------------------------

priors_brm <- c(
  # Fixed effects — weakly informative (small shrinkage)
  set_prior("normal(0, 1)", class = "b"), 
  
  # Intercept — mild regularization (helps model intercept stability)
  set_prior("normal(0, 1)", class = "Intercept"), 
  
  # Residual SD (sigma):
  set_prior("student_t(3, 0, 1)", class = "sigma"),
  
  # General SDs for random effects — conservative, heavy tails
  set_prior("student_t(3, 0, 2.5)", class = "sd"), 
  
  # Specific shrinkage for random slope of year by timeseries
  set_prior("normal(0, 0.5)", class = "sd", group = "timeseries", coef = "year_z")
)

#-----------------------------------------------------------
# Define modeling formulae
#-----------------------------------------------------------

# Model without leaf_off (used if no data for that variable)
formula_nolo <- bf(
  leaf_out_z ~ leaf_out_prev_z + LSF2 + Tmean.EQ1.SO1_z + Tmean.SO1.EQ2_z + 
    Tmean.EQ2.SO2_z + Tmax_spring_z + RH_z + SWrad_z + (year_z || timeseries)
)

# Model including leaf_off (used if data is available for that species)
formula_lo <- bf(
  leaf_out_z ~ leaf_out_prev_z + LSF2 + Tmean.EQ1.SO1_z + Tmean.SO1.EQ2_z + leaf_off_z + 
    Tmean.EQ2.SO2_z + Tmax_spring_z + RH_z + SWrad_z + (year_z || timeseries)
)

#-----------------------------------------------------------
# Setup: model loop across phenological periods and species
#-----------------------------------------------------------

periods <- c("early", "late")          # Define observation periods
species_list <- unique(PEP.df$species) # Extract species names
models <- list()                       # Create empty list to store models

#-----------------------------------------------------------
# Loop over each period × species combination
#-----------------------------------------------------------

for (p in periods) {
  models[[p]] <- list()                # Create sublist for each period
  
  for (sp in species_list) {
    cat("Running model for Period:", p, "| Species:", sp, "\n")
    
    # Subset the data for this species and period
    data_sub <- PEP.df %>% filter(period == p, species == sp)
    
    # Skip if there are too few observations for stable estimation
    if (nrow(data_sub) < 30) {
      cat("Skipping species due to insufficient data:", sp, "\n")
      next
    }
    
    #-----------------------------------------------------------
    # Use different formula depending on whether 'leaf_off' is available
    #-----------------------------------------------------------
    
    include_leaf_off <- any(!is.na(data_sub$leaf_off_z))
    formula <- if (include_leaf_off) formula_lo else formula_nolo
    
    #-----------------------------------------------------------
    # Fit the model using brms and CmdStanR backend
    #-----------------------------------------------------------
    
    models[[p]][[sp]] <- brm(
      formula = formula,               # The model formula, chosen dynamically depending on whether 'leaf_off' is available
      data = data_sub,                 # Subset of data for this species and period
      family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
      prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
      
      backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
      #init = 0,                       # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
      
      chains = 4,                      # Number of MCMC chains to run (default is 4)
      cores = 12,                      # Number of CPU cores used in total
      threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
      
      iter = 4000,                     # Total number of iterations per chain (including warmup)
      warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
      
      control = list(
        adapt_delta = 0.95,            # Target acceptance probability for the sampler (higher = fewer divergences)
        max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
      ),
      
      seed = 123                       # Set random seed for reproducibility
    )
  }
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_full_models_noSpring.rds"))
```

</details>

<details>

<summary>No Scale model</summary>

```{r}
#-----------------------------------------------------------
# Define priors
#-----------------------------------------------------------

priors_brm <- c(
  # Fixed effects — weakly informative (small shrinkage)
  set_prior("normal(0, 1)", class = "b"), 
  
  # Intercept — mild regularization (helps model intercept stability)
  set_prior("normal(0, 1)", class = "Intercept"), 
  
  # Residual SD (sigma):
  set_prior("student_t(3, 0, 1)", class = "sigma"),
  
  # General SDs for random effects — conservative, heavy tails
  set_prior("student_t(3, 0, 2.5)", class = "sd"), 
  
  # Specific shrinkage for random slope of year by timeseries
  set_prior("normal(0, 0.5)", class = "sd", group = "timeseries", coef = "year_z_global")
)

#-----------------------------------------------------------
# Define modeling formulae
#-----------------------------------------------------------

# Model without leaf_off (used if no data for that variable)
formula_nolo <- bf(
  leaf_out_z_global ~ leaf_out_prev_z_global + LSF2 + Tmean.EQ1.SO1_z_global + Tmean.SO1.EQ2_z_global + 
    Tmean.EQ2.SO2_z_global + Tmax_z_global + RH_z_global + SWrad_z_global + (year_z_global || timeseries)
)

# Model including leaf_off (used if data is available for that species)
formula_lo <- bf(
  leaf_out_z_global ~ leaf_out_prev_z_global + LSF2 + Tmean.EQ1.SO1_z_global + Tmean.SO1.EQ2_z_global + leaf_off_z_global + 
    Tmean.EQ2.SO2_z_global + Tmax_z_global + RH_z_global + SWrad_z_global + (year_z_global || timeseries)
)

#-----------------------------------------------------------
# Setup: model loop across phenological periods and species
#-----------------------------------------------------------

periods <- c("early", "late")          # Define observation periods
species_list <- unique(PEP.df$species) # Extract species names
models <- list()                       # Create empty list to store models

#-----------------------------------------------------------
# Loop over each period × species combination
#-----------------------------------------------------------

for (p in periods) {
  models[[p]] <- list()                # Create sublist for each period
  
  for (sp in species_list) {
    cat("Running model for Period:", p, "| Species:", sp, "\n")
    
    # Subset the data for this species and period
    data_sub <- PEP.df %>% filter(period == p, species == sp)
    
    # Skip if there are too few observations for stable estimation
    if (nrow(data_sub) < 30) {
      cat("Skipping species due to insufficient data:", sp, "\n")
      next
    }
    
    #-----------------------------------------------------------
    # Use different formula depending on whether 'leaf_off' is available
    #-----------------------------------------------------------
    
    include_leaf_off <- any(!is.na(data_sub$leaf_off_z))
    formula <- if (include_leaf_off) formula_lo else formula_nolo
    
    #-----------------------------------------------------------
    # Fit the model using brms and CmdStanR backend
    #-----------------------------------------------------------
    
    models[[p]][[sp]] <- brm(
      formula = formula,               # The model formula, chosen dynamically depending on whether 'leaf_off' is available
      data = data_sub,                 # Subset of data for this species and period
      family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
      prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
      
      backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
      #init = 0,                       # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
      
      chains = 4,                      # Number of MCMC chains to run (default is 4)
      cores = 12,                      # Number of CPU cores used in total
      threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
      
      iter = 4000,                     # Total number of iterations per chain (including warmup)
      warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
      
      control = list(
        adapt_delta = 0.95,            # Target acceptance probability for the sampler (higher = fewer divergences)
        max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
      ),
      
      seed = 123                       # Set random seed for reproducibility
    )
  }
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_full_models_noScale.rds"))
```

</details>

<details>

<summary>No Year model</summary>

```{r}
#-----------------------------------------------------------
# Define priors
#-----------------------------------------------------------

priors_brm <- c(
  # Fixed effects — weakly informative (small shrinkage)
  set_prior("normal(0, 1)", class = "b"), 
  
  # Intercept — mild regularization (helps model intercept stability)
  set_prior("normal(0, 1)", class = "Intercept"), 
  
  # Residual SD (sigma):
  set_prior("student_t(3, 0, 1)", class = "sigma"),
  
  # General SDs for random effects — conservative, heavy tails
  set_prior("student_t(3, 0, 2.5)", class = "sd")
)

#-----------------------------------------------------------
# Define modeling formulae
#-----------------------------------------------------------

# Model without leaf_off (used if no data for that variable)
formula_nolo <- bf(
  leaf_out_z ~ leaf_out_prev_z + LSF2 + Tmean.EQ1.SO1_z + Tmean.SO1.EQ2_z + 
    Tmean.EQ2.SO2_z + Tmax_z + RH_z + SWrad_z + (1 | timeseries)
)

# Model including leaf_off (used if data is available for that species)
formula_lo <- bf(
  leaf_out_z ~ leaf_out_prev_z + LSF2 + Tmean.EQ1.SO1_z + Tmean.SO1.EQ2_z + leaf_off_z + 
    Tmean.EQ2.SO2_z + Tmax_z + RH_z + SWrad_z + (1 | timeseries)
)

#-----------------------------------------------------------
# Setup: model loop across phenological periods and species
#-----------------------------------------------------------

periods <- c("early", "late")          # Define observation periods
species_list <- unique(PEP.df$species) # Extract species names
models <- list()                       # Create empty list to store models

#-----------------------------------------------------------
# Loop over each period × species combination
#-----------------------------------------------------------

for (p in periods) {
  models[[p]] <- list()                # Create sublist for each period
  
  for (sp in species_list) {
    cat("Running model for Period:", p, "| Species:", sp, "\n")
    
    # Subset the data for this species and period
    data_sub <- PEP.df %>% filter(period == p, species == sp)
    
    # Skip if there are too few observations for stable estimation
    if (nrow(data_sub) < 30) {
      cat("Skipping species due to insufficient data:", sp, "\n")
      next
    }
    
    #-----------------------------------------------------------
    # Use different formula depending on whether 'leaf_off' is available
    #-----------------------------------------------------------
    
    include_leaf_off <- any(!is.na(data_sub$leaf_off_z))
    formula <- if (include_leaf_off) formula_lo else formula_nolo
    
    #-----------------------------------------------------------
    # Fit the model using brms and CmdStanR backend
    #-----------------------------------------------------------
    
    models[[p]][[sp]] <- brm(
      formula = formula,               # The model formula, chosen dynamically depending on whether 'leaf_off' is available
      data = data_sub,                 # Subset of data for this species and period
      family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
      prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
      
      backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
      #init = 0,                       # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
      
      chains = 4,                      # Number of MCMC chains to run (default is 4)
      cores = 12,                      # Number of CPU cores used in total
      threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
      
      iter = 4000,                     # Total number of iterations per chain (including warmup)
      warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
      
      control = list(
        adapt_delta = 0.95,            # Target acceptance probability for the sampler (higher = fewer divergences)
        max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
      ),
      
      seed = 123                       # Set random seed for reproducibility
    )
  }
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_full_models_noYear.rds"))
```

</details>

<br>

# 3. Reproducibility

<details>

<summary>show</summary>

```{r}
# Datetime
Sys.time()

# Session info
sessionInfo()
```

</details>
