---
title: Spring phenology tier2 project
author: Constantin Zohner
date: "last updated 30 May 2025"

subtitle: Run Bayesian hierarchical models with alternative drivers (PEP725 data)
output:
  html_document:
    highlight: haddock
    toc: false
    use_bookdown: true
    df_print: paged
    fig_caption: yes
---

<style>
body
  { counter-reset: source-line 0; }
pre.numberSource code
  { counter-reset: none; }
</style>

<br>

### Driver names
- leaf_out_prev2: Leaf-out date two years ago
- LSF4: Binary indicator of late spring frost events (< -4°C) occurring after leaf-out in previous year
- leaf_off: Leaf senescence date in the previous year
- Tmean: Daily mean air temperature (°C) in previous year:
  - EQ2 = September equinox
  - SO2 = December solstice
  - GS  = May 1 - Sep 30
- RH: Spring relative air humidity (%)
- SWrad: Spring (preseason) incoming short-wave solar radiation (W/m²)
- Tmean: Spring (preseason) Daily mean air temperature (°C)

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  class.source = "numberLines lineAnchors"
  )
```

<br><br>

## 1. Load packages and data
<details><summary>get packages</summary>
```{r, message=F, warning=F}
library(tidyverse)
library(data.table)
library(car)
library(brms)
#install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
#cmdstanr::install_cmdstan()
library(cmdstanr)
options(brms.backend = "cmdstanr")
```
</details>

<details><summary>get data</summary>
```{r}
#####################
## Set directories ##
#####################



PEP.dir    = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_input/PEP_drivers_final_spring/Merged_file"
output.dir = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_output/ModelResults/Bayesian_models"



##############################################################################################################################################
##############################################################################################################################################



#################
## Import data ##
#################



#-----------------------------------------------------------
# define model variables
#-----------------------------------------------------------

full_variables = c("leaf_out", "leaf_out_prev", "leaf_off",
                   "LSF4",
                   "Tmean.GS","Tmean.EQ2.SO2",
                   "Tmean","RH","SWrad",
                   "year")

#-----------------------------------------------------------
# Load full data
#-----------------------------------------------------------

PEP.df <- fread(paste(PEP.dir, "pep_drivers_spring_data_clean_preseason.csv", sep="/")) %>% 
  dplyr::select(species, year, timeseries, lat, lon, alt_dem, all_of(full_variables))

#-----------------------------------------------------------
#  Add leaf-out date two years ago
#-----------------------------------------------------------

PEP.df =  PEP.df %>% 
  mutate(year2 = year) %>%  # temporary copy of year for join
  inner_join(
    PEP.df %>%
      transmute(timeseries, year = year + 2, leaf_out_prev2 = leaf_out),  # shift year for join
    by = c("timeseries", "year")
  ) %>%
  select(-year2)  # optional cleanup
nrow(PEP.df)

#-----------------------------------------------------------
# Reduced data including all leaf-off observations
#-----------------------------------------------------------

PEPred.df <- PEP.df %>% 
  filter(!is.na(leaf_off)) 
nrow(PEPred.df)



##############################################################################################################################################
##############################################################################################################################################



####################
## Filter periods ##
####################



#-----------------------------------------------------------
# Define periods
#-----------------------------------------------------------

min.year.full = min(PEP.df$year) # Earliest year
min.year.full
max.year.full = max(PEP.df$year) # Latest year
max.year.full
year.center.full = round(median(c(min.year.full, max.year.full))) # central year
year.center.full
year.center.full-(min.year.full-1) #length of each period
max.year.full-year.center.full

#-----------------------------------------------------------
# Early data full
#-----------------------------------------------------------

PEP_early.df <- PEP.df %>% 
  mutate(period = "early") %>% 
  filter(!alt_dem>600,
         year>=min.year.full, 
         year<=year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()

#-----------------------------------------------------------
# Early data reduced
#-----------------------------------------------------------

PEPred_early.df <- PEPred.df %>% 
  mutate(period = "early") %>% 
  filter(!alt_dem>600,
         year>=min.year.full, 
         year<=year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()

#-----------------------------------------------------------
# Late data full
#-----------------------------------------------------------

PEP_late.df <- PEP.df %>% 
  mutate(period = "late") %>% 
  filter(!alt_dem>600,
         year>year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()

#-----------------------------------------------------------
# Late data reduced
#-----------------------------------------------------------

PEPred_late.df <- PEPred.df %>% 
  mutate(period = "late") %>% 
  filter(!alt_dem>600,
         year>year.center.full) %>% 
  group_by(timeseries)%>%
  #delete time series with fewer than 20 years
  filter(n() >= 20) %>% 
  ungroup()



##############################################################################################################################################
##############################################################################################################################################



###############################
## Merge early and late data ##
###############################



#-----------------------------------------------------------
# Reduced data
#-----------------------------------------------------------

PEPred.df = bind_rows(PEPred_early.df, PEPred_late.df) %>% 
  filter(!species %in% c("Larix decidua"))
# Clean up
rm(PEPred_early.df, PEPred_late.df)
nrow(PEPred.df)

#-----------------------------------------------------------
# Full data
#-----------------------------------------------------------

PEP.df = bind_rows(PEP_early.df, PEP_late.df) %>% 
  # Remove all species that are already present in PEPred.df 
  # ensures that species with valid 'leaf_off' data are handled exclusively via PEPred.df
  filter(!species %in% unique(PEPred.df$species))
# Clean up
rm(PEP_early.df, PEP_late.df)
nrow(PEP.df)

#-----------------------------------------------------------
# Merge and scale continuous variables
#-----------------------------------------------------------

# Define continuous_vars after leaf_out_prev2 exists
binary_vars <- c("LSF2", "LSF4")
continuous_vars <- setdiff(c(full_variables, "leaf_out_prev2"), binary_vars)

PEP.df = bind_rows(PEP.df, PEPred.df) %>%
  
  # Group by observation period and site (i.e., timeseries = unique site-location combination)
  group_by(period, timeseries) %>%
  
  # Scale each continuous predictor within each site and period group
  # This removes across-site variation in mean/variance,
  # ensuring that model coefficients reflect *within-site* effects over time
  #
  # → Why is this important?
  # Because we're interested in whether *interannual variation* in predictors
  # (e.g., temperature, RH) explains *interannual shifts* in phenology,
  # not whether warmer sites leaf out earlier on average.
  #
  # Without this grouping, variables would be scaled globally,
# and site-level means could dominate effect estimates —
# falsely inflating signals driven by spatial gradients rather than temporal dynamics.
#
# Example: If warm lowland sites always have earlier leaf-out, a global
# correlation between Tmean and leaf_out could appear strong,
# even if within-site interannual correlations are weak or inconsistent.

mutate(across(all_of(continuous_vars), scale)) %>%
  
  # Remove group structure
  ungroup()
nrow(PEP.df)

# Clean up
rm(PEPred.df)
```
</details>

<br>

## 2. Check multicollinearity using variance inflation factors (VIFs)
<details><summary>show code</summary>
```{r}
#==================================================
# FULL MODELS — Includes all general predictors
# Adds 'leaf_off' if available for the species
#==================================================

VIF_full.df <- PEP.df %>%
  group_by(species) %>%
  group_modify(~ {
    data_subset <- .x
    
    # Determine if 'leaf_off' is usable for this species
    include_leaf_off <- any(!is.na(data_subset$leaf_off))
    
    # Conditionally construct model formula
    formula <- if (include_leaf_off) {
      leaf_out ~ leaf_out_prev2 + LSF4 + leaf_off + 
        Tmean.GS + Tmean.EQ2.SO2 +
        Tmean + RH + SWrad + year
    } else {
      leaf_out ~ leaf_out_prev2 + LSF4 + 
        Tmean.GS + Tmean.EQ2.SO2 +
        Tmean + RH + SWrad + year
    }
    
    #--------------------------------------------------
    # Fit model and compute VIFs
    #--------------------------------------------------
    model <- lm(formula, data = data_subset)
    vif_vals <- vif(model)
    
    # Return tidy VIF output
    data.frame(
      Variable = names(vif_vals),
      VIF = as.numeric(vif_vals)
    )
  }) %>%
  ungroup() %>%
  
  # Reshape into wide format: each species gets its own column
  pivot_wider(names_from = species, values_from = VIF)

# View resulting wide-format VIF table
VIF_full.df
```
</details>

<br>

## 3. Run Bayesian hierarchical models
<details><summary>show code</summary>
```{r, message=F, warning=F, results='hide'}
#-----------------------------------------------------------
# Define priors
#-----------------------------------------------------------

priors_brm <- c(
  # Fixed effects — weakly informative (small shrinkage)
  set_prior("normal(0, 1)", class = "b"), 
  
  # Intercept — mild regularization (helps model intercept stability)
  set_prior("normal(0, 1)", class = "Intercept"), 
  
  # General SDs for random effects — conservative, heavy tails
  set_prior("student_t(3, 0, 2.5)", class = "sd"), 
  
  # Specific shrinkage for random slope of year by timeseries
  set_prior("normal(0, 0.5)", class = "sd", group = "timeseries", coef = "year")
)

#-----------------------------------------------------------
# Define modeling formulae
#-----------------------------------------------------------

# To account for temporal trends in phenology and potential non-independence of observations across years within each site, 
# we included year as a random slope grouped by timeseries using an uncorrelated random effects structure (year || timeseries). 
# This formulation allows both the baseline phenology (intercept) and the temporal trend (slope of year) to vary across sites, 
# while simplifying model complexity by not estimating the correlation between them.
#
# By doing so, we capture spatial heterogeneity in long-term phenological trends and reduce the risk of inflated fixed-effect 
# estimates caused by unmodeled site-specific trajectories. This specification preserves our ability to detect interannual 
# relationships between climate predictors and leaf-out timing within sites, while controlling for varying temporal baselines 
# across locations in a computationally efficient and statistically robust manner.

# Model without leaf_off (used if no data for that variable)
formula_nolo <- bf(
  leaf_out ~ leaf_out_prev2 + LSF4 +
        Tmean.GS + Tmean.EQ2.SO2 +
        Tmean + RH + SWrad + (year || timeseries)
)

# Model including leaf_off (used if data is available for that species)
formula_lo <- bf(
  leaf_out ~ leaf_out_prev2 + LSF4 + leaf_off + 
        Tmean.GS + Tmean.EQ2.SO2 +
        Tmean + RH + SWrad + (year || timeseries)
)

#-----------------------------------------------------------
# Setup: model loop across phenological periods and species
#-----------------------------------------------------------

periods <- c("early", "late")          # Define observation periods
species_list <- unique(PEP.df$species) # Extract species names
models <- list()                       # Create empty list to store models

#-----------------------------------------------------------
# Loop over each period × species combination
#-----------------------------------------------------------

for (p in periods) {
  models[[p]] <- list()                # Create sublist for each period
  
  for (sp in species_list) {
    cat("Running model for Period:", p, "| Species:", sp, "\n")
    
    # Subset the data for this species and period
    data_sub <- PEP.df %>% filter(period == p, species == sp)
    
    # Skip if there are too few observations for stable estimation
    if (nrow(data_sub) < 30) {
      cat("Skipping species due to insufficient data:", sp, "\n")
      next
    }
    
    #-----------------------------------------------------------
    # Use different formula depending on whether 'leaf_off' is available
    #-----------------------------------------------------------
    
    include_leaf_off <- any(!is.na(data_sub$leaf_off))
    formula <- if (include_leaf_off) formula_lo else formula_nolo
    
    #-----------------------------------------------------------
    # Fit the model using brms and CmdStanR backend
    #-----------------------------------------------------------
    
    models[[p]][[sp]] <- brm(
      formula = formula,               # The model formula, chosen dynamically depending on whether 'leaf_off' is available
      data = data_sub,                 # Subset of data for this species and period
      family = gaussian(),             # Specifies a Gaussian likelihood (appropriate for continuous 'leaf_out' response)
      prior = priors_brm,              # Priors for intercept, fixed effects, and random effect standard deviations
  
      backend = "cmdstanr",            # Use CmdStanR backend for efficient and parallel sampling
      init = 0,                        # Initialize all model parameters at 0 (can help avoid divergent transitions early on)
      
      chains = 4,                      # Number of MCMC chains to run (default is 4)
      cores = 12,                      # Number of CPU cores used in total
      threads = threading(threads = 3),# Run 3 threads per chain (parallelized within-chain sampling; total = 4 × 3 = 12)
      
      iter = 4000,                     # Total number of iterations per chain (including warmup)
      warmup = 1000,                   # Number of warmup (burn-in) iterations per chain
      
      control = list(
        adapt_delta = 0.99,            # Target acceptance probability for the sampler (higher = fewer divergences)
        max_treedepth = 15             # Maximum depth of the NUTS tree (helps with complex posterior geometry)
      ),
      
      seed = 123                       # Set random seed for reproducibility
    )
  }
}

#-----------------------------------------------------------
# Save all fitted models to disk
#-----------------------------------------------------------

saveRDS(models, file = file.path(output.dir, "brms_full_models_alternative.rds"))
```
</details>

<br>

## 4. Reproducibility
<details><summary>show code</summary>
```{r}
## datetime
Sys.time()

## session info
sessionInfo()
```