---
title: Spring phenology tier2 project
author: Constantin Zohner
date: "last updated 6 May 2025"

subtitle: R code generating the preseason drivers for the 15-year moving window analysis for leaf-out observations and model predictions
output:
  html_document:
    highlight: haddock
    toc: false
    use_bookdown: true
    df_print: paged
---

<style>
body
  { counter-reset: source-line 0; }
pre.numberSource code
  { counter-reset: none; }
</style>

<br>

### Driver names
- Tmean: Mean daily temperature
- Tmin: Minimum daily temperature
- Tmax: Maximum daily temperature
- SWrad: Net short-wave radiation
- RH: Relative humidity

### Species names
- Deciduous trees: Aesculus hippocastanum, Alnus glutinosa, Betula pendula, Fagus sylvatica, Fraxinus excelsior, Quercus robur
- Deciduous shrub: Corylus avellana, Ribes uva-crispa
- Deciduous needleleaf: Larix decidua
- Evergreen needleleaf: Picea abies, Pinus sylvestris

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  class.source = "numberLines lineAnchors"
  )
```

<br><br>

```{r, message=FALSE, warning=FALSE, attr.source='.numberLines'}
#####################
# Required packages #
#####################



require(data.table)
require(tidyverse)
require(pbmcapply)
require(chillR)



##############################################################################################################################################
##############################################################################################################################################



################
# Define Paths #
################



# 1. Input
##########

# PEP data
PEP.dir = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_input/PEP_drivers_final_spring/Merged_file"

# Modeled data
model.dir = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_input/PEP_models"

# Climate
EOBS.dir = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_input/EOBS_data"


# 2. Output
###########

Output.dir1 = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_input/PEP_drivers_final_spring_MW/MW15"
Output.dir2 = "/Users/consti/Desktop/PhD/Publication_material/40_Spring_phenology_tier2/PEP725_Analysis/Analysis_input/PEP_drivers_final_spring_MW/MW15/Missing_observations"



##############################################################################################################################################
##############################################################################################################################################



#################
## Import data ##
#################



# Phenology data
################

PEP.df   = fread(paste(PEP.dir, "pep_drivers_spring_data.csv", sep="/")) %>% 
  dplyr::select(c(s_id, species, timeseries, year, alt_dem, leaf_out))


# Modeled data
##############

model.df = fread(paste(model.dir, "pep_model_predictions.csv", sep="/")) %>% 
  dplyr::select(-c(V1, leaf_out))
  



##############################################################################################################################################
##############################################################################################################################################



################
## Merge data ##
################



# Merge and filter data
PEP.df = inner_join(PEP.df, model.df, by = c("s_id", "species", "timeseries", "year")) %>% 
  group_by(timeseries)%>%
  filter(!alt_dem>600, year>=1966) %>%
  filter(n() >= 20) %>% 
  ungroup()

# Long format 
PEP.df = pivot_longer(PEP.df, 
                       cols = c(leaf_out, leaf_out_GDD, leaf_out_GDD.chill.lo, leaf_out_GDD.chill.hi, leaf_out_GDD.chill.DL), 
                       names_to = "phenology", values_to = "leaf_out")

# Year vector
year.vector  = c(min(PEP.df$year):(max(PEP.df$year)-14))

# Counter
i=1

# Create List object
List.year = replicate(length(year.vector), data.frame())
names(List.year) = year.vector

# Loop to bind individual moving-window data frames
for(MW.year in year.vector) {
  
  PEP.df.sub = PEP.df %>% 
    # filter years within 15-year moving window
    filter(year >= MW.year,
           year < MW.year + 15) %>% 
    # Group by timeseries and phenology
    group_by(timeseries, phenology) %>%
    # Delete timeseries <12 years
    filter(n() >= 12) %>%
    # Add mean leaf-out date
    mutate(leaf_out_mean = round(mean(leaf_out)),
           leaf_out_sd   = sd(leaf_out)) %>% 
    ungroup() %>% 
    mutate(MWyear = MW.year,
           ts_yr  = paste(timeseries,phenology,year,MW.year, sep="_")) 
  
  # store dataframe in year list
  List.year[[i]] = PEP.df.sub
  
  # Increase counter and print
  i=i+1
  #print(MW.year)
}  

# Merge list to data frame
PEP.MW.df = do.call(rbind.data.frame, List.year)


#-------------------------------------------------------------------------


## Import daily climatic datasets from E-OBS
############################################

#define climate variables
list.files(EOBS.dir)
vn <- c("tg_ens_mean_0.1deg_reg_v30.0e.csv", # daily mean temperature 
        "tn_ens_mean_0.1deg_reg_v30.0e.csv", # daily minimum temperature
        "tx_ens_mean_0.1deg_reg_v30.0e.csv", # daily maximum temperature
        "hu_ens_mean_0.1deg_reg_v30.0e.csv", # daily mean relative humidity 
        "qq_ens_mean_0.1deg_reg_v29.0e.csv") # global radiation

#create empty list
DataList <- replicate(length(vn),data.frame())
#loop through climate variables
for(i in 1:length(vn)) {
  #read data
  data = fread(paste0(EOBS.dir, "/", vn[i]))
  #add table to list
  DataList[[i]] = data }
#add names to list
names(DataList)=vn



##############################################################################################################################################
##############################################################################################################################################



######################
## Helper functions ##
######################



#################################
# Replace NA with neighbor mean #
#################################


replace_na_with_neighbor_mean <- function(x) {
  if (!is.numeric(x)) return(x)  # Ensure only numeric columns are processed
  
  na_indices <- which(is.na(x))  # Identify NA indices
  known_indices <- which(!is.na(x))  # Identify non-NA values
  
  if (length(known_indices) == 0) {
    return(x)  # If all values are NA, return as is
  }
  
  for (i in na_indices) {
    # Find nearest non-NA before and after
    before_idx <- ifelse(any(known_indices < i), max(known_indices[known_indices < i]), NA)
    after_idx <- ifelse(any(known_indices > i), min(known_indices[known_indices > i]), NA)
    
    if (!is.na(before_idx) & !is.na(after_idx)) {
      # Compute mean of the surrounding values
      x[i] <- mean(c(x[before_idx], x[after_idx]), na.rm = TRUE)
    } else if (!is.na(before_idx)) {
      x[i] <- x[before_idx]  # If NA is at the end, take the previous value
    } else if (!is.na(after_idx)) {
      x[i] <- x[after_idx]  # If NA is at the beginning, take the next value
    }
  }
  
  return(x)
}



##############################################################################################################################################
##############################################################################################################################################



##############################
## Merge datasets into list ##
##############################



# Identifier 1 (all site x year combinations)
PEP.MW.df$site_year = paste0(PEP.MW.df$s_id,"_",PEP.MW.df$year)

# Identifier 2 (all timeseries x year combinations)
timeseries_year = unique(PEP.MW.df$ts_yr)
length(timeseries_year)

# add PEP data to list
DataList[[6]] = PEP.MW.df

rm(data, PEP.MW.df, PEP.df)
names(DataList)=c(vn,"PEP")
names(DataList)




############################################################
## Loop through each observation using parallel computing ##
############################################################



parallelCalc <- function(timeseries_years){ 

  # Subset input data by time-point
  #################################
  
  #phenology data
  pheno.sub  <- DataList[[6]][which(DataList[[6]]$ts_yr==timeseries_years),]
  
  # Define the length of the period of interest in calendar units
  year1     <- as.character(pheno.sub$year-1)
  year2     <- as.character(pheno.sub$year)
  start_doy <- paste(year1,"-01-01", sep="") 
  end_doy   <- paste(year2,"-06-30", sep="")
  days      <- seq(as.Date(start_doy), as.Date(end_doy), by="days")
  
  #get number of days in year 1 and year 2
  days_year1 <- length(seq(as.Date(start_doy), as.Date(paste(year1,"-12-31", sep="")), by="days"))
  days_year2 <- length(seq(as.Date(paste0(year2,"-01-01")), as.Date(end_doy), by="days"))
  
  
  # ------------------------------------------------------------------------
  # Short Name        ; Long Name                               ; Unit [d-1]
  # ------------------------------------------------------------------------
  # SWRAD             ; Surface shortwave downwelling radiation ; W m-2
  # RH                ; Relative humidity                       ; %
  # TMIN              ; Minimum Air Temperature                 ; degC
  # TMEAN             ; Mean Air Temperature                    ; degC
  # TMAX              ; Maximum Air Temperature                 ; degC
  # ------------------------------------------------------------------------
  
  
  #Tmean
  T_mean.sub2 <- DataList[[1]][which(DataList[[1]]$site_year==pheno.sub$site_year),]%>%
    dplyr::select(as.character(1:days_year2))
  
  #Skip timeseries for which there is no data
  if (
    # 1) no climate data for current year and/or
    nrow(T_mean.sub2)==0 |
    # 2) no climate data for previous year
    nrow(DataList[[1]][which(DataList[[1]]$s_id==pheno.sub$s_id & DataList[[1]]$year==year1),])==0
  ) {
    write.table(pheno.sub, file=paste0(Output.dir2, '/', timeseries_years, '.csv'), 
                sep=',', row.names = F, col.names = T)
  } else {
    
    T_mean.sub1 <- DataList[[1]][which(DataList[[1]]$s_id==pheno.sub$s_id & DataList[[1]]$year==year1),]%>% 
      dplyr::select(as.character(1:days_year1))
    T_mean.sub  <- cbind(T_mean.sub1, T_mean.sub2)
    
    #Tmin
    T_min.sub1 <- DataList[[2]][which(DataList[[2]]$s_id==pheno.sub$s_id & DataList[[2]]$year==year1),]%>% 
      dplyr::select(as.character(1:days_year1))
    T_min.sub2 <- DataList[[2]][which(DataList[[2]]$site_year==pheno.sub$site_year),]%>%
      dplyr::select(as.character(1:days_year2))
    T_min.sub  <- cbind(T_min.sub1, T_min.sub2)
    
    #Tmax
    T_max.sub1 <- DataList[[3]][which(DataList[[3]]$s_id==pheno.sub$s_id & DataList[[3]]$year==year1),]%>% 
      dplyr::select(as.character(1:days_year1))
    T_max.sub2 <- DataList[[3]][which(DataList[[3]]$site_year==pheno.sub$site_year),]%>%
      dplyr::select(as.character(1:days_year2))
    T_max.sub  <- cbind(T_max.sub1, T_max.sub2)
    
    #RH
    rh.sub1 <- DataList[[5]][which(DataList[[5]]$s_id==pheno.sub$s_id & DataList[[5]]$year==year1),]%>% 
      dplyr::select(as.character(1:days_year1))
    rh.sub2 <- DataList[[5]][which(DataList[[5]]$site_year==pheno.sub$site_year),]%>%
      dplyr::select(as.character(1:days_year2))
    rh.sub  <- cbind(rh.sub1, rh.sub2)
    
    #SWrad
    swrad.sub1 <- DataList[[5]][which(DataList[[5]]$s_id==pheno.sub$s_id & DataList[[5]]$year==year1),]%>% 
      dplyr::select(as.character(1:days_year1))
    swrad.sub2 <- DataList[[5]][which(DataList[[5]]$site_year==pheno.sub$site_year),]%>%
      dplyr::select(as.character(1:days_year2))
    swrad.sub  <- cbind(swrad.sub1, swrad.sub2)
    
    
    ##############################################################################################################################################
    
    
    ###############################
    # Create table of daily climate
    ###############################
    
    
    # Generate sub-dataframe to store results
    factors.sub <- pheno.sub %>% 
      dplyr::select(s_id,species,timeseries,phenology,alt_dem,
                    MWyear,year,
                    leaf_out,leaf_out_mean,leaf_out_sd)
    factors.sub = as.data.frame(factors.sub) 
    
    #create table
    daily_vals <- data.frame(Year     = lubridate::year(as.Date(days,origin=days[1])),
                             Month    = lubridate::month(as.Date(days,origin=days[1])),
                             Day      = lubridate::day(as.Date(days,origin=days[1])),
                             DOY      = as.numeric(names(T_mean.sub)),
                             Tmin     = as.numeric(T_min.sub), 
                             Tmean    = as.numeric(T_mean.sub), 
                             Tmax     = as.numeric(T_max.sub), 
                             SWrad    = as.numeric(swrad.sub),
                             RH       = as.numeric(rh.sub)) %>% 
      #replace NAs with neighbor mean
      mutate(across(c(Tmin, Tmean, Tmax, SWrad, RH), replace_na_with_neighbor_mean))
    
    
    ##############################################################################################################################################
    
    
    #####################
    # Get important dates
    #####################
    
    
    #row of mean leaf-out date
    DOY_out_mean = which(daily_vals$DOY == pheno.sub$leaf_out_mean & daily_vals$Year==year2)   
    
    
    ##############################################################################################################################################
    
    
    ############################
    # Get preseason temperatures
    ############################
    
    
    ## Calculate the average, SD and slopes of preseason temperatures 
    # 10 to 120 days prior to mean leaf-out date
    
    #get preseason length vector (10 to 120 days with 10-day steps)
    preseason.lengths = seq(10, 120, 10)
    
    #loop through preseasons
    for(preseason.length in preseason.lengths) {
      
      #name columns
      #means
      preseason.Tmean    <- paste0("Tmean.PS",  preseason.length)
      preseason.Tmax     <- paste0("Tmax.PS",   preseason.length)
      preseason.Tmin     <- paste0("Tmin.PS",   preseason.length)
      preseason.SWrad    <- paste0("SWrad.PS",  preseason.length)
      preseason.RH       <- paste0("RH.PS",     preseason.length)
      #sd
      preseason.sd.Tmean <- paste0("Tmean.IntraSD.PS", preseason.length)
      preseason.sd.Tmax  <- paste0("Tmax.IntraSD.PS",  preseason.length)
      preseason.sd.Tmin  <- paste0("Tmin.IntraSD.PS",  preseason.length)
      #slopes
      preseason.slope.Tmean <- paste0("Tmean.slope.PS", preseason.length)
      preseason.slope.Tmax  <- paste0("Tmax.slope.PS",  preseason.length)
      preseason.slope.Tmin  <- paste0("Tmin.slope.PS",  preseason.length)
      
      #add columns to table
      factors.sub = factors.sub %>%
        mutate(
          #means
          !!preseason.Tmean    := mean(daily_vals$Tmean[(DOY_out_mean-preseason.length):DOY_out_mean]),
          !!preseason.Tmax     := mean(daily_vals$Tmax[(DOY_out_mean-preseason.length):DOY_out_mean]),
          !!preseason.Tmin     := mean(daily_vals$Tmin[(DOY_out_mean-preseason.length):DOY_out_mean]),
          !!preseason.SWrad    := mean(daily_vals$SWrad[(DOY_out_mean-preseason.length):DOY_out_mean]),
          !!preseason.RH       := mean(daily_vals$RH[(DOY_out_mean-preseason.length):DOY_out_mean]),
          #sd
          !!preseason.sd.Tmean := sd(daily_vals$Tmean[(DOY_out_mean-preseason.length):DOY_out_mean]),
          !!preseason.sd.Tmax  := sd(daily_vals$Tmax[(DOY_out_mean-preseason.length):DOY_out_mean]),
          !!preseason.sd.Tmin  := sd(daily_vals$Tmin[(DOY_out_mean-preseason.length):DOY_out_mean]),
          #slope
          !!preseason.slope.Tmean := coef(lm(daily_vals$Tmean[(DOY_out_mean-preseason.length):DOY_out_mean] ~ daily_vals$DOY[(DOY_out_mean-preseason.length):DOY_out_mean]))[2],
          !!preseason.slope.Tmax  := coef(lm(daily_vals$Tmax[(DOY_out_mean-preseason.length):DOY_out_mean] ~ daily_vals$DOY[(DOY_out_mean-preseason.length):DOY_out_mean]))[2],
          !!preseason.slope.Tmin  := coef(lm(daily_vals$Tmin[(DOY_out_mean-preseason.length):DOY_out_mean] ~ daily_vals$DOY[(DOY_out_mean-preseason.length):DOY_out_mean]))[2])
      
    }
    
    
    ##############################################################################################################################################
    
    
    # Safe the table  
    return(factors.sub)
  }
}



##############################################################################################################################################
##############################################################################################################################################



#initialize the loop
outputlist         <- mclapply(timeseries_year[1:1000], parallelCalc, mc.cores=12, mc.preschedule=T)
climate.factors.df <- rbindlist(outputlist)



##############################################################################################################################################
##############################################################################################################################################



###################
## Safe the data ##
###################



#Safe table
write.csv(climate.factors.df, paste(Output.dir1, "pep_preseasonsMW15_spring_data_subset.csv", sep="/"))



##############################################################################################################################################
##############################################################################################################################################



#####################
## Reproducibility ##	
#####################



## date time
Sys.time()

## session info
sessionInfo()



##############################################################################################################################################
#############################################################THE END##########################################################################
##############################################################################################################################################
```